### 2023-08-04 jds.rmd
Supplemental material for JDS paper

---
title: "Evaluating Perceptual Judgements on 3D Printed Bar Charts"
author:
- first: Tyler
  first-init: T.
  last: Wiederich
  marker: 1
  footnote-id: 1
  email: twiederich2@huskers.unl.edu
- first: Susan
  first-init: S.
  last: VanderPlas
  marker: 1
  email: susan.vanderplas@unl.edu
short-title: Perceptual Judgements on 3D Printed Bar Charts
output:
  jds.rmd::pdf_article: default
affiliation:
- institution: "University of Nebraska-Lincoln"
  prefix: Department of Statistics
  country: United States of America
  marker: 1
classoption:
- letterpaper
- inpress
- dvipsnames
preamble: |
  \usepackage{amsfonts,amsmath,amssymb,amsthm} 
  \usepackage{booktabs} 
  \usepackage{lipsum} 
  \PassOptionsToPackage{dvipsnames}{xcolor} 
  \newcommand{\svp}[1]{{\textcolor{ForestGreen}{#1}}} 
  \newcommand{\tw}[1]{{\textcolor{Red}{#1}}}
abstract: "The use of 3D data visualizations has limitations when the third dimension
  does not convey any additional information to the viewer. Numerous studies advocate
  to avoid these types of graphs whenever possible, but these studies are almost entirely
  focused on the 2D projections of the 3D graphs. This paper describes the partial
  replication of a well-known paper in data visualization and its adaptation to focus
  on 3D printed bar charts. While current results from our study do not show differences
  between 2D and 3D graphs, we will use this study to provide a baseline for future studies dealing with graphics in 3D environments.\n"
keywords:
- graphics
- 3D bar charts
- 3D printing
bibliography: JDS2023.bib
footnote:
- content: Tyler Wiederich
  type: corresp
  id: 1
---

```{r, include = F}
knitr::opts_chunk$set(echo = F, dpi = 300)

library(tidyverse)
library(RSQLite)
library(lme4)

con = dbConnect(SQLite(), '../../experiment_interface/department.db')
dbListTables(con)
results = dbReadTable(con, 'results')
users = dbReadTable(con, 'user')
userMatrix = dbReadTable(con, 'userMatrix')
dbDisconnect(con)

valid.users <- users %>% 
  mutate(subject = paste0(nickname, participantUnique)) %>% 
  filter(age != 'Under 19')

#Fill in correct values for incorrect 3d graph kits
# 1 result, need to fill in fileID, graphtype, and plot
results = results %>% 
  mutate(fileID = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', 1, fileID),
         graphtype = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', 'Type1', graphtype),
         plot = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', '3dPrint', plot))

load('../../experiment_interface/data/set85data.Rdata')
load('../../experiment_interface/data/kits.Rdata')

trueRatios = datasets %>% 
  mutate(ratio.df = map(data, function(x)(x[!is.na(x[,'IDchr']),4])),
         trueRatio = map(ratio.df, function(x)(x[1,'Height'] / x[2,'Height']))) %>% 
  unnest(trueRatio) %>% 
  filter(fileID != 15) %>% 
  #mutate(Height = Height * 100) %>% 
  select(fileID, Height)

res = results %>% 
  left_join(trueRatios, by = 'fileID') %>% 
  mutate(response = log2(abs(byHowMuch - Height*100) + 1/8),
         subject = paste0(nickname, participantUnique),
         ratioLabel = round(100*Height, 1)) %>% 
  filter(subject %in% valid.users$subject) %>% 
  arrange(appStartTime) %>% 
  filter(whichIsSmaller == 'Triangle (▲)',
         as.Date.POSIXct(appStartTime) <= '2023-05-22') #Only verified results from before the start of SDSS

```

# Introduction

## Elementary Graphical Tasks
@cleveland_graphical_1984 published a paper that sets up the foundation for how accurately people extract quantitative information using elementary perceptual tasks (EPTs). 
Elementary Perceptual Tasks, according to these experiments, include assessing graphical elements such as position along a common scale, length, angle, and volume, and estimating the corresponding numerical value of these representations.
They found that estimates of ratios between positions along a common scale had smaller log errors than estimates of ratios between lengths. The study relied entirely on estimation accuracy, which may not always be relevant when extracting information from graphs. For example, estimation is less relevant when ordering values by size.
@heerCrowdsourcingGraphicalPerception2010b replicated some parts of @cleveland_graphical_1984 in an online setting using Mechanical Turk, largely replicating the results of the original study while demonstrating the utility of the Mechanical Turk platform.
<!-- While the two experiments in their paper highlighted the accuracy for position, length, and angle EPTs, they theorized that the accuracy of judgements using volume was worse than that for judgments using position along a common scale.  -->

Further studies have shown that 3D graphs are less accurate at portraying numeric information that 2D graphs [@barfield_effects_1989;@fisher_data_1997]. In certain contexts and conditions, there is some research suggesting that 3D graphs may better encode information \citep{brath_3d_2014}.

## 3D Graphical Perception
Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
This occasionally causes problems: the line-width illusion, for instance, has been attributed to implicit 3D perception of two-dimensional stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams [@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013].

The use of 3D graphics have been explored in multiple studies. 
@fisher_data_1997 explored the preference of using either 2D or 3D graphs and found that subjects tended to use simpler 2D graphs when tasked with extracting information. @barfield_effects_1989 compared 2D and 3D graphs presented on paper and on computers.
Their results showed that the accuracy of subject answers depended on their skill level. 
Novice subjects were more accurate with 2D paper graphs and experienced managers were more accurate with 3D computer graphs. 
For both experience levels, participants were more confident in their answers using 2D graphs over other presentations of data. 
There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may better encode information.
@brath_3d_2014 highlights the intrinsic attributes of 3D graphs and its benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data. 
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since 2014. 
Digital graphics has developed quickly, along with the hardware necessary to support these software developments. 
As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving the artificial charts into a more natural setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of technological developments, but because these charts provide the opportunity to make visual graphics accessible to those with limited or absent vision.

Here, we provide the process of replication and modernization of testing perceptual judgments to 2D graphs, 3D graphs projected in 2D environments, and 3D printed bar graphs.

## Selected Components from Cleveland and McGill

Cleveland and McGill provided a theory and tested for the ordering of perceptual importance for the elements of length, position, and angle. 
Their first experiment, referenced as the position-length experiment, used five types of bar charts.
Two of these were grouped bar charts and the other three were stacked bar charts.
Each chart had two bars used for comparison and participants were asked to determine which bar was smaller and give their perceived ratio of the smaller bar to the larger bar.
The two grouped bar charts are for the perceptual element of position along a common scale, where one has zero distance between bars and the other has a fixed distance between bars.
These grouped bar charts will be referenced as adjacent and separated graph types in this paper, respectively.

Our study replicates the procedure for the comparisons of the two grouped bar charts, but with an objective of detecting differences in accuracy between 2D graphs, 3D digital graphs, and 3D printed graphs.

# Methods

Our study is designed to replicate and expand upon the position-length experiment from Cleveland and McGill as closely as possible.
In this section, we will discuss the replication process and the design of our modified version of this experiment.

## Replicating Cleveland and McGill

The first step of replicating the position-length experiment was to determine the heights of the bars that  participants use for comparisons.
These values for the bar heights are linear on a log scale and are given by

\[s_i=10\cdot 10^{(i-1)/12}, \qquad i=1,...,10\]

Each graph presents two bars from the values given above where the participants are asked to judge the ratio of the smaller bar to the larger bar. The ratio of bars used by Cleveland and McGill were 17.8, 26.1, 38.3, 46.4 (twice), 56.2, 68.1 (twice), and 82.5 (twice).
The exact numeric comparisons were not disclosed, but the comparison values used in our study were subjected to the constraints of having the same ratio values and that no value was used more than twice.

Each graph is presented so that there are ten bars where only two of the bars are marked for identification.
Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.
Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).

## Stimuli Construction

The graphs share a common layout across all formats, where two groupings of five bars are identified by "A" and "B", respectively, and circles and triangles are used to identify the bars participants should compare. Example graphs are shown in
Figure \@ref(fig:plotTypes).
There are some graphical elements that cannot be easily portrayed via 3D printing. For this reason, all graph types do not have axes, grid lines, or floating titles.
<!-- The aspect ratio of the plots presented in the original paper is approximately 0.825:1. The aspect ratio of the plots presented in our study is closer to 0.864:1, but -->
<!-- The aspect ratio of the 2d plots (including only the bars is 2:1), while the aspect ratio of the 3d printed charts is 11:9.5. This is something we probably need to fix.... Until then, let's not go there -->

```{r plotTypes, fig.cap = "Two dimensional, two-dimensional digital rendering, and 3D-printed charts used in this study.", out.width = "100%"}
knitr::include_graphics("plot-types.png", dpi = 300)
```

The ggplot2 [@ggplot2] package was utilized to create the 2D bar charts. 
The scale axis was removed, leaving only the bars and a bar grouping identifier.
The bars used for comparisons had the identifying mark at a height of 5 out of 100 for the 2D plots, and the 3D plots had the identifying marks on top of the bars.

The 3D renderings and 3D printed charts were both created using OpenSCAD [@kintelOpenSCADDocumentation2023], which creates STL files from markup describing the object's geometric composition. 
Charts were composed of a platform, raised labels for the A and B groups of bars, circle and triangle markers indicating the bars of interest, and the bars themselves; values for the bar heights were inserted into the markup using R [@R].
In addition, an ID code was engraved into the bottom of the platform to uniquely identify each object; this allows the researchers to ensure that the 3D printed charts are correctly allocated to stimulus sets.

Digital renderings of the generated STL files were created using @rgl, which integrates into @shiny using the @mozillafoundationWebGL2D3D2023 extension.
Rendered 3D charts were initially angled corresponding to the default 3D bar charts present in Microsoft Excel, but WebGL's interactivity allows the user to rotate, scale, and otherwise interact with the chart to change the angle. 3D renderings were colored to correspond to the 3D printed chart filament color.
The default `rgl` lighting was replaced with three lights located in fixed positions around the rendered figure. 
The lights were positioned so that one was behind the rendered figure<!--(x: 910, y: -1000, z:1000)-->, another in front of the figure<!--(x: 780, y: 1030, z: 1000)-->, and one light below the figure<!--(x: 65, y: 15, z: -100)-->.
3D charts were printed with colored filament corresponding to a specific ratio comparison; this allowed researchers to visually assess kits to ensure that they contained five unique ratios.
Colors corresponding to each ratio were assigned randomly to ensure that chart color provided no useful information about the ratio value.
As printed, the base of the chart was 13cm x 3cm x 1cm, with the highest bar rising 9.5cm above the chart base.
Raised letters and shapes were 2mm above the base or bar, respectively.

## Experiment Design

Participants were provided a kit with five 3D-printed charts, comprising five of the seven unique ratio comparisons; we then used the kit ID to ensure that participants saw computer-rendered charts with ratios corresponding to those in the kit of physical charts.
This ensured that the experimental design was balanced across chart type and randomized with respect to the type of comparison (adjacent or separated).
We printed 21 kits containing five 3d-printed charts each.
<!-- There are 21 kits in total so that each combination of ratios is accounted for. -->
<!-- Each graph is randomly assigned either an adjacent or separated comparison. -->

```{r studyDesign, fig.cap = "A graphical representation of the study design. Only five of the seven ratios were used in each kit where the ratios are randomly selected. A total of 21 kits were created to include all combinations of the five ratios."}
knitr::include_graphics("study-design.png")
```

## Participant Recruitment

One interesting facet of @cleveland_graphical_1984 is the participant recruitment methodology: "For each experiment the subjects fell into two categories: (1) a group of females, mostly housewives, without substantial technical experience; (2) a mixture of males and females with substantial technical training and working in technical jobs. 
Most of the subjects in the position-length experiment participated in the position-angle experiment; in all cases repeat subjects judged the position-angle graphs first." 
It would seem likely that the authors recruited individuals within their respective departments as well as their wives.
In the spirit of replicating the study, members of the UNL Statistics department and their spouses, partners, and roommates were asked to participate in our study; this replicates the spirit of the original study without the implicit assumptions that graduate students and professors are (1) largely male, (2) heterosexual, and (3) have unemployed partners.

A total of 48 participants completed the study; demographics are shown in \@ref(fig:demographics).

```{r demographics, fig.cap="Demographic breakdown of participants in the study. All subjects were recruited from faculty and students in the statistics department at University of Nebraska-Lincoln.", message=FALSE, warning=FALSE, out.width='85%'}
library(patchwork)
valid.users$userAppStartTime <- as.POSIXct(valid.users$userAppStartTime,
           origin = '1970-1-1')

age.levels = sort(unique(valid.users$age))
p1 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
  group_by(age) %>% 
  summarise(count = n()) %>% 
  ggplot(mapping = aes(x = factor(age, levels = age.levels),
                       y = count)) + 
  geom_bar(stat = 'identity') +
  geom_text(mapping = aes(y = count+1, label = count),
             size = 2) +
  labs(title = 'Age Groups',
       x = '',
       y = 'Count') +
  theme_bw() + 
  theme(plot.title = element_text(hjust=0.5),
        axis.text.x = element_text(angle = 45, hjust=1, size = 8))

p2 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
  group_by(gender) %>% 
  summarise(count = n()) %>% 
  ggplot(mapping = aes(x = gender,
                       y = count)) + 
  geom_bar(stat = 'identity') +
  geom_text(mapping = aes(y = count+1, label = count),
             size = 2) +
  labs(title = 'Gender Groups',
       x = '',
       y = 'Count') +
  theme_bw() + 
  theme(plot.title = element_text(hjust=0.5),
        axis.text.x = element_text(angle = 45, hjust=1, size = 8))

educ.levels = sort(unique(valid.users$education))[c(2,4,3,1)]
p3 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
  group_by(education) %>% 
  summarise(count = n()) %>% 
  ggplot(mapping = aes(x = factor(education, levels = educ.levels),
                       y = count)) + 
  geom_bar(stat = 'identity') +
  geom_text(mapping = aes(y = count+2, label = count),
             size = 2) +
  labs(title = 'Education Levels',
       x = '',
       y = 'Count') +
  theme_bw() + 
  theme(plot.title = element_text(hjust=0.5),
        axis.text.x = element_text(angle = 45, hjust=1, size=8))

p = p1 + p2 + p3
p
```

## Data Collection

A Shiny applet was used for data collection, along with the provided kit of 3D printed charts.
Participants provided informed consent through the applet, and then were asked for demographic information (age, gender, education level). 
Then, participants were shown a "practice" page which allowed them to experiment with the data collection interface and practice estimating the ratio between the bars, as shown in Figure \@ref(fig:practice)

```{r practice, fig.cap = "Screenshot of Shiny application practice screen. Three 2D bar charts with different ratios were provided, along with sliders indicating the correct proportion. Participants could practice with the sliders and preview the questions that would be asked as part of the task.", out.width = "80%"}
knitr::include_graphics("03-Practice-2.png")
```

Directly before the experiment started, participants were asked to provide the kit ID, along with directions indicating that if the instructions indicated that participants should use a 3D chart for a task, the participant should select a chart from the kit, enter the ID code of that chart from the bottom of the object, and complete the requested task.
Participants were also instructed to make quick judgments for each graph and not to measure or estimate ratios using physical objects.

```{r experiment3dRender, fig.cap = "Screenshot of the applet collecting data for a 3D rendered chart task. Participants were asked to select which bar (circle or triangle) was smaller, and then to estimate the ratio of the smaller bar to the larger bar.", out.width = "80%"}
knitr::include_graphics("05-Experiment-05-filled-in.png")
```

Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in Figure \@ref(fig:experiment3dRender).

# Results

All responses that incorrectly identified the smaller bar were removed from the study before analysis.

## Midmeans of Log Absolute Errors
Cleveland and McGill used 
$$\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$$
to measure accuracy of their participant's responses. 
In their study, log base 2 seemed appropriate due to "average relative errors changing by factors less than 10."
They also added 1/8 to prevent distortions when the errors were close to zero. 
@heerCrowdsourcingGraphicalPerception2010b followed the same analysis method, replicating many (but not all) of the results presented in the original paper.

Figure \@ref(fig:midmeans-log-errors) shows the midmeans of the log absolute errors compared to the true ratio of the bars for each graph type and comparison type. 
The results tend to indicate that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types.

```{r midmeans-log-errors, fig.cap='Midmeans of log absolute errors for the true ratio of bars. Each overlaying line are computed with loess.', message=FALSE, warning=FALSE, out.width='80%'}
plot.types = c('2D', 'Rendered 3D', '3D Printed')
names(plot.types) = c('2dDigital', '3dDigital', '3dPrint')

res %>% 
  group_by(Height, graphtype, plot) %>% 
  summarize(midmean = mean(response, trim=0.25, na.rm = T)) %>% 
  ggplot(mapping = aes(x = Height, y = midmean,
                       color = graphtype)) +
  geom_point() +
  geom_smooth(se = F, alpha=1/2) +
  scale_color_discrete(labels = c('Adjacent', 'Separated')) +
  facet_wrap(~plot, labeller = labeller(plot = plot.types)) +
  labs(title = '',
       x = 'True Proportional Difference (%)',
       y = 'Midmeans of Log Error',
       color = 'Comparison Type') +
  theme_bw() +
  theme(legend.position = 'bottom')
```
This is somewhat different than the results in @cleveland_graphical_1984 and @heerCrowdsourcingGraphicalPerception2010b; in both cases the midmean log absolute errors increased until about 55% of the true proportional difference and then decreased.
It is possible that this difference is due to the fact that we did not require an explicit numerical estimate of the ratio but instead asked participants to indicate the ratio on a slider (which is essentially a number line). 
This should reduce the cognitive load required to transition from spatial comparison to numerical comparison and then to compute the proportion, but may also have impacted the results.

## Linear Mixed Effects Model

In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a linear mixed effects model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio. This allows us to test for significant differences (though given the midmean log absolute error plots, we do not expect to find any) as well as to quantify effect sizes for future studies. The formal statistical model is as follows:

$$y_{ijklm}=\mu+S_i+R_j+G(R)_{(k)j}+T_l+\epsilon_{ijklm}$$

\noindent where

- $y_{ijklm}=\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$

- $S_i\sim N(0,\sigma^2_S)$ is the effect of the $i^{th}$ subject

- $R_j$ is the effect of the $j^{th}$ ratio

- $G(R)_{(k)j}$ is the effect of the $k^{th}$ graph type nested in the $j^{th}$ ratio

- $T_l$ is the effect of the $l^{th}$ comparison type

- $\epsilon_{ijklm}\sim N(0,\sigma^2_\epsilon)$ is the random error

```{r, include=FALSE}
mod = lmer(response ~ (1|subject) + ratio + type + ratio:plot,
     data = res)
anova(mod)
```

No differences were detected for the true ratio of bars (p-value = .7881), whether the bars were adjacent or separated (p-value = .3375), or for the plot nested within the true ratio (p-value = .6868). 
While Heer \& Bostock provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data at this time.

# Discussion and Future Work

Previous work in 3D graphics would suggest that the errors for the 3D graphs would be larger than the errors for the 2D graphs. 
While we did not find any significant results indicating that 3D graphs are read less accurately, there are two possibilities that might account for this discrepancy.

The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger. However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @cleveland_graphical_1984 do not include significance testing beyond the graphical display of confidence intervals.

The second possibility is more interesting: we examined 3D charts using rendered 3D graphs and 3D printed charts; both of these options allow for participants to interact with the chart, rotating it, and generally perceiving it as one might perceive any other 3D, real, object.
This is a far cry from the 3D perspective charts in the original study, which have a fixed angle and perspective and are thus not equivalent to our 3D charts. 
Future studies should include an additional fixed 3D perspective bar chart, which will at least enable us to examine whether modern 3D rendering environments allow for more accurate conclusions than fixed 3D perspectives. Future iterations of this study will include "traditional" 3D graphs created by Microsoft Excel (that is, graphs with a fixed 3D perspective rendered in 2D). This option will allow us to examine fixed perspective 3D plots compared to 3D renderings and 2D plots; it will also enable online data collection in addition to the in-person data collection used in this experiment.

Another interesting aspect of our study is that the method used to record participant estimates is different from the method used in the original study as well as Heer \& Bostock's replication study. 
A method similar to our slider input, marking position on a line, was used in @spenceVisualPsychophysicsSimple1990, but Spence asked participants to estimate $A/(A+B)$, where we asked participants to estimate $A/B$; thus, our results are still not directly comparable to previous studies.
We expect that the specific ratio estimated would also have an effect on observed participant errors.

The slider method for input of ratio estimates should be easier for participants, as it does not require explicit transformation to the numerical domain. 
What is clear is that it would be beneficial to assess the impact of measurement method on participant errors directly, so that the results of these different studies might be explained and interpreted with regard both to the stimuli used and the measurement method employed in the experiment.
Future iterations of this experiment will likely address this estimation difference; such modifications in experimental design are relatively straightforward in Shiny and will provide useful insight into the design of future experiments evaluating the perception of statistical graphics.

<!-- The goal of this study was to verify that previous research into graphics is upheld under a new set of circumstances. It started with a question, followed by research into prior work, and then by designing and conducting a study. This general framework of research is useful to all fields of study and is taught in most introductory statistics courses. The next step of this study is to collect more data while simultaneously providing students the opportunity to learn about the research process from both the perspectives of an experiment participant and the views of the researchers. At University of Nebraska-Lincoln, students taking introductory statistics will have the opportunity to participate in the experiment while also completing pre/post experiment surveys and reflections on a two-page abstract and a presentation of results. -->

## Supplemental Material
Stimuli, code, and data for this experiment are provided at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).

### 2023-08-04 paper-code.R

@@ -1,0 +1,156 @@
+## ---- include = F-------------------------------------------------------------------------------------------
+knitr::opts_chunk$set(echo = F, dpi = 300)
+
+library(tidyverse)
+library(RSQLite)
+library(lme4)
+
+con = dbConnect(SQLite(), '../../experiment_interface/department.db')
+dbListTables(con)
+results = dbReadTable(con, 'results')
+users = dbReadTable(con, 'user')
+userMatrix = dbReadTable(con, 'userMatrix')
+dbDisconnect(con)
+
+valid.users <- users %>% 
+  mutate(subject = paste0(nickname, participantUnique)) %>% 
+  filter(age != 'Under 19')
+
+
+#Fill in correct values for incorrect 3d graph kits
+# 1 result, need to fill in fileID, graphtype, and plot
+results = results %>% 
+  mutate(fileID = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', 1, fileID),
+         graphtype = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', 'Type1', graphtype),
+         plot = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', '3dPrint', plot))
+
+
+
+load('../../experiment_interface/data/set85data.Rdata')
+load('../../experiment_interface/data/kits.Rdata')
+
+
+trueRatios = datasets %>% 
+  mutate(ratio.df = map(data, function(x)(x[!is.na(x[,'IDchr']),4])),
+         trueRatio = map(ratio.df, function(x)(x[1,'Height'] / x[2,'Height']))) %>% 
+  unnest(trueRatio) %>% 
+  filter(fileID != 15) %>% 
+  #mutate(Height = Height * 100) %>% 
+  select(fileID, Height)
+
+res = results %>% 
+  left_join(trueRatios, by = 'fileID') %>% 
+  mutate(response = log2(abs(byHowMuch - Height*100) + 1/8),
+         subject = paste0(nickname, participantUnique),
+         ratioLabel = round(100*Height, 1)) %>% 
+  filter(subject %in% valid.users$subject) %>% 
+  arrange(appStartTime) %>% 
+  filter(whichIsSmaller == 'Triangle (▲)',
+         as.Date.POSIXct(appStartTime) <= '2023-05-22') #Only verified results from before the start of SDSS
+
+
+
+
[…]

### 2024-01-16 jds.rmd
smoothing out the abstract and introduction

@@
 output:
-  jds.rmd::pdf_article: default
+  jds.rmd::pdf_article: 
+    extra_dependencies: "subfig"
 affiliation:
@@
   \newcommand{\tw}[1]{{\textcolor{Red}{#1}}}
-abstract: "The use of 3D data visualizations has limitations when the third dimension
-  does not convey any additional information to the viewer. Numerous studies advocate
-  to avoid these types of graphs whenever possible, but these studies are almost entirely
-  focused on the 2D projections of the 3D graphs. This paper describes the partial
-  replication of a well-known paper in data visualization and its adaptation to focus
-  on 3D printed bar charts. While current results from our study do not show differences
-  between 2D and 3D graphs, we will use this study to provide a baseline for future studies dealing with graphics in 3D environments.\n"
+abstract: "Graphical design principles typically recommend minimizing the dimensionality of a visualization - for instance, using only 2 dimensions for bar charts rather than providing a 3D rendering, because this extra complexity may result in a decrease in accuracy.
+This advice has been oft repeated, but the underlying experimental evidence is focused on fixed 2D projections of 3D charts.
+In this paper, we describe an experiment which attempts to establish whether the decrease in accuracy extends to 3D virtual renderings and 3D printed charts.
+We replicate the grouped bar chart comparisons in the  1984 Cleveland & McGill study, assessing the accuracy of numerical estimates using different types of 3D and 2D renderings."
 keywords:
@@
 
-con = dbConnect(SQLite(), '../../experiment_interface/department.db')
+con = dbConnect(SQLite(), '../2022-Tyler-3D_Graphical_Perception/experiment_interface/department.db')
 dbListTables(con)
@@
 
-load('../../experiment_interface/data/set85data.Rdata')
-load('../../experiment_interface/data/kits.Rdata')
+load('../2022-Tyler-3D_Graphical_Perception/experiment_interface/data/set85data.Rdata')
+load('../2022-Tyler-3D_Graphical_Perception/experiment_interface/data/kits.Rdata')
 
@@
 # Introduction
+Good communication requires both that the information be transmitted correctly and that the intended recipient be able to decode and understand the transmitted information accurately.
+In order to communicate effectively, we must use graphical forms that accurately convey information relevant to the task in question.
+In many cases, this means we must understand how accurately people can read quantitative information off of charts.
+While accuracy is not the only quantity of interest in graphical investigations [@hullmanPursuitErrorSurvey2019], it is an important factor in assessing the utility of many different data graphics.
+
+The accuracy of graphical forms has been studied for almost a century [@vonhuhnFurtherStudiesGraphic1927;@eellsRelativeMeritsCircles1926;@croxtonGraphicComparisonsBars1932;@croxtonBarChartsCircle1927], as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions.
+This is particularly true in areas like graphics which are affected by the immense technological innovation in hardware and software which has taken place since the early 1990s.
 
 ## Elementary Graphical Tasks
-@cleveland_graphical_1984 published a paper that sets up the foundation for how accurately people extract quantitative information using elementary perceptual tasks (EPTs). 
+@clevelandGraphical1984 established the comparative accuracy of different "elementary perceptual tasks" (EPTs).
 Elementary Perceptual Tasks, according to these experiments, include assessing graphical elements such as position along a common scale, length, angle, and volume, and estimating the corresponding numerical value of these representations.
-They found that estimates of ratios between positions along a common scale had smaller log errors than estimates of ratios between lengths.
-The study relied entirely on estimation accuracy, which may not always be relevant when extracting information from graphs.
-For example, estimation is less relevant when ordering values by size.
-@heerCrowdsourcingGraphicalPerception2010b replicated some parts of @cleveland_graphical_1984 in an online setting using Mechanical Turk, largely replicating the results of the original study while demonstrating the utility of the Mechanical Turk platform.
-<!-- While the two experiments in their paper highlighted the accuracy for position, length, and angle EPTs, they theorized that the accuracy of judgements using volume was worse than that for judgments using position along a common scale.
--->
+The study relied entirely on estimation accuracy, which may not always be relevant when extracting information from graphs. For example, estimation is less relevant when ordering values by size.
+As a result of the Cleveland and McGill [-@clevelandGraphical1984] study, it is possible to assemble an ordering of perceptual accuracy for the elements of length, position, and angle. 
+@heerCrowdsourcingGraphicalPerception2010b replicated some parts of @clevelandGraphical1984 in an online setting using Mechanical Turk, largely validating the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.
 
-Further studies have shown that 3D graphs are less accurate at portraying numeric information that 2D graphs [@barfield_effects_1989;@fisher_data_1997].
-In certain contexts and conditions, there is some research suggesting that 3D graphs may better encode information \citep{brath_3d_2014}.
+The first experiment in @clevelandGraphical1984 (the position-length experiment), used five types of bar charts: two types of grouped bar charts and three types of stacked bar charts.
+Each chart had two bars marked for comparison; participants were asked to determine which bar was smaller and give the  perceived ratio of the smaller bar to the larger bar.
+@fig-bar-types show the two types of grouped bar charts.
+We are primarily interested in the grouped bar charts (in part because 3D printing is not yet inexpensive enough to make moderate-scale stacked bar chart experiments viable), which consisted of two comparison bars which were either adjacent or in separate groups.
+These grouped bar charts will be referenced as adjacent and separated graph types in this paper, respectively.
+
+```{r bar-types}
+#| message: false
+#| error: false
+#| warning: false
+#| fig-width: 4
+#| fig-height: 4
+#| fig-cap: "@clevelandGraphical1984 used two different types of grouped bar charts, which are replicated here."
+#| fig-subcap: 
+#|    - "Adjacent"
+#|    - "Separated"
+#| layout-ncol: 2
+library(ggplot2)
+
+datasets$data[[1]] %>%
+  ggplot(aes(x = Group, y = Height, group = GroupOrder)) + 
+  geom_col(position = "dodge", fill = "white", color = "black") + 
+  geom_text(aes(x = Group, y = 5, group = GroupOrder, label = IDchr), position = position_dodge(0.9)) + 
+  theme_bw() + 
+  theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
+  ggtitle("Adjacent Chart")
+
+datasets$data[[1]] %>%
+  ggplot(aes(x = factor(Order %%2), y = Height, group = GroupOrder)) + 
+  geom_col(position = "dodge", fill = "white", color = "black") + 
+  geom_text(aes(x = factor(Order %%2), y = 5, group = GroupOrder, label = IDchr), position = position_dodge(0.9)) + 
+  theme_bw() + 
+  theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
+  ggtitle("Separated Chart")
+
+```
 
@@
 Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
-This occasionally causes problems: the line-width illusion, for instance, has been attributed to implicit 3D perception of two-dimensional stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams [@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013].
+This occasionally causes problems: the line-width illusion, for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams [@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013].
 
 The use of 3D graphics have been explored in multiple studies. 
-@fisher_data_1997 explored the preference of using either 2D or 3D graphs and found that subjects tended to use simpler 2D graphs when tasked with extracting information.
-@barfield_effects_1989 compared 2D and 3D graphs presented on paper and on computers.
-Their results showed that the accuracy of subject answers depended on their skill level. 
-Novice subjects were more accurate with 2D paper graphs and experienced managers were more accurate with 3D computer graphs. 
-For both experience levels, participants were more confident in their answers using 2D graphs over other presentations of data. 
-There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may better encode information.
-@brath_3d_2014 highlights the intrinsic attributes of 3D graphs and its benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.
+@fisher_data_1997 explored user preference for 2D or 3D charts and found that subjects tended to prefer simpler 2D graphs when tasked with extracting information. 
+@barfield_effects_1989 compared 2D and 3D graphs presented on paper and on computers, showing that the accuracy of subject answers depended on their skill level: novice subjects were more accurate with 2D paper graphs, while experienced managers were more accurate with 3D computer graphs.
+For both experience levels, participants were more confident in their answers when using 2D graphs. 
+There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may better encode information - for instance, when X and Y are used to represent spatial dimensions, it may be preferable to use a 3D chart to convey numerical information instead of using color, which is perceived much less accurately.
+@brath_3d_2014 highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.
 
 There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data. 
-However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since 2014. 
+However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.
 Digital graphics has developed quickly, along with the hardware necessary to support these software developments. 
-As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving the artificial charts into a more natural setting.
-As a result, it is reasonable to reconsider the use of 3D charts, not only because of technological developments, but because these charts provide the opportunity to make visual graphics accessible to those with limited or absent vision.
+As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
+As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision.
 
-Here, we provide the process of replication and modernization of testing perceptual judgments to 2D graphs, 3D graphs projected in 2D environments, and 3D printed bar graphs.
-
-## Selected Components from Cleveland and McGill
-
-Cleveland and McGill provided a theory and tested for the ordering of perceptual importance for the elements of length, position, and angle. 
-Their first experiment, referenced as the position-length experiment, used five types of bar charts.
-Two of these were grouped bar charts and the other three were stacked bar charts.
-Each chart had two bars used for comparison and participants were asked to determine which bar was smaller and give their perceived ratio of the smaller bar to the larger bar.
-The two grouped bar charts are for the perceptual element of position along a common scale, where one has zero distance between bars and the other has a fixed distance between bars.
-These grouped bar charts will be referenced as adjacent and separated graph types in this paper, respectively.
-
-Our study replicates the procedure for the comparisons of the two grouped bar charts, but with an objective of detecting differences in accuracy between 2D graphs, 3D digital graphs, and 3D printed graphs.
+In this paper, we discuss a study designed to examine Cleveland & McGill's experiments on grouped bar charts using modern graphics in two and three dimensions.
+This study lays the groundwork for additional empirical studies on the use of 3D graphics, rendered and 3D printed, for visualizing complex data.
+In order to explore perception of fully 3D graphics, it is prudent to start with the simplest possible 3D graphic: one in which the third dimension is not necessary, so that we can easily compare to two-dimensional representations without loss of information.
+<!-- This is included because of a graphics paper section comment that really annoyed me -->
+In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from @clevelandGraphics1984 in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs.
+The next section presents the results, and we conclude the paper by discussing this experiment in the context of existing work on the perception of 2D and 3D graphical elements.
 
@@
 ```{r plotTypes, fig.cap = "Two dimensional, two-dimensional digital rendering, and 3D-printed charts used in this study.", out.width = "100%"}
-knitr::include_graphics("plot-types.png", dpi = 300)
+knitr::include_graphics("_images/plot-types.png", dpi = 300)
 ```
@@
 A total of 21 kits were created to include all combinations of the five ratios."}
-knitr::include_graphics("study-design.png")
+knitr::include_graphics("_images/study-design.png")
 ```
@@
 
-One interesting facet of @cleveland_graphical_1984 is the participant recruitment methodology: "For each experiment the subjects fell into two categories: (1) a group of females, mostly housewives, without substantial technical experience; (2) a mixture of males and females with substantial technical training and working in technical jobs.
+One interesting facet of @clevelandGraphical1984 is the participant recruitment methodology: "For each experiment the subjects fell into two categories: (1) a group of females, mostly housewives, without substantial technical experience; (2) a mixture of males and females with substantial technical training and working in technical jobs.
 Most of the subjects in the position-length experiment participated in the position-angle experiment; in all cases repeat subjects judged the position-angle graphs first." 
@@
 Participants could practice with the sliders and preview the questions that would be asked as part of the task.", out.width = "80%"}
-knitr::include_graphics("03-Practice-2.png")
+knitr::include_graphics("_images/03-Practice-2.png")
 ```
@@
 Participants were asked to select which bar (circle or triangle) was smaller, and then to estimate the ratio of the smaller bar to the larger bar.", out.width = "80%"}
-knitr::include_graphics("05-Experiment-05-filled-in.png")
+knitr::include_graphics("_images/05-Experiment-05-filled-in.png")
 ```
@@
 ```
-This is somewhat different than the results in @cleveland_graphical_1984 and @heerCrowdsourcingGraphicalPerception2010b; in both cases the midmean log absolute errors increased until about 55% of the true proportional difference and then decreased.
+This is somewhat different than the results in @clevelandGraphical1984 and @heerCrowdsourcingGraphicalPerception2010b; in both cases the midmean log absolute errors increased until about 55% of the true proportional difference and then decreased.
 It is possible that this difference is due to the fact that we did not require an explicit numerical estimate of the ratio but instead asked participants to indicate the ratio on a slider (which is essentially a number line).
@@
 The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger.
-However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @cleveland_graphical_1984 do not include significance testing beyond the graphical display of confidence intervals.
+However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @clevelandGraphical1984 do not include significance testing beyond the graphical display of confidence intervals.
 

### 2024-01-17 jds.rmd
Updates and add refs and code for sine illusion

@@
 Each chart had two bars marked for comparison; participants were asked to determine which bar was smaller and give the  perceived ratio of the smaller bar to the larger bar.
-@fig-bar-types show the two types of grouped bar charts.
+@fig:bar-types show the two types of grouped bar charts.
 We are primarily interested in the grouped bar charts (in part because 3D printing is not yet inexpensive enough to make moderate-scale stacked bar chart experiments viable), which consisted of two comparison bars which were either adjacent or in separate groups.
@@
 Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
-This occasionally causes problems: the line-width illusion, for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams [@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013].
+This occasionally causes problems: the line-width illusion[@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013], for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in @fig:sine-illusion).
+While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception.
+Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the stimuli.
+
+```{r sine-illusion}
+#| cache: false
+#| fig-width: 4
+#| fig-height: 4
+#| out-width: "30%"
+#| fig-cap: "An illustration of the sine illusion\\citep{vanderplasSignsSineIllusion2015}, also known as the line-width illusion.
+All vertical lines in figures (a) and (c) are the same length, but the lines in the middle of the curve appear to be much shorter.
+The illusion results when implicit perceptual corrections useful for perceiving the size of objects with depth are applied to 2D stimuli with no actual depth."
+#| fig-subcap:
+#|    - "Sine Illusion"
+#|    - "3D surface with similar features"
+#|    - "Perceived vs. actual length"
+#' Function to data frame
+#'  
+#' @param n number of values 
+#' @param ell extent of vertical range
+#' @param x range of horizontal values
+#' @param f function
+#' @param fp first derivative of function f
+#' @param f2p second derivative of function f
+#' @return data frame with n function values and derivatives along the x axis for a range given by x
+#' @example
+#' f <- function(x) 2*sin(x)
+#' fp <- function(x) 2*cos(x)
+#' f2p <- function(x) -2*sin(x)
+#' dframe <- function.frame(n=50, ell=1, x=c(0,2*pi), f, fp, f2p)
+#' require(ggplot2)
+#' qplot(x, y=ystart, yend=yend, xend=x, geom="segment", data=dframe) +
+#'    geom_point(aes(x,y))
+function.frame <- function(n=200, ell=1, x, f, fprime, f2prime) {
+  x <- seq(x[1], x[2], length=n)
+  if (length(ell) != length(x)) ell <- rep(ell, length(x))
+  y <- f(x)
+  ystart <- y - ell/2
+  yend <- y + ell/2
+  # now correct for line illusion in vertical direction
+  dy <- diff(range(y))
+  dyl <- diff(range(c(ystart, yend)))
+  # fprime and f2prime are sensitive to the aspect ratio of a plot
+  # we represent it in framework of dy and dy+len
+  # needs to be fixed by factor a:  
+  a <- dy/dyl
+
+  fp <- a*fprime(x)
+  f2p <- a*f2prime(x)
+  data.frame(x,y,ystart, yend, fp, f2p, a)
+}
+
+#' This function has to be split in parts and renamed
+#' 
+#' Right now this function is a convoluted mess.  We need to have separate
+#' functions that 
+#' (1) create a data frame for a given function,
+#' its first and second derivatives and range in x
+#' (this is what the function function.frame is supposed to do), 
+#' (2) a trig transform function
+#' (3) a quadratic approach function - better even, make (2) and (3) one function and use a parameter to decide on the method to use.
+#' @example
+#' f <- function(x) 2*sin(x)
+#' fp <- function(x) 2*cos(x)
+#' f2p <- function(x) -2*sin(x)
+#' dframe <- createSine(n=50, ell=1, x=c(0,2*pi), f, fp, f2p)
+#' require(ggplot2)
+#' qplot(x=x, xend=x, y=y+ellx4.u, yend=y-ellx4.l, geom="segment", data=dframe, linetype=I(1)) + 
+#'    theme_bw() + coord_fixed(ratio=1) + 
+#'    xlab("x") + ylab("y") + 
+#'    scale_x_continuous(breaks=seq(0, 2*pi, by=pi/2), minor_breaks=minor.axis.correction,
+#'      labels=c("0", expression(paste(pi,"/2")), expression(pi), expression(paste("3",pi, "/2")), 
+#'      expression(paste("2",pi))))
+
+createSine <- function(n=200, len=1, f=f, fprime=fprime, f2prime=f2prime, a=0, b=2*pi) {
+  #  if(getquadapprox & !is.function(f2prime)) f2prime <- function(x) -1*f(x) # for backwards compatibility
+  x <- seq(a, b, length=n+2)[(2:(n+1))]
+  ell <- rep(len, length=length(x))
+  fx <- f(x)
+  ystart <- fx - .5*ell
+  yend <- fx + .5*ell
+
+  # now correct for line illusion in vertical direction
+  dy <- diff(range(fx))
+  dx <- diff(range(x))
+  # fprime works in framework of dx and dy, but we represent it in framework of dx and dy+len
+  # needs to be fixed by factor a:  
+#  a <- dy/(dy + len) 
+  dyl <- diff(range(c(ystart, yend)))
+  a <- dy/(dyl) 
+  # ellx is based on the "trig" correction
+  ellx <- ell / cos(atan(abs(a*fprime(x))))
+  # ellx2 is based on linear approximation of f  
+  ellx2 <- ell * sqrt(1 + a^2*fprime(x)^2)
+
+  # make this a data frame - ggplot2 doesn't do well with floating vectors
+  dframe <- data.frame(x=x, xstart=x, xend=x, y=fx, ystart=ystart, yend=yend, ell=ell, ellx = ellx, ellx2=ellx2)
+
+  # third adjustment is based on quadratic approximation of f.
+  # this needs two parts: correction above and below f(x)  
+  #   if(getquadapprox & is.function(f2prime)){
+  #     secseg <- do.call("rbind", lapply(dframe$x, function(i) getSecantSegment(i, dframe, f, fprime, f2prime)))
+  #     dframe$ellx3.u <- secseg$sec.ell1
+  #     dframe$ellx3.l <- secseg$sec.ell2
+  #   }
+
+  fp <- a*fprime(x)
+  f2p <- a*f2prime(x)
+  v <- 1 + fp^2
+  lambdapinv <- 0.5*(sqrt(v^2-f2p*fp^2*ell) + v)    
+#  lambdaminv <- -0.5*(sqrt(v^2+f2p*fp^2*ell) + v)
+  lambdaminv <- 0.5*(sqrt(v^2+f2p*fp^2*ell) + v)
+
+  dframe$ellx4.l <- 0.5*abs(lambdapinv)/sqrt(v)
+  dframe$ellx4.u <- 0.5*abs(lambdaminv)/sqrt(v)
+#  dframe$lambdam <- 1/lambdaminv
+#  dframe$lambdap <- 1/lambdapinv
+  fp <- fprime(x)
+  f2p <- f2prime(x)
+  v <- 1 + fp^2
+#  lambda1 <- (-v + sqrt(v^2 - fp^2*f2p*ell))/(fp^2*f2p) # these two work
+#  lambda2 <- (-v + sqrt(v^2 + fp^2*f2p*ell))/(fp^2*f2p) # these two work
+
+  lambdaa <- -2*(sqrt(v^2 - f2p*fp^2*ell) + v)^-1 # identical to top
+  lambdab <- 2*(sqrt(v^2 + f2p*fp^2*ell) + v)^-1
+
+  dframe$lambdam <- lambdaa
+  dframe$lambdap <- lambdab
+
+  # qplot(a*(sqrt(1 + a^2*fprime(x)^2)*abs(lambdab))^-1, ellx4.u, data=dframe)
+  # qplot(a*(sqrt(1 + a^2*fprime(x)^2)*abs(lambdaa))^-1, ellx4.l, data=dframe)
+  # ggplot(aes(x, y+ell*(a*sqrt(v)*abs(lambdaa))^-1, xend=x, yend=y-ell*(a*sqrt(v)*abs(lambdab))^-1), data=dframe) + geom_segment()
+
+  dframe
+}
+
+#' what is the function doing? - in one sentence
+#' 
+#' function description in a paragraph
+#' @param x0 is a vector of locations for which the secant segments are supposed to be calculated
+#' @param df df is a data frame
+#' @param f function
+#' @param fp first derivative of function f
+#' @param f2p second derivative of function f
+#' @return data frame consisting of .... what is the output of the function?
+#' @example
+#' # need an example here of how to use the function
+getSecantSegment <- function(x0, df, f, fprime, f2prime){
+# find the closest values of the grid df corresponding to x0
+  # doesn't work right now
+  ell.x0 <- sapply(x0, function(x) df$ell[which.min(x-df$x)])
+
+  dy <- diff(range(with(df, c(y+ell.x0, y-ell.x0))))
+  dx <- diff(range(df$x))
+  a <- dx/dy
+
+  #---- Actual Roots
+  #   secSlope   <- -a/fprime(x0)
+  #   temp       <- seq(min(df$x)-pi, max(df$x)+pi, .0001)
+  #   leftend    <- temp[which.min(abs(f(temp) + ell.x1 - secSlope*(temp-x1)))]
+  #   rightend   <- temp[which.min(abs(f(temp) - ell.x1 - secSlope*(temp-x1)))]
+  #----
+
+  #---- Approximation
+  lambda1 <- (-(fprime(x0)^2 + 1) + sqrt((fprime(x0)^2 + 1)^2 - 2*fprime(x0)^2*f2prime(x0)*ell.x0))/(fprime(x0)^2*f2prime(x0))
+  lambda2 <- (-(fprime(x0)^2 + 1) + sqrt((fprime(x0)^2 + 1)^2 + 2*fprime(x0)^2*f2prime(x0)*ell.x0))/(fprime(x0)^2*f2prime(x0))
+
+  x1 <- lambda1*fprime(x0)+x0
+  x2 <- lambda2*fprime(x0)+x0
+  y1 <- f(x0)-lambda1
+  y2 <- f(x0)-lambda2
+  #----
+
+  #---- Approximation V2 ----
+  #   dx <- fprime(x0)
+  #   dx1 <- 1+dx^2   
+  #   ddx <- f2prime(x0)
+  #   
+  #   corr1 <- 1/a*(((dx1)-sqrt(dx1^2-2*ddx*(dx1-1)^2*ell.x0))/(ddx*dx))
+  #   corr2 <- 1/a*(((dx1)-sqrt(dx1^2+2*ddx*(dx1-1)^2*ell.x0))/(ddx*dx))
+  #   if(abs(dx)<.01) {
+  #     dx1 <- 2
+  #     corr1 <- corr2 <- 0
+  #   }
+  #   if(abs(dx)<.01){
+  #     ddx <- 1
+  #   }
+  #   
+  #   x1 = corr1 + x0
+  #   x2 = corr2 + x0
+  #   y1 = f(x1)+ell.x0
+  #   y2 = f(x2)-ell.x0
+  #----
+
+  df2 <- data.frame(x=x0, y=f(x0), deriv=fprime(x0),
+                    sec.xstart=x1, sec.xend = x2, 
+                    sec.ystart=y1, sec.yend = y2,
+                    ell.orig = 2*ell.x0)
+
+  df2$sec.ell1 <- a*.5*ell.x0/with(df2, sqrt((sec.yend-y)^2+(sec.xend-x)^2))
+  df2$sec.ell2 <- a*.5*ell.x0/with(df2, sqrt((y-sec.ystart)^2+(x-sec.xstart)^2))
+  #   df2$sec.ell1 <- with(df2, sqrt((sec.yend-y)^2+(sec.xend-x)^2))
+  #   df2$sec.ell2 <- with(df2, sqrt((y-sec.ystart)^2+(x-sec.xstart)^2))
+  df2$type <- "Perceived Width"
+  df2$a <- a
+  return(df2)
+}
+
+correctx <- function(z, fprime, a=0, b=2*pi, w=1) {
+  # w = 1/(shrink+1)
+  const <- integrate(function(x) abs(fprime(x)), a, b)$value
+  trans <- sapply(z, function(i) integrate(function(x) abs(fprime(x)), a, i)$value*(b-a)/const + a)
+  # alternatively to the rowMeans, you could report back  
+  # trans*(1-w) + z*w
+  trans*w + z*(1-w)
+}
+
+f <- function(x) 2*sin(x)
+fprime <- function(x) 2*cos(x)
+f2prime <- function(x) -2*sin(x)
+
+x <- seq(0, 2*pi, length=42)[2:41]
+data <- do.call("rbind", lapply(seq(-.5, .5, 1), function(i) data.frame(x=x, y=3*sin(x), z=i)))
+
+data.persp <- reshape2::acast(data, x~z, value.var="y")
+x <- sort(unique(data$x))
+y <- sort(unique(data$y))
+z <- sort(unique(data$z))
+dframe <- createSine(n = 50, len = 1, f=f, fprime=fprime, f2prime=f2prime)
+dframe %>%
+  ggplot(aes(x = xstart, xend = xend, y = ystart, yend = yend)) + geom_segment() + coord_fixed() + 
+  theme_void()
+
+persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1.75, 1.75), scale=FALSE, box=FALSE, expand=3/pi, d=50) # , ltheta=0, lphi=-15
+
+# persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1.75, 1.75), scale=FALSE, box=FALSE, expand=3/pi, d=3) # , ltheta=0, lphi=-15
+
+dframe <- createSine(n = 150, len = 1, f=f, fprime=fprime, f2prime=f2prime)
+dframe$ystartcts <- dframe$ystart
+dframe$yendcts <- dframe$yend
+dframe[1:150,c(2, 3, 5, 6)] <- NA
+dframe[(1:15)*10-5, c(2, 3)] <- dframe[(1:15)*10-5, 1] 
+dframe[(1:15)*10-5, 5] <- dframe[(1:15)*10-5, 4] - .5
+dframe[(1:15)*10-5, 6] <- dframe[(1:15)*10-5, 4] + .5
+dframe$type <- "Vertical Width"
+
+idx <- which(!is.na(dframe$xstart))
+dframe$ell <- dframe$ell/2
+dframe.1 <- getSecantSegment(dframe$xstart[idx], dframe, f, fprime, f2prime)
+dframe.1$x <- dframe$x[idx]
+dframe.1$y <- dframe$y[idx]
+dframe.1$ystartcts <- dframe$ystartcts[idx]
+dframe.1$yendcts <- dframe$yendcts[idx]
+names(dframe.1) <- c("x", "y", "deriv", "xstart", "xend", "ystart", "yend", "ell", "ell.quad1", "ell.quad2", "type", "a", "ystartcts", "yendcts")
+dframe.1$vangle <- with(dframe.1, atan(deriv))
+dframe <- bind_rows(dframe, dframe.1)
+dframe$type <- factor(dframe$type)
+
+ggplot(aes(x=x, y=y), data=dframe, colour=I("grey50")) + 
+  geom_line(color = "grey") + theme_void() + 
+  geom_line(aes(y=ystartcts), colour="grey50", alpha = .25) +
+  geom_line(aes(y=yendcts), colour="grey50", alpha = .25) +
+  geom_segment(data=subset(dframe, !is.na(type)), 
+               aes(x=xstart, xend = xend, y=ystart, yend=yend, colour=type, linetype=type), linewidth=0.8)  + 
+  xlab("") + ylab("")  +
+  coord_equal(ratio=1) + scale_colour_manual("", values=c("blue", "grey30")) + 
+  # geom_text(aes(label=paste("theta", "%~~%", round(abs(vangle)/pi*180), "^o", sep=""), 
+                # x=pmax(xstart, xend)/2+x/2+.35 , y=y-sign(vangle)*.6+.02), colour="blue",
+            # data=dframe.1, parse=TRUE, hjust=.9, vjust=.5, size=3) + 
+  scale_x_continuous(breaks=seq(0, 2*pi, by=pi/2), 
+                     labels=c("0", expression(paste(pi,"/2")), expression(pi), 
+                              expression(paste("3",pi, "/2")), expression(paste("2",pi)))) +
+  scale_linetype_manual("", values=c("11", "solid")) + 
+  theme(legend.key.width = unit(3, "line"), plot.margin = unit(c(0,0,0,0), "cm"), legend.position=c(.25, .2))
+
+```
 
@@
 As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
-As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision.
+As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].
 
@@
 
-Our study is designed to replicate and expand upon the position-length experiment from Cleveland and McGill as closely as possible.
-In this section, we will discuss the replication process and the design of our modified version of this experiment.
+Our study is designed to replicate and expand upon the position-length experiment from Cleveland & McGill as closely as possible, with the hope of being able to integrate this study with previously reported results in a consistent manner.
+In this section, we discuss the replication process and the design of our version of this experiment.
 
@@
 The exact numeric comparisons were not disclosed, but the comparison values used in our study were subjected to the constraints of having the same ratio values and that no value was used more than twice.
+\svp{Can we show the comparison values in a table?}
 
-Each graph is presented so that there are ten bars where only two of the bars are marked for identification.
+Each graph is presented so that there are ten bars where only two of the bars are marked for identification, \svp{as shown in @fig:bar-types.}
 Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.

### 2024-01-18 jds.rmd
compiled pdf

@@
 ```{r sine-illusion}
+#| message: false
+#| error: false
+#| warning: false
 #| cache: false
@@
 #| fig-cap: "An illustration of the sine illusion\\citep{vanderplasSignsSineIllusion2015}, also known as the line-width illusion.
-All vertical lines in figures (a) and (c) are the same length, but the lines in the middle of the curve appear to be much shorter.
-The illusion results when implicit perceptual corrections useful for perceiving the size of objects with depth are applied to 2D stimuli with no actual depth."
+All vertical lines are the same length, but the lines in the middle of the curve appear to be much shorter.
+The illusion results when implicit perceptual corrections useful for perceiving the size of objects with depth are applied to 2D stimuli with no actual depth.
+As 3D heuristics occasionally cause inaccurate communication when applied to 2D objects, it is reasonable to think that there might be some situations where charts making use of a realistic third dimension might be perceived more accurately than their 2D equivalents."
 #| fig-subcap:
@@
 The 3D renderings and 3D printed charts were both created using OpenSCAD [@kintelOpenSCADDocumentation2023], which creates STL files from markup describing the object's geometric composition. 
-Charts were composed of a platform, raised labels for the A and B groups of bars, circle and triangle markers indicating the bars of interest, and the bars themselves; values for the bar heights were inserted into the markup using R [@R].
-In addition, an ID code was engraved into the bottom of the platform to uniquely identify each object; this allows the researchers to ensure that the 3D printed charts are correctly allocated to stimulus sets.
+Charts were composed of a platform, with raised text labels centered in front of the A and B groups of bars.
+Bars were created by inserting bar heights into the OpenSCAD template using R, and then raised circle and triangle markers were positioned on top of the bars to be compared to provide a tactile and visual indicator.
+In addition, an ID code was engraved into the bottom of the platform to uniquely identify each object; this allowed the researchers to ensure that each kit contained the correct charts throughout the experiment while minimizing the risk of stickers or other identifiers becoming detached from the charts.
+As printed, the base of each chart was 13cm x 3cm x 1cm, with the highest bar rising 9.5cm above the chart base.
+Raised letters and shapes were 2mm above the base or bar, respectively.
+The color of the filament used to print each chart corresponded to the estimation ratio (though this was not obvious to participants), allowing the researchers to quickly determine that a kit of 3D charts contained no ratio duplication.
+Colors were randomly assigned to ratios so that participants could not gain information about the ratio from any perceived ordering of the colors.
 
-Digital renderings of the generated STL files were created using @rgl, which integrates into @shiny using the @mozillafoundationWebGL2D3D2023 extension.
-Rendered 3D charts were initially angled corresponding to the default 3D bar charts present in Microsoft Excel, but WebGL's interactivity allows the user to rotate, scale, and otherwise interact with the chart to change the angle.
-3D renderings were colored to correspond to the 3D printed chart filament color.
+Digital renderings of the generated STL files were created using RGL [@rgl], which integrates into Shiny [@shiny] using the WebGL [@mozillafoundationWebGL2D3D2023] extension.
+The RGL rendering of the STL file was initially angled to correspond to the perspective of default 3D bar charts present in Microsoft Excel, but in the WebGL interactive environment, participants could rotate the charts as they pleased to create a useful estimate of the bar ratio.
+Rendered charts were colored to correspond with the filament color used to print the physical chart for consistency.
 The default `rgl` lighting was replaced with three lights located in fixed positions around the rendered figure. 
 The lights were positioned so that one was behind the rendered figure<!--(x: 910, y: -1000, z:1000)-->, another in front of the figure<!--(x: 780, y: 1030, z: 1000)-->, and one light below the figure<!--(x: 65, y: 15, z: -100)-->.
-3D charts were printed with colored filament corresponding to a specific ratio comparison; this allowed researchers to visually assess kits to ensure that they contained five unique ratios.
-Colors corresponding to each ratio were assigned randomly to ensure that chart color provided no useful information about the ratio value.
-As printed, the base of the chart was 13cm x 3cm x 1cm, with the highest bar rising 9.5cm above the chart base.
-Raised letters and shapes were 2mm above the base or bar, respectively.
+This arrangement provided a consistent visual experience and minimized shadows that prevented visual estimation of the bars.
 
@@
 
-Participants were provided a kit with five 3D-printed charts, comprising five of the seven unique ratio comparisons; we then used the kit ID to ensure that participants saw computer-rendered charts with ratios corresponding to those in the kit of physical charts.
+Participants were provided a kit with five 3D-printed charts corresponding to five of the seven unique ratio comparisons.
+In the Shiny application, we asked participants for the kit ID, which was used to look up the corresponding digital renderings.
 This ensured that the experimental design was balanced across chart type and randomized with respect to the type of comparison (adjacent or separated).
-We printed 21 kits containing five 3d-printed charts each.
+We printed 21 kits containing five 3d-printed charts each, where 2 of the 7 ratios were randomly removed from the total set.
 <!-- There are 21 kits in total so that each combination of ratios is accounted for. -->
@@
 Figure \@ref(fig:midmeans-log-errors) shows the midmeans of the log absolute errors compared to the true ratio of the bars for each graph type and comparison type. 
-The results tend to indicate that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types.
+The results suggest that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types. In addition, there is a tendency
 
-```{r midmeans-log-errors, fig.cap='Midmeans of log absolute errors for the true ratio of bars. Each overlaying line are computed with loess.', message=FALSE, warning=FALSE, out.width='80%'}
+```{r midmeans-log-errors, fig.cap='Midmeans and observed values of log absolute errors for the true ratio of bars.
+Summary lines are computed from raw data using a loess smooth.', message=FALSE, warning=FALSE, out.width='80%'}
 plot.types = c('2D', 'Rendered 3D', '3D Printed')
@@
 
-res %>% 
+tmp <- res %>% 
   group_by(Height, graphtype, plot) %>% 
   summarize(midmean = mean(response, trim=0.25, na.rm = T)) %>%
-  ggplot(mapping = aes(x = Height, y = midmean,
-                       color = graphtype)) +
-  geom_point() +
-  geom_smooth(se = F, alpha=1/2) +
+  mutate(type = "Midmean") %>%
+  rename(response = midmean)
+
+res %>% 
+  mutate(type = "Raw Data") %>%
+  bind_rows(tmp) %>%
+  # group_by(Height, graphtype, plot) %>% 
+  ggplot(mapping = aes(x = Height*100, y = response,
+                       color = graphtype, fill = graphtype)) +
+  geom_smooth(se = T, alpha=.125,  expand = F, method = "loess") +
+  geom_point(aes(shape = type, alpha = type, size = type)) +
+  scale_shape_manual("Data", values = c(19, 1)) + 
+  scale_size_manual("Data", values = c(2, 1)) + 
+  scale_alpha_manual("Data", values = c(1, .25)) + 
   scale_color_discrete(labels = c('Adjacent', 'Separated')) +
+  scale_x_continuous(limits = c(0, 100)) + 
   facet_wrap(~plot, labeller = labeller(plot = plot.types)) +
@@
        x = 'True Proportional Difference (%)',
-       y = 'Midmeans of Log Error',
+       y = 'Log Error',
        color = 'Comparison Type') +

### 2024-01-22 initialsubmission.tex
Updates and Reviewer 1 corrections

\documentclass[letterpaper,inpress,dvipsnames]{jdsart}

\setcounter{page}{1}
\pubmonth{July}
\pubyear{2022}
\volume{xx}
\issue{xx}
\doi{0000}

\usepackage[utf8]{inputenc}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{amsfonts,amsmath,amssymb,amsthm} 
\usepackage{booktabs} 
\usepackage{lipsum} 
\PassOptionsToPackage{dvipsnames}{xcolor} 
\newcommand{\svp}[1]{{\textcolor{ForestGreen}{#1}}} 
\newcommand{\tw}[1]{{\textcolor{Red}{#1}}}

\begin{document}
\begin{frontmatter}

\title{Evaluating Perceptual Judgements on 3D Printed Bar Charts}
\runtitle{Perceptual Judgements on 3D Printed Bar Charts}

\author[1]{
  \inits{T.}
  \fnms{Tyler}
  \snm{Wiederich}  \thanksref{1}  \ead{twiederich2@huskers.unl.edu}}
\author[1]{
  \inits{S.}
  \fnms{Susan}
  \snm{VanderPlas}  \ead{susan.vanderplas@unl.edu}}

\thankstext[type=corresp,id=1]{Tyler Wiederich}
\address[1]{Department of Statistics, 
  \institution{University of Nebraska-Lincoln}, \cny{United States of America}}

\begin{abstract}
The use of 3D data visualizations has limitations when the third dimension does not convey any additional information to the viewer. Numerous studies advocate to avoid these types of graphs whenever possible, but these studies are almost entirely focused on the 2D projections of the 3D graphs. This paper describes the partial replication of a well-known paper in data visualization and its adaptation to focus on 3D printed bar charts. While current results from our study do not show differences between 2D and 3D graphs, we will use this study to provide a baseline for future studies dealing with graphics in 3D environments.
\end{abstract}

\begin{keywords}
\kwd{graphics}\kwd{3D bar charts}\kwd{3D printing}.
\end{keywords}

\end{frontmatter}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{elementary-graphical-tasks}{%
\subsection{Elementary Graphical Tasks}\label{elementary-graphical-tasks}}

\citet{cleveland_graphical_1984} published a paper that sets up the foundation for how accurately people extract quantitative information using elementary perceptual tasks (EPTs).
Elementary Perceptual Tasks, according to these experiments, include assessing graphical elements such as position along a common scale, length, angle, and volume, and estimating the corresponding numerical value of these representations.
They found that estimates of ratios between positions along a common scale had smaller log errors than estimates of ratios between lengths. The study relied entirely on estimation accuracy, which may not always be relevant when extracting information from graphs. For example, estimation is less relevant when ordering values by size.
\citet{heerCrowdsourcingGraphicalPerception2010b} replicated some parts of \citet{cleveland_graphical_1984} in an online setting using Mechanical Turk, largely replicating the results of the original study while demonstrating the utility of the Mechanical Turk platform.

Further studies have shown that 3D graphs are less accurate at portraying numeric information that 2D graphs \citep{barfield_effects_1989, fisher_data_1997}. In certain contexts and conditions, there is some research suggesting that 3D graphs may better encode information \citep{brath_3d_2014}.

\hypertarget{d-graphical-perception}{%
\subsection{3D Graphical Perception}\label{d-graphical-perception}}

Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
This occasionally causes problems: the line-width illusion, for instance, has been attributed to implicit 3D perception of two-dimensional stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams \citep{vanderplasSignsSineIllusion2015, daySineIllusion1991, hofmannCommonAnglePlots2013}.

The use of 3D graphics have been explored in multiple studies.
\citet{fisher_data_1997} explored the preference of using either 2D or 3D graphs and found that subjects tended to use simpler 2D graphs when tasked with extracting information. \citet{barfield_effects_1989} compared 2D and 3D graphs presented on paper and on computers.
Their results showed that the accuracy of subject answers depended on their skill level.
Novice subjects were more accurate with 2D paper graphs and experienced managers were more accurate with 3D computer graphs.
For both experience levels, participants were more confident in their answers using 2D graphs over other presentations of data.
There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may better encode information.
\citet{brath_3d_2014} highlights the intrinsic attributes of 3D graphs and its benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data.
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since 2014.
Digital graphics has developed quickly, along with the hardware necessary to support these software developments.
As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving the artificial charts into a more natural setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of technological developments, but because these charts provide the opportunity to make visual graphics accessible to those with limited or absent vision.

Here, we provide the process of replication and modernization of testing perceptual judgments to 2D graphs, 3D graphs projected in 2D environments, and 3D printed bar graphs.

\hypertarget{selected-components-from-cleveland-and-mcgill}{%
\subsection{Selected Components from Cleveland and McGill}\label{selected-components-from-cleveland-and-mcgill}}

Cleveland and McGill provided a theory and tested for the ordering of perceptual importance for the elements of length, position, and angle.
Their first experiment, referenced as the position-length experiment, used five types of bar charts.
Two of these were grouped bar charts and the other three were stacked bar charts.
Each chart had two bars used for comparison and participants were asked to determine which bar was smaller and give their perceived ratio of the smaller bar to the larger bar.
The two grouped bar charts are for the perceptual element of position along a common scale, where one has zero distance between bars and the other has a fixed distance between bars.
These grouped bar charts will be referenced as adjacent and separated graph types in this paper, respectively.

Our study replicates the procedure for the comparisons of the two grouped bar charts, but with an objective of detecting differences in accuracy between 2D graphs, 3D digital graphs, and 3D printed graphs.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

Our study is designed to replicate and expand upon the position-length experiment from Cleveland and McGill as closely as possible.
In this section, we will discuss the replication process and the design of our modified version of this experiment.

\hypertarget{replicating-cleveland-and-mcgill}{%
\subsection{Replicating Cleveland and McGill}\label{replicating-cleveland-and-mcgill}}

The first step of replicating the position-length experiment was to determine the heights of the bars that participants use for comparisons.
These values for the bar heights are linear on a log scale and are given by

\[s_i=10\cdot 10^{(i-1)/12}, \qquad i=1,...,10\]

Each graph presents two bars from the values given above where the participants are asked to judge the ratio of the smaller bar to the larger bar. The ratio of bars used by Cleveland and McGill were 17.8, 26.1, 38.3, 46.4 (twice), 56.2, 68.1 (twice), and 82.5 (twice).
The exact numeric comparisons were not disclosed, but the comparison values used in our study were subjected to the constraints of having the same ratio values and that no value was used more than twice.

Each graph is presented so that there are ten bars where only two of the bars are marked for identification.
Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.
Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at \url{https://github.com/TWiedRW/2023-JDS-3dcharts}.

\hypertarget{stimuli-construction}{%
\subsection{Stimuli Construction}\label{stimuli-construction}}

The graphs share a common layout across all formats, where two groupings of five bars are identified by ``A'' and ``B'', respectively, and circles and triangles are used to identify the bars participants should compare. Example graphs are shown in
Figure \ref{fig:plotTypes}.
There are some graphical elements that cannot be easily portrayed via 3D printing. For this reason, all graph types do not have axes, grid lines, or floating titles.

\begin{figure}
\includegraphics[width=1\linewidth]{_images/plot-types} \caption{Two dimensional, two-dimensional digital rendering, and 3D-printed charts used in this study.}\label{fig:plotTypes}
\end{figure}

The ggplot2 \citep{ggplot2} package was utilized to create the 2D bar charts.
The scale axis was removed, leaving only the bars and a bar grouping identifier.
The bars used for comparisons had the identifying mark at a height of 5 out of 100 for the 2D plots, and the 3D plots had the identifying marks on top of the bars.

The 3D renderings and 3D printed charts were both created using OpenSCAD \citep{kintelOpenSCADDocumentation2023}, which creates STL files from markup describing the object's geometric composition.
Charts were composed of a platform, raised labels for the A and B groups of bars, circle and triangle markers indicating the bars of interest, and the bars themselves; values for the bar heights were inserted into the markup using R \citep{R}.
In addition, an ID code was engraved into the bottom of the platform to uniquely identify each object; this allows the researchers to ensure that the 3D printed charts are correctly allocated to stimulus sets.

Digital renderings of the generated STL files were created using \citet{rgl}, which integrates into \citet{shiny} using the \citet{mozillafoundationWebGL2D3D2023} extension.
Rendered 3D charts were initially angled corresponding to the default 3D bar charts present in Microsoft Excel, but WebGL's interactivity allows the user to rotate, scale, and otherwise interact with the chart to change the angle. 3D renderings were colored to correspond to the 3D printed chart filament color.
The default \texttt{rgl} lighting was replaced with three lights located in fixed positions around the rendered figure.
The lights were positioned so that one was behind the rendered figure, another in front of the figure, and one light below the figure.
3D charts were printed with colored filament corresponding to a specific ratio comparison; this allowed researchers to visually assess kits to ensure that they contained five unique ratios.
Colors corresponding to each ratio were assigned randomly to ensure that chart color provided no useful information about the ratio value.
As printed, the base of the chart was 13cm x 3cm x 1cm, with the highest bar rising 9.5cm above the chart base.
Raised letters and shapes were 2mm above the base or bar, respectively.

\hypertarget{experiment-design}{%
\subsection{Experiment Design}\label{experiment-design}}

Participants were provided a kit with five 3D-printed charts, comprising five of the seven unique ratio comparisons; we then used the kit ID to ensure that participants saw computer-rendered charts with ratios corresponding to those in the kit of physical charts.
This ensured that the experimental design was balanced across chart type and randomized with respect to the type of comparison (adjacent or separated).
We printed 21 kits containing five 3d-printed charts each.

\begin{figure}
\includegraphics[width=6.4in]{_images/study-design} \caption{A graphical representation of the study design. Only five of the seven ratios were used in each kit where the ratios are randomly selected. A total of 21 kits were created to include all combinations of the five ratios.}\label{fig:studyDesign}
\end{figure}

\hypertarget{participant-recruitment}{%
\subsection{Participant Recruitment}\label{participant-recruitment}}

One interesting facet of \citet{cleveland_graphical_1984} is the participant recruitment methodology: ``For each experiment the subjects fell into two categories: (1) a group of females, mostly housewives, without substantial technical experience; (2) a mixture of males and females with substantial technical training and working in technical jobs.
Most of the subjects in the position-length experiment participated in the position-angle experiment; in all cases repeat subjects judged the position-angle graphs first.''
It would seem likely that the authors recruited individuals within their respective departments as well as their wives.
In the spirit of replicating the study, members of the UNL Statistics department and their spouses, partners, and roommates were asked to participate in our study; this replicates the spirit of the original study without the implicit assumptions that graduate students and professors are (1) largely male, (2) heterosexual, and (3) have unemployed partners.

A total of 48 participants completed the study; demographics are shown in \ref{fig:demographics}.

\begin{figure}
\includegraphics[width=0.85\linewidth]{jds_files/figure-latex/demographics-1} \caption{Demographic breakdown of participants in the study. All subjects were recruited from faculty and students in the statistics department at University of Nebraska-Lincoln.}\label{fig:demographics}
\end{figure}

\hypertarget{data-collection}{%
\subsection{Data Collection}\label{data-collection}}

A Shiny applet was used for data collection, along with the provided kit of 3D printed charts.
Participants provided informed consent through the applet, and then were asked for demographic information (age, gender, education level).
Then, participants were shown a ``practice'' page which allowed them to experiment with the data collection interface and practice estimating the ratio between the bars, as shown in Figure \ref{fig:practice}

\begin{figure}
\includegraphics[width=0.8\linewidth]{_images/03-Practice-2} \caption{Screenshot of Shiny application practice screen. Three 2D bar charts with different ratios were provided, along with sliders indicating the correct proportion. Participants could practice with the sliders and preview the questions that would be asked as part of the task.}\label{fig:practice}
\end{figure}

Directly before the experiment started, participants were asked to provide the kit ID, along with directions indicating that if the instructions indicated that participants should use a 3D chart for a task, the participant should select a chart from the kit, enter the ID code of that chart from the bottom of the object, and complete the requested task.
Participants were also instructed to make quick judgments for each graph and not to measure or estimate ratios using physical objects.

\begin{figure}
\includegraphics[width=0.8\linewidth]{_images/05-Experiment-05-filled-in} \caption{Screenshot of the applet collecting data for a 3D rendered chart task. Participants were asked to select which bar (circle or triangle) was smaller, and then to estimate the ratio of the smaller bar to the larger bar.}\label{fig:experiment3dRender}
\end{figure}

Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in Figure \ref{fig:experiment3dRender}.

\hypertarget{results}{%
\section{Results}\label{results}}

All responses that incorrectly identified the smaller bar were removed from the study before analysis.

\hypertarget{midmeans-of-log-absolute-errors}{%
\subsection{Midmeans of Log Absolute Errors}\label{midmeans-of-log-absolute-errors}}

Cleveland and McGill used
\[\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)\]
to measure accuracy of their participant's responses.
In their study, log base 2 seemed appropriate due to ``average relative errors changing by factors less than 10.''
They also added 1/8 to prevent distortions when the errors were close to zero.
\citet{heerCrowdsourcingGraphicalPerception2010b} followed the same analysis method, replicating many (but not all) of the results presented in the original paper.

Figure \ref{fig:midmeans-log-errors} shows the midmeans of the log absolute errors compared to the true ratio of the bars for each graph type and comparison type.
The results tend to indicate that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types.

\begin{figure}
\includegraphics[width=0.8\linewidth]{jds_files/figure-latex/midmeans-log-errors-1} \caption{Midmeans of log absolute errors for the true ratio of bars. Each overlaying line are computed with loess.}\label{fig:midmeans-log-errors}
\end{figure}

This is somewhat different than the results in \citet{cleveland_graphical_1984} and \citet{heerCrowdsourcingGraphicalPerception2010b}; in both cases the midmean log absolute errors increased until about 55\% of the true proportional difference and then decreased.
It is possible that this difference is due to the fact that we did not require an explicit numerical estimate of the ratio but instead asked participants to indicate the ratio on a slider (which is essentially a number line).
This should reduce the cognitive load required to transition from spatial comparison to numerical comparison and then to compute the proportion, but may also have impacted the results.

\hypertarget{linear-mixed-effects-model}{%
\subsection{Linear Mixed Effects Model}\label{linear-mixed-effects-model}}

In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a linear mixed effects model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio. This allows us to test for significant differences (though given the midmean log absolute error plots, we do not expect to find any) as well as to quantify effect sizes for future studies. The formal statistical model is as follows:

\[y_{ijklm}=\mu+S_i+R_j+G(R)_{(k)j}+T_l+\epsilon_{ijklm}\]

\noindent where

\begin{itemize}
\item
  \(y_{ijklm}=\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)\)
\item
  \(S_i\sim N(0,\sigma^2_S)\) is the effect of the \(i^{th}\) subject
\item
  \(R_j\) is the effect of the \(j^{th}\) ratio
\item
  \(G(R)_{(k)j}\) is the effect of the \(k^{th}\) graph type nested in the \(j^{th}\) ratio
\item
  \(T_l\) is the effect of the \(l^{th}\) comparison type
\item
  \(\epsilon_{ijklm}\sim N(0,\sigma^2_\epsilon)\) is the random error
\end{itemize}

No differences were detected for the true ratio of bars (p-value = .7881), whether the bars were adjacent or separated (p-value = .3375), or for the plot nested within the true ratio (p-value = .6868).
While Heer \& Bostock provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data at this time.

\hypertarget{discussion-and-future-work}{%
\section{Discussion and Future Work}\label{discussion-and-future-work}}

Previous work in 3D graphics would suggest that the errors for the 3D graphs would be larger than the errors for the 2D graphs.
While we did not find any significant results indicating that 3D graphs are read less accurately, there are two possibilities that might account for this discrepancy.

The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger. However, we should note that the analysis methods used in \citet{heerCrowdsourcingGraphicalPerception2010b} and \citet{cleveland_graphical_1984} do not include significance testing beyond the graphical display of confidence intervals.

The second possibility is more interesting: we examined 3D charts using rendered 3D graphs and 3D printed charts; both of these options allow for participants to interact with the chart, rotating it, and generally perceiving it as one might perceive any other 3D, real, object.
This is a far cry from the 3D perspective charts in the original study, which have a fixed angle and perspective and are thus not equivalent to our 3D charts.
Future studies should include an additional fixed 3D perspective bar chart, which will at least enable us to examine whether modern 3D rendering environments allow for more accurate conclusions than fixed 3D perspectives. Future iterations of this study will include ``traditional'' 3D graphs created by Microsoft Excel (that is, graphs with a fixed 3D perspective rendered in 2D). This option will allow us to examine fixed perspective 3D plots compared to 3D renderings and 2D plots; it will also enable online data collection in addition to the in-person data collection used in this experiment.

Another interesting aspect of our study is that the method used to record participant estimates is different from the method used in the original study as well as Heer \& Bostock's replication study.
A method similar to our slider input, marking position on a line, was used in \citet{spenceVisualPsychophysicsSimple1990}, but Spence asked participants to estimate \(A/(A+B)\), where we asked participants to estimate \(A/B\); thus, our results are still not directly comparable to previous studies.
We expect that the specific ratio estimated would also have an effect on observed participant errors.

The slider method for input of ratio estimates should be easier for participants, as it does not require explicit transformation to the numerical domain.
What is clear is that it would be beneficial to assess the impact of measurement method on participant errors directly, so that the results of these different studies might be explained and interpreted with regard both to the stimuli used and the measurement method employed in the experiment.
Future iterations of this experiment will likely address this estimation difference; such modifications in experimental design are relatively straightforward in Shiny and will provide useful insight into the design of future experiments evaluating the perception of statistical graphics.

\hypertarget{supplemental-material}{%
\subsection{Supplemental Material}\label{supplemental-material}}

Stimuli, code, and data for this experiment are provided at \url{https://github.com/TWiedRW/2023-JDS-3dcharts}.

\bibliography{JDS2023.bib}
\bibliographystyle{jds}

\end{document}

### 2024-01-22 jds.rmd
Updates and Reviewer 1 corrections

@@
   \usepackage{amsfonts,amsmath,amssymb,amsthm} 
+  \usepackage{threeparttable}
+  \usepackage{cleveref}
   \usepackage{booktabs} 
@@
 ```{r, include = F}
-knitr::opts_chunk$set(echo = F, dpi = 300)
+knitr::opts_chunk$set(echo = F, dpi = 300, message =F, warning = F, error = F, dev = "ragg_png")
 
@@
 
-res = results %>% 
+res_all = results %>% 
   left_join(trueRatios, by = 'fileID') %>% 
@@
   filter(subject %in% valid.users$subject) %>% 
-  arrange(appStartTime) %>% 
+  arrange(appStartTime)
+
+res <- res_all %>% 
   filter(whichIsSmaller == 'Triangle (▲)',
@@
 Each chart had two bars marked for comparison; participants were asked to determine which bar was smaller and give the  perceived ratio of the smaller bar to the larger bar.
-@fig:bar-types show the two types of grouped bar charts.
+\Cref{fig:bar-types} shows the two types of grouped bar charts.
 We are primarily interested in the grouped bar charts (in part because 3D printing is not yet inexpensive enough to make moderate-scale stacked bar chart experiments viable), which consisted of two comparison bars which were either adjacent or in separate groups.
@@
 ```{r bar-types}
-#| message: false
-#| error: false
-#| warning: false
 #| fig-width: 4
 #| fig-height: 4
-#| fig-cap: "@clevelandGraphical1984 used two different types of grouped bar charts, which are replicated here."
-#| fig-subcap: 
-#|    - "Adjacent"
-#|    - "Separated"
+#| fig-cap: "\\citet{clevelandGraphical1984} used two different types of grouped bar charts: comparisons between adjacent bars, and comparisons between separated bars.
+It is widely acknowledged that comparisons between separated bars (Type 3 comparisons, in Cleveland \\& McGill's terminology) are more difficult and error-prone."
+#| out-width: ".49\\linewidth"
+#| fig-show: "hold"
 #| layout-ncol: 2
 library(ggplot2)
-
+library(ragg)
 datasets$data[[1]] %>%
@@
   geom_col(position = "dodge", fill = "white", color = "black") + 
-  geom_text(aes(x = Group, y = 5, group = GroupOrder, label = IDchr), position = position_dodge(0.9)) + 
+  geom_point(aes(x = Group, y = 5, group = GroupOrder, shape = IDchr), size = 3, position = position_dodge(0.9)) + 
+  guides(shape = "none") + 
   theme_bw() + 
   theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
-  ggtitle("Adjacent Chart")
+  ggtitle("Adjacent (Type 1)")
 
@@
   geom_col(position = "dodge", fill = "white", color = "black") + 
-  geom_text(aes(x = factor(Order %%2), y = 5, group = GroupOrder, label = IDchr), position = position_dodge(0.9)) + 
+  geom_point(aes(x = factor(Order %%2), y = 5, group = GroupOrder, shape = IDchr), size = 3, position = position_dodge(0.9)) +
+  guides(shape = "none") + 
   theme_bw() + 
   theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
-  ggtitle("Separated Chart")
+  ggtitle("Separated (Type 3)")
 
@@
 #| fig-height: 4
-#| out-width: "30%"
+#| out-width: "33%"
 #| fig-cap: "An illustration of the sine illusion\\citep{vanderplasSignsSineIllusion2015}, also known as the line-width illusion.
@@
 
-persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1.75, 1.75), scale=FALSE, box=FALSE, expand=3/pi, d=50) # , ltheta=0, lphi=-15
+par(mar = c(0, 0, 0, 0))
+persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1, 1), scale=FALSE, box=FALSE, expand=2.5/pi, d=50) # , ltheta=0, lphi=-15
 
@@
 However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.
-Digital graphics has developed quickly, along with the hardware necessary to support these software developments. 
+Digital graphics have developed quickly, along with the hardware necessary to support these software developments. 
 As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
@@
 
-Each graph is presented so that there are ten bars where only two of the bars are marked for identification, \svp{as shown in @fig:bar-types.}
-Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.
+\svp{In each graph, there are two sets of five bars each, and two of the bars are marked for identification with a circle and triangle}, \svp{as in \Cref{fig:bar-types}.}
+Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.\svp{What does excessive noise mean??}
 Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).
@@
 Example graphs are shown in
-Figure \@ref(fig:plotTypes).
+\Cref{fig:plotTypes}.
 There are some graphical elements that cannot be easily portrayed via 3D printing. For this reason, all graph types do not have axes, grid lines, or floating titles.
@@
 
-```{r plotTypes, fig.cap = "Two dimensional, two-dimensional digital rendering, and 3D-printed charts used in this study.", out.width = "100%"}
-knitr::include_graphics("_images/plot-types.png", dpi = 300)
+```{r plotTypes, fig.cap = "Two dimensional, 3D digital rendering, and 3D-printed charts used in this study.",out.width = c("30%", "30%", "35%"), fig.show = "hold"}
+knitr::include_graphics("_images/Type1-Rep01.png", dpi = 300)
+knitr::include_graphics("_images/RenderedChart.png", dpi = 300)
+knitr::include_graphics("_images/Kit_of_charts.png", dpi = 300)
 ```
@@
 
-Participants were provided a kit with five 3D-printed charts corresponding to five of the seven unique ratio comparisons.
-In the Shiny application, we asked participants for the kit ID, which was used to look up the corresponding digital renderings.
-This ensured that the experimental design was balanced across chart type and randomized with respect to the type of comparison (adjacent or separated).
-We printed 21 kits containing five 3d-printed charts each, where 2 of the 7 ratios were randomly removed from the total set.
+A diagram of the experimental design is provided in \Cref{fig:studyDesign}.
+The study is set up as a modified randomized incomplete block design, comprised of 3 blocks (presentation mode), 7 bar ratios (5 are selected for each participant).
+At each combination of ratio and presentation type, the comparison type, adjacent (Type 1) or separated (Type 3), is randomly selected.
+This process was used to create 21 kits of 5 3D printed charts and 10 virtual charts.
+There is thus a slight deviation from a completely randomized incomplete block design, in that all possible combinations of ratios, presentation modes, and comparison types are not present in our design.
+However, this approach provided the best experimental design given the constraints of managing physical stimuli; it took nearly a month of continuous 3D printing to create all of the charts for the kits that we assembled.
+A dataset containing the kit composition breakdown is available in the github repository for this project.
+
 <!-- There are 21 kits in total so that each combination of ratios is accounted for. -->
@@
 ```{r studyDesign, fig.cap = "A graphical representation of the study design.
-Only five of the seven ratios were used in each kit where the ratios are randomly selected.
-A total of 21 kits were created to include all combinations of the five ratios."}
-knitr::include_graphics("_images/study-design.png")
+Only five of the seven ratios were used in each kit; at least one of the smallest or largest ratios were included along with 4 other charts; each selected ratio was displayed in all 3 mediums.
+For each medium $\\times$ ratio combination, the comparison type (separated or adjacent) was randomly determined.
+A total of 21 kits of 3D printed charts were created to include all combinations of the five ratios."}
+knitr::include_graphics("_images/design.pdf")
 ```
@@
 Most of the subjects in the position-length experiment participated in the position-angle experiment; in all cases repeat subjects judged the position-angle graphs first." 
-It would seem likely that the authors recruited individuals within their respective departments as well as their wives.
-In the spirit of replicating the study, members of the UNL Statistics department and their spouses, partners, and roommates were asked to participate in our study; this replicates the spirit of the original study without the implicit assumptions that graduate students and professors are (1) largely male, (2) heterosexual, and (3) have unemployed partners.
+It seems likely that the authors recruited individuals within their respective departments as well as their wives.
+In order to replicate this atypical participant recruitment method in the modern era, members of the UNL Statistics department and their spouses, partners, and roommates were asked to participate in our study.
+This replicates the spirit of the recruitment method without the implicit assumptions that graduate students and professors are male, heterosexual, and have unemployed spouses.
 
-A total of 48 participants completed the study; demographics are shown in \@ref(fig:demographics).
+A total of 48 participants completed the study; demographics are shown in \Cref{fig:demographics}.
 
-```{r demographics, fig.cap="Demographic breakdown of participants in the study.
-All subjects were recruited from faculty and students in the statistics department at University of Nebraska-Lincoln.", message=FALSE, warning=FALSE, out.width='85%'}
+```{r demographics, fig.cap="Demographic characteristics of participants in the study.", message=FALSE, warning=FALSE, out.width='100%', fig.width = 9, fig.height = 3.5}
 library(patchwork)
@@
   summarise(count = n()) %>% 
-  ggplot(mapping = aes(x = factor(age, levels = age.levels),
-                       y = count)) + 
+  ggplot(mapping = aes(x = factor(age, levels = rev(age.levels)), y = count)) + 
   geom_bar(stat = 'identity') +
-  geom_text(mapping = aes(y = count+1, label = count),
-             size = 2) +
-  labs(title = 'Age Groups',
-       x = '',
-       y = 'Count') +
+  geom_text(aes(x = factor(age, levels = age.levels), y = count+.5, label = age), hjust = 0) + 
+  coord_flip() + 
+  labs(title = 'Age', y = 'Count') +
   theme_bw() + 
-  theme(plot.title = element_text(hjust=0.5),
-        axis.text.x = element_text(angle = 45, hjust=1, size = 8))
+  scale_y_continuous(expand = expand_scale(mult = 0, add = c(.25, 4))) + 
+  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())
 
@@
   summarise(count = n()) %>% 
-  ggplot(mapping = aes(x = gender,
-                       y = count)) + 
+  ggplot(mapping = aes(x = gender, y = count)) + 
   geom_bar(stat = 'identity') +
-  geom_text(mapping = aes(y = count+1, label = count),
-             size = 2) +
-  labs(title = 'Gender Groups',
-       x = '',
-       y = 'Count') +
+  geom_text(aes(x = gender, y = count + .5, label = gender), hjust = 0) + 
+  coord_flip() + 
+  labs(title = 'Gender', y = 'Count') +
   theme_bw() + 
-  theme(plot.title = element_text(hjust=0.5),
-        axis.text.x = element_text(angle = 45, hjust=1, size = 8))
+  scale_y_continuous(expand = expand_scale(mult = 0, add = c(.25, 4))) + 
+  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())
 
@@
   summarise(count = n()) %>% 
-  ggplot(mapping = aes(x = factor(education, levels = educ.levels),
-                       y = count)) + 
+  ggplot(mapping = aes(x = factor(education, levels = educ.levels), y = count)) + 
   geom_bar(stat = 'identity') +
-  geom_text(mapping = aes(y = count+2, label = count),
-             size = 2) +
-  labs(title = 'Education Levels',
-       x = '',
-       y = 'Count') +
+  coord_flip() + 
+  geom_text(aes(x = education, y = count + .5, label = str_replace(education, "Graduate Degree", "Graduate\nDegree")), hjust = 0) + 
+  scale_y_continuous(expand = expand_scale(mult = 0, add = c(.25, 15))) + 
   theme_bw() + 
-  theme(plot.title = element_text(hjust=0.5),
-        axis.text.x = element_text(angle = 45, hjust=1, size=8))
+  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
+  labs(title = 'Education', y = 'Count')
 
@@
 
+While the results of any sample with recruitment methodology like this are not generalizable to the public, running our initial study on a comparable population to that of the initial study serves a purpose: other replications, such as @heerCrowdsourcingGraphicalPerception2010b, used online samples of paid participants, who likely have less experience using charts and graphics and making numerical estimations than those in a statistics department (and individuals who cohabitate with them).
+We have every intention of running this experiment in other populations, but the convenience of online sampling is not compatible with physical objects such as our 3D printed charts, so we will have to recruit in-person participants.
+Thus, an initial study in a population comparable to @clevelandGraphical1984 is a reasonable first step into testing 3D charts.
+
 ## Data Collection
@@
 Participants provided informed consent through the applet, and then were asked for demographic information (age, gender, education level). 
-Then, participants were shown a "practice" page which allowed them to experiment with the data collection interface and practice estimating the ratio between the bars, as shown in Figure \@ref(fig:practice)
+Then, participants were shown a "practice" page which allowed them to experiment with the data collection interface and practice estimating the ratio between the bars, as shown in \Cref{fig:practice}.
+This practice interface did not provide any feedback as to participant correctness, but was merely intended to familiarize the participants with the questions which would be asked as well as the process of estimating the ratio between the two bars.
 
@@
 
-Directly before the experiment started, participants were asked to provide the kit ID, along with directions indicating that if the instructions indicated that participants should use a 3D chart for a task, the participant should select a chart from the kit, enter the ID code of that chart from the bottom of the object, and complete the requested task.
-Participants were also instructed to make quick judgments for each graph and not to measure or estimate ratios using physical objects.
+After the practice screen, participants were asked to provide the kit ID.
+Participants were also instructed that when indicated, they should choose a 3D chart from the kit and select the ID code of the chart (inscribed on the bottom) from the drop-down menu before completing the estimation task.
+Participants were also instructed to make quick judgements for each graph and not to measure or estimate ratios using physical objects.
 
@@
 
-Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in Figure \@ref(fig:experiment3dRender).
+Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in \Cref{fig:experiment3dRender}.
+It should be noted that we chose to use a slider for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries [@ruudUncertaintyCausesRounding2014a;@maineriSliderBarsMultiDevice2021].
+Another benefit of this modification is that it should reduce participants' cognitive load, allowing them to focus on the ratio between bars rather than the numerical translation process.
+This differs from @clevelandGraphical1984, but maintains the spirit of the study while using methods which were not available for data entry at the time.
 
@@
 
-All responses that incorrectly identified the smaller bar were removed from the study before analysis.
+All responses ($n = `r nrow(res_all) - nrow(res)`$) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check.
 
 ## Midmeans of Log Absolute Errors
-Cleveland and McGill used 
+\citeauthor{clevelandGraphical1984} used 
 $$\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$$
@@
 
-Figure \@ref(fig:midmeans-log-errors) shows the midmeans of the log absolute errors compared to the true ratio of the bars for each graph type and comparison type. 
+\Cref{fig:midmeans-log-errors} shows the midmeans of the log absolute errors compared to the true ratio of the bars for each graph type and comparison type. 
 The results suggest that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types. In addition, there is a tendency
@@
   scale_alpha_manual("Data", values = c(1, .25)) + 
+  scale_fill_discrete(labels = c('Adjacent', 'Separated')) +
   scale_color_discrete(labels = c('Adjacent', 'Separated')) +
@@
        x = 'True Proportional Difference (%)',
-       y = 'Log Error',
+       y = 'Log Error',fill = 'Comparison Type',
        color = 'Comparison Type') +
@@
 It is possible that this difference is due to the fact that we did not require an explicit numerical estimate of the ratio but instead asked participants to indicate the ratio on a slider (which is essentially a number line).
-This should reduce the cognitive load required to transition from spatial comparison to numerical comparison and then to compute the proportion, but may also have impacted the results.
+It is possible that this change explains the lack of a difference between chart modalities we see, but there are other potential explanations as well. 
+One interesting facet of this data is that there is a consistent reduction in log error when the true proportional difference is near 50%. 
+This may be because of implicit anchoring - we can typically bisect a span relatively accurately, which means that it might be expected that when the true proportion is near 50%, we would both notice that and be able to accurately transfer that information to the slider.
+This pattern is more noticable in separated (Type 3) comparisons, which also makes sense - as these comparisons are known to be less accurate, a locally  minimal error near 50% with increased error relative to adjacent (Type 1) comparisons could easily explain the exaggerated trends we see in \Cref{fig:midmeans-log-errors}.
 
@@
 
-$$y_{ijklm}=\mu+S_i+R_j+G(R)_{(k)j}+T_l+\epsilon_{ijklm}$$
+$$y_{ijklm}=\mu+S_i + R_j + G_k +T_l + \epsilon_{ijklm}$$
 
@@
 
-- $S_i\sim N(0,\sigma^2_S)$ is the effect of the $i^{th}$ subject
+- $S_i\sim N(0,\sigma^2_S)$ is the random effect of the $i^{th}$ subject
 
@@
 
-- $G(R)_{(k)j}$ is the effect of the $k^{th}$ graph type nested in the $j^{th}$ ratio
+- $G_k$ is the effect of the $k^{th}$ display type
 
@@
 
-```{r, include=FALSE}
-mod = lmer(response ~ (1|subject) + ratio + type + ratio:plot,
+```{r, include=T}
+library(papaja)
+library(lme4)
+
+mod <- lmer(response ~ (1|subject) + ratio + plot + type,
             data = res)
-anova(mod)
+modsum <- summary(mod)
+modaov <- anova(mod)
+modaovsum <- summary(modaov)
+
+f_df <- as.numeric(length(modsum$residuals)-modsum$ngrps - 1)
+
+apa_lm <- apa_print(mod)
+apa_table(apa_lm$table, caption = "Analysis of Fixed Effects")
 ```
 
-No differences were detected for the true ratio of bars (p-value = .7881), whether the bars were adjacent or separated (p-value = .3375), or for the plot nested within the true ratio (p-value = .6868).
-While Heer \& Bostock provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data at this time.
+No differences were detected for the true ratio of bars ($t = `r modsum$coefficients[2,3]`$), whether the bars were adjacent or separated ($t = `r modsum$coefficients[5,3]`$), or for the plot nested within the true ratio ($F(`r modaov$npar[2]`, `r f_df`) = `r modaov[2,4]`$).
+While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes.
+
+## Interactivity with 3D Rendered Charts
+
+When 3D rendered plots were shown, we recorded the number of user interactions (clicks, rotations) with the WebGL plot.
+\Cref{fig:clicks} shows the distribution of number of clicks over all 3D rendered trials in the experiment, as well as the number of clicks relative to the accuracy measure $\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$.
+It does not appear that participants who interacted with the WebGL rendering were more accurate than their peers who did not.
+
+```{r clicks, fig.cap = "Participant clicks on the WebGL interface (left) in order to interact with the plot.
+Most participants did not interact with the WebGL interface, perhaps because they did not realize that interacting was an option, but some participants utilized the interactivity heavily.
+(Right) Participant interactivity was not associated with increased accuracy.", fig.width = 4, fig.height = 4, out.width = ".45\\textwidth", fig.show = "hold"}
+res %>%
+  filter(plot == "3dDigital") %>%
+  ggplot(aes(x = plot3dClicks)) + geom_bar() + 
+  xlab("# Clicks on WebGL Plot") + ylab("Count") + ggtitle("Distribution of WebGL Interaction") + 
+  theme_bw()
+
+res %>%
+  filter(plot == "3dDigital") %>%
+  ggplot(aes(x = plot3dClicks, y = response)) + geom_point() + 
+  xlab("# Clicks on WebGL Plot") + ylab("Count") + ggtitle("Accuracy and Participant Interactivity") + 
+  geom_smooth(method = "lm", se = T) + 
+  theme_bw()
+```
 
@@
 The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger.
-However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @clevelandGraphical1984 do not include significance testing beyond the graphical display of confidence intervals.
+However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @clevelandGraphical1984 do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.
 
@@
 The slider method for input of ratio estimates should be easier for participants, as it does not require explicit transformation to the numerical domain. 
-What is clear is that it would be beneficial to assess the impact of measurement method on participant errors directly, so that the results of these different studies might be explained and interpreted with regard both to the stimuli used and the measurement method employed in the experiment.
+What is clear is that it would be beneficial to assess the impact of the measurement method on participant errors directly, so that the results of these different studies might be explained and interpreted with regard both to the stimuli used and the measurement method employed in the experiment.
 Future iterations of this experiment will likely address this estimation difference; such modifications in experimental design are relatively straightforward in Shiny and will provide useful insight into the design of future experiments evaluating the perception of statistical graphics.

### 2024-01-22 jds.tex
Updates and Reviewer 1 corrections

--- 2024-01-22 jds.rmd
@@
----
-title: "Evaluating Perceptual Judgements on 3D Printed Bar Charts"
-author:
-- first: Tyler
-  first-init: T.
-  last: Wiederich
-  marker: 1
-  footnote-id: 1
-  email: twiederich2@huskers.unl.edu
-- first: Susan
-  first-init: S.
-  last: VanderPlas
-  marker: 1
-  email: susan.vanderplas@unl.edu
-short-title: Perceptual Judgements on 3D Printed Bar Charts
-output:
-  jds.rmd::pdf_article: 
-    extra_dependencies: "subfig"
-affiliation:
-- institution: "University of Nebraska-Lincoln"
-  prefix: Department of Statistics
-  country: United States of America
-  marker: 1
-classoption:
-- letterpaper
-- inpress
-- dvipsnames
-preamble: |
+\documentclass[letterpaper,inpress,dvipsnames]{jdsart}
+
+\setcounter{page}{1}
+\pubmonth{July}
+\pubyear{2022}
+\volume{xx}
+\issue{xx}
+\doi{0000}
+
+\usepackage[utf8]{inputenc}
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{subfig}
+
 \usepackage{amsfonts,amsmath,amssymb,amsthm} 
@@
 \newcommand{\tw}[1]{{\textcolor{Red}{#1}}}
-abstract: "Graphical design principles typically recommend minimizing the dimensionality of a visualization - for instance, using only 2 dimensions for bar charts rather than providing a 3D rendering, because this extra complexity may result in a decrease in accuracy.
+
+\begin{document}
+\begin{frontmatter}
+
+\title{Evaluating Perceptual Judgements on 3D Printed Bar Charts}
+\runtitle{Perceptual Judgements on 3D Printed Bar Charts}
+
+\author[1]{
+  \inits{T.}
+  \fnms{Tyler}
+  \snm{Wiederich}  \thanksref{1}  \ead{twiederich2@huskers.unl.edu}}
+\author[1]{
+  \inits{S.}
+  \fnms{Susan}
+  \snm{VanderPlas}  \ead{susan.vanderplas@unl.edu}}
+
+\thankstext[type=corresp,id=1]{Tyler Wiederich}
+\address[1]{Department of Statistics, 
+  \institution{University of Nebraska-Lincoln}, \cny{United States of America}}
+
+\begin{abstract}
+Graphical design principles typically recommend minimizing the dimensionality of a visualization - for instance, using only 2 dimensions for bar charts rather than providing a 3D rendering, because this extra complexity may result in a decrease in accuracy.
 This advice has been oft repeated, but the underlying experimental evidence is focused on fixed 2D projections of 3D charts.
 In this paper, we describe an experiment which attempts to establish whether the decrease in accuracy extends to 3D virtual renderings and 3D printed charts.
-We replicate the grouped bar chart comparisons in the  1984 Cleveland & McGill study, assessing the accuracy of numerical estimates using different types of 3D and 2D renderings."
-keywords:
-- graphics
-- 3D bar charts
-- 3D printing
-bibliography: JDS2023.bib
-footnote:
-- content: Tyler Wiederich
-  type: corresp
-  id: 1
----
+We replicate the grouped bar chart comparisons in the 1984 Cleveland \& McGill study, assessing the accuracy of numerical estimates using different types of 3D and 2D renderings.
+\end{abstract}
 
-```{r, include = F}
-knitr::opts_chunk$set(echo = F, dpi = 300, message =F, warning = F, error = F, dev = "ragg_png")
+\begin{keywords}
+\kwd{graphics}\kwd{3D bar charts}\kwd{3D printing}.
+\end{keywords}
 
-library(tidyverse)
-library(RSQLite)
-library(lme4)
+\end{frontmatter}
 
-con = dbConnect(SQLite(), '../2022-Tyler-3D_Graphical_Perception/experiment_interface/department.db')
-dbListTables(con)
-results = dbReadTable(con, 'results')
-users = dbReadTable(con, 'user')
-userMatrix = dbReadTable(con, 'userMatrix')
-dbDisconnect(con)
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-valid.users <- users %>% 
-  mutate(subject = paste0(nickname, participantUnique)) %>% 
-  filter(age != 'Under 19')
-
-#Fill in correct values for incorrect 3d graph kits
-# 1 result, need to fill in fileID, graphtype, and plot
-results = results %>% 
-  mutate(fileID = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', 1, fileID),
-         graphtype = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', 'Type1', graphtype),
-         plot = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', '3dPrint', plot))
-
-load('../2022-Tyler-3D_Graphical_Perception/experiment_interface/data/set85data.Rdata')
-load('../2022-Tyler-3D_Graphical_Perception/experiment_interface/data/kits.Rdata')
-
-trueRatios = datasets %>% 
-  mutate(ratio.df = map(data, function(x)(x[!is.na(x[,'IDchr']),4])),
-         trueRatio = map(ratio.df, function(x)(x[1,'Height'] / x[2,'Height']))) %>% 
-  unnest(trueRatio) %>% 
-  filter(fileID != 15) %>% 
-  #mutate(Height = Height * 100) %>% 
-  select(fileID, Height)
-
-res_all = results %>% 
-  left_join(trueRatios, by = 'fileID') %>% 
-  mutate(response = log2(abs(byHowMuch - Height*100) + 1/8),
-         subject = paste0(nickname, participantUnique),
-         ratioLabel = round(100*Height, 1)) %>% 
-  filter(subject %in% valid.users$subject) %>% 
-  arrange(appStartTime)
-
-res <- res_all %>% 
-  filter(whichIsSmaller == 'Triangle (▲)',
-         as.Date.POSIXct(appStartTime) <= '2023-05-22') #Only verified results from before the start of SDSS
-
-```
-
-# Introduction
 Good communication requires both that the information be transmitted correctly and that the intended recipient be able to decode and understand the transmitted information accurately.
@@
 In many cases, this means we must understand how accurately people can read quantitative information off of charts.
-While accuracy is not the only quantity of interest in graphical investigations [@hullmanPursuitErrorSurvey2019], it is an important factor in assessing the utility of many different data graphics.
+While accuracy is not the only quantity of interest in graphical investigations \citep{hullmanPursuitErrorSurvey2019}, it is an important factor in assessing the utility of many different data graphics.
 
-The accuracy of graphical forms has been studied for almost a century [@vonhuhnFurtherStudiesGraphic1927;@eellsRelativeMeritsCircles1926;@croxtonGraphicComparisonsBars1932;@croxtonBarChartsCircle1927], as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions.
+The accuracy of graphical forms has been studied for almost a century \citep{vonhuhnFurtherStudiesGraphic1927, eellsRelativeMeritsCircles1926, croxtonGraphicComparisonsBars1932, croxtonBarChartsCircle1927}, as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions.
 This is particularly true in areas like graphics which are affected by the immense technological innovation in hardware and software which has taken place since the early 1990s.
 
-## Elementary Graphical Tasks
-@clevelandGraphical1984 established the comparative accuracy of different "elementary perceptual tasks" (EPTs).
+\hypertarget{elementary-graphical-tasks}{%
+\subsection{Elementary Graphical Tasks}\label{elementary-graphical-tasks}}
+
+\citet{clevelandGraphical1984} established the comparative accuracy of different ``elementary perceptual tasks'' (EPTs).
 Elementary Perceptual Tasks, according to these experiments, include assessing graphical elements such as position along a common scale, length, angle, and volume, and estimating the corresponding numerical value of these representations.
 The study relied entirely on estimation accuracy, which may not always be relevant when extracting information from graphs. For example, estimation is less relevant when ordering values by size.
-As a result of the Cleveland and McGill [-@clevelandGraphical1984] study, it is possible to assemble an ordering of perceptual accuracy for the elements of length, position, and angle. 
-@heerCrowdsourcingGraphicalPerception2010b replicated some parts of @clevelandGraphical1984 in an online setting using Mechanical Turk, largely validating the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.
+As a result of the Cleveland and McGill \citeyearpar{clevelandGraphical1984} study, it is possible to assemble an ordering of perceptual accuracy for the elements of length, position, and angle.
+\citet{heerCrowdsourcingGraphicalPerception2010b} replicated some parts of \citet{clevelandGraphical1984} in an online setting using Mechanical Turk, largely validating the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.
 
-The first experiment in @clevelandGraphical1984 (the position-length experiment), used five types of bar charts: two types of grouped bar charts and three types of stacked bar charts.
+The first experiment in \citet{clevelandGraphical1984} (the position-length experiment), used five types of bar charts: two types of grouped bar charts and three types of stacked bar charts.
 Each chart had two bars marked for comparison; participants were asked to determine which bar was smaller and give the perceived ratio of the smaller bar to the larger bar.
@@
 
-```{r bar-types}
-#| fig-width: 4
-#| fig-height: 4
-#| fig-cap: "\\citet{clevelandGraphical1984} used two different types of grouped bar charts: comparisons between adjacent bars, and comparisons between separated bars.
-It is widely acknowledged that comparisons between separated bars (Type 3 comparisons, in Cleveland \\& McGill's terminology) are more difficult and error-prone."
-#| out-width: ".49\\linewidth"
-#| fig-show: "hold"
-#| layout-ncol: 2
-library(ggplot2)
-library(ragg)
-datasets$data[[1]] %>%
-  ggplot(aes(x = Group, y = Height, group = GroupOrder)) + 
-  geom_col(position = "dodge", fill = "white", color = "black") + 
-  geom_point(aes(x = Group, y = 5, group = GroupOrder, shape = IDchr), size = 3, position = position_dodge(0.9)) + 
-  guides(shape = "none") + 
-  theme_bw() + 
-  theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
-  ggtitle("Adjacent (Type 1)")
+\begin{figure}
+\includegraphics[width=.49\linewidth]{jds_files/figure-latex/bar-types-1} \includegraphics[width=.49\linewidth]{jds_files/figure-latex/bar-types-2} \caption{\citet{clevelandGraphical1984} used two different types of grouped bar charts: comparisons between adjacent bars, and comparisons between separated bars.
+It is widely acknowledged that comparisons between separated bars (Type 3 comparisons, in Cleveland \& McGill's terminology) are more difficult and error-prone.}\label{fig:bar-types}
+\end{figure}
 
-datasets$data[[1]] %>%
-  ggplot(aes(x = factor(Order %%2), y = Height, group = GroupOrder)) + 
-  geom_col(position = "dodge", fill = "white", color = "black") + 
-  geom_point(aes(x = factor(Order %%2), y = 5, group = GroupOrder, shape = IDchr), size = 3, position = position_dodge(0.9)) +
-  guides(shape = "none") + 
-  theme_bw() + 
-  theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
-  ggtitle("Separated (Type 3)")
+\hypertarget{d-graphical-perception}{%
+\subsection{3D Graphical Perception}\label{d-graphical-perception}}
 
-```
-
-## 3D Graphical Perception
 Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
-This occasionally causes problems: the line-width illusion[@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013], for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in @fig:sine-illusion).
+This occasionally causes problems: the line-width illusion\citep{vanderplasSignsSineIllusion2015, daySineIllusion1991, hofmannCommonAnglePlots2013}, for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in \citet{fig:sine-illusion}).
 While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception.
@@
 
-```{r sine-illusion}
-#| message: false
-#| error: false
-#| warning: false
-#| cache: false
-#| fig-width: 4
-#| fig-height: 4
-#| out-width: "33%"
-#| fig-cap: "An illustration of the sine illusion\\citep{vanderplasSignsSineIllusion2015}, also known as the line-width illusion.
+\begin{figure}
+\subfloat[Sine Illusion\label{fig:sine-illusion-1}]{\includegraphics[width=0.33\linewidth]{jds_files/figure-latex/sine-illusion-1} }\subfloat[3D surface with similar features\label{fig:sine-illusion-2}]{\includegraphics[width=0.33\linewidth]{jds_files/figure-latex/sine-illusion-2} }\subfloat[Perceived vs.
+actual length\label{fig:sine-illusion-3}]{\includegraphics[width=0.33\linewidth]{jds_files/figure-latex/sine-illusion-3} }\caption{An illustration of the sine illusion\citep{vanderplasSignsSineIllusion2015}, also known as the line-width illusion.
 All vertical lines are the same length, but the lines in the middle of the curve appear to be much shorter.
 The illusion results when implicit perceptual corrections useful for perceiving the size of objects with depth are applied to 2D stimuli with no actual depth.
-As 3D heuristics occasionally cause inaccurate communication when applied to 2D objects, it is reasonable to think that there might be some situations where charts making use of a realistic third dimension might be perceived more accurately than their 2D equivalents."
-#| fig-subcap:
-#|    - "Sine Illusion"
-#|    - "3D surface with similar features"
-#|    - "Perceived vs. actual length"
-#' Function to data frame
-#'  
-#' @param n number of values 
-#' @param ell extent of vertical range
-#' @param x range of horizontal values
-#' @param f function
-#' @param fp first derivative of function f
-#' @param f2p second derivative of function f
-#' @return data frame with n function values and derivatives along the x axis for a range given by x
-#' @example
-#' f <- function(x) 2*sin(x)
-#' fp <- function(x) 2*cos(x)
-#' f2p <- function(x) -2*sin(x)
-#' dframe <- function.frame(n=50, ell=1, x=c(0,2*pi), f, fp, f2p)
-#' require(ggplot2)
-#' qplot(x, y=ystart, yend=yend, xend=x, geom="segment", data=dframe) +
-#'    geom_point(aes(x,y))
-function.frame <- function(n=200, ell=1, x, f, fprime, f2prime) {
-  x <- seq(x[1], x[2], length=n)
-  if (length(ell) != length(x)) ell <- rep(ell, length(x))
-  y <- f(x)
-  ystart <- y - ell/2
-  yend <- y + ell/2
-  # now correct for line illusion in vertical direction
-  dy <- diff(range(y))
-  dyl <- diff(range(c(ystart, yend)))
-  # fprime and f2prime are sensitive to the aspect ratio of a plot
-  # we represent it in framework of dy and dy+len
-  # needs to be fixed by factor a:  
-  a <- dy/dyl
-
-  fp <- a*fprime(x)
-  f2p <- a*f2prime(x)
-  data.frame(x,y,ystart, yend, fp, f2p, a)
-}
-
-#' This function has to be split in parts and renamed
-#' 
-#' Right now this function is a convoluted mess.  We need to have separate
-#' functions that 
-#' (1) create a data frame for a given function,
-#' its first and second derivatives and range in x
-#' (this is what the function function.frame is supposed to do), 
-#' (2) a trig transform function
-#' (3) a quadratic approach function - better even, make (2) and (3) one function and use a parameter to decide on the method to use.
-#' @example
-#' f <- function(x) 2*sin(x)
-#' fp <- function(x) 2*cos(x)
-#' f2p <- function(x) -2*sin(x)
-#' dframe <- createSine(n=50, ell=1, x=c(0,2*pi), f, fp, f2p)
-#' require(ggplot2)
-#' qplot(x=x, xend=x, y=y+ellx4.u, yend=y-ellx4.l, geom="segment", data=dframe, linetype=I(1)) + 
-#'    theme_bw() + coord_fixed(ratio=1) + 
-#'    xlab("x") + ylab("y") + 
-#'    scale_x_continuous(breaks=seq(0, 2*pi, by=pi/2), minor_breaks=minor.axis.correction,
-#'      labels=c("0", expression(paste(pi,"/2")), expression(pi), expression(paste("3",pi, "/2")), 
-#'      expression(paste("2",pi))))
-
-createSine <- function(n=200, len=1, f=f, fprime=fprime, f2prime=f2prime, a=0, b=2*pi) {
-  #  if(getquadapprox & !is.function(f2prime)) f2prime <- function(x) -1*f(x) # for backwards compatibility
-  x <- seq(a, b, length=n+2)[(2:(n+1))]
-  ell <- rep(len, length=length(x))
-  fx <- f(x)
-  ystart <- fx - .5*ell
-  yend <- fx + .5*ell
-
-  # now correct for line illusion in vertical direction
-  dy <- diff(range(fx))
-  dx <- diff(range(x))
-  # fprime works in framework of dx and dy, but we represent it in framework of dx and dy+len
-  # needs to be fixed by factor a:  
-#  a <- dy/(dy + len) 
-  dyl <- diff(range(c(ystart, yend)))
-  a <- dy/(dyl) 
-  # ellx is based on the "trig" correction
-  ellx <- ell / cos(atan(abs(a*fprime(x))))
-  # ellx2 is based on linear approximation of f  
-  ellx2 <- ell * sqrt(1 + a^2*fprime(x)^2)
-
-  # make this a data frame - ggplot2 doesn't do well with floating vectors
-  dframe <- data.frame(x=x, xstart=x, xend=x, y=fx, ystart=ystart, yend=yend, ell=ell, ellx = ellx, ellx2=ellx2)
-
-  # third adjustment is based on quadratic approximation of f.
-  # this needs two parts: correction above and below f(x)  
-  #   if(getquadapprox & is.function(f2prime)){
-  #     secseg <- do.call("rbind", lapply(dframe$x, function(i) getSecantSegment(i, dframe, f, fprime, f2prime)))
-  #     dframe$ellx3.u <- secseg$sec.ell1
-  #     dframe$ellx3.l <- secseg$sec.ell2
-  #   }
-
-  fp <- a*fprime(x)
-  f2p <- a*f2prime(x)
-  v <- 1 + fp^2
-  lambdapinv <- 0.5*(sqrt(v^2-f2p*fp^2*ell) + v)    
-#  lambdaminv <- -0.5*(sqrt(v^2+f2p*fp^2*ell) + v)
-  lambdaminv <- 0.5*(sqrt(v^2+f2p*fp^2*ell) + v)
-
-  dframe$ellx4.l <- 0.5*abs(lambdapinv)/sqrt(v)
-  dframe$ellx4.u <- 0.5*abs(lambdaminv)/sqrt(v)
-#  dframe$lambdam <- 1/lambdaminv
-#  dframe$lambdap <- 1/lambdapinv
-  fp <- fprime(x)
-  f2p <- f2prime(x)
-  v <- 1 + fp^2
-#  lambda1 <- (-v + sqrt(v^2 - fp^2*f2p*ell))/(fp^2*f2p) # these two work
-#  lambda2 <- (-v + sqrt(v^2 + fp^2*f2p*ell))/(fp^2*f2p) # these two work
-
-  lambdaa <- -2*(sqrt(v^2 - f2p*fp^2*ell) + v)^-1 # identical to top
-  lambdab <- 2*(sqrt(v^2 + f2p*fp^2*ell) + v)^-1
-
-  dframe$lambdam <- lambdaa
-  dframe$lambdap <- lambdab
-
-  # qplot(a*(sqrt(1 + a^2*fprime(x)^2)*abs(lambdab))^-1, ellx4.u, data=dframe)
-  # qplot(a*(sqrt(1 + a^2*fprime(x)^2)*abs(lambdaa))^-1, ellx4.l, data=dframe)
-  # ggplot(aes(x, y+ell*(a*sqrt(v)*abs(lambdaa))^-1, xend=x, yend=y-ell*(a*sqrt(v)*abs(lambdab))^-1), data=dframe) + geom_segment()
-
-  dframe
-}
-
-#' what is the function doing? - in one sentence
-#' 
-#' function description in a paragraph
-#' @param x0 is a vector of locations for which the secant segments are supposed to be calculated
-#' @param df df is a data frame
-#' @param f function
-#' @param fp first derivative of function f
-#' @param f2p second derivative of function f
-#' @return data frame consisting of .... what is the output of the function?
-#' @example
-#' # need an example here of how to use the function
-getSecantSegment <- function(x0, df, f, fprime, f2prime){
-# find the closest values of the grid df corresponding to x0
-  # doesn't work right now
-  ell.x0 <- sapply(x0, function(x) df$ell[which.min(x-df$x)])
-
-  dy <- diff(range(with(df, c(y+ell.x0, y-ell.x0))))
-  dx <- diff(range(df$x))
-  a <- dx/dy
-
-  #---- Actual Roots
-  #   secSlope   <- -a/fprime(x0)
-  #   temp       <- seq(min(df$x)-pi, max(df$x)+pi, .0001)
-  #   leftend    <- temp[which.min(abs(f(temp) + ell.x1 - secSlope*(temp-x1)))]
-  #   rightend   <- temp[which.min(abs(f(temp) - ell.x1 - secSlope*(temp-x1)))]
-  #----
-
-  #---- Approximation
-  lambda1 <- (-(fprime(x0)^2 + 1) + sqrt((fprime(x0)^2 + 1)^2 - 2*fprime(x0)^2*f2prime(x0)*ell.x0))/(fprime(x0)^2*f2prime(x0))
-  lambda2 <- (-(fprime(x0)^2 + 1) + sqrt((fprime(x0)^2 + 1)^2 + 2*fprime(x0)^2*f2prime(x0)*ell.x0))/(fprime(x0)^2*f2prime(x0))
-
-  x1 <- lambda1*fprime(x0)+x0
-  x2 <- lambda2*fprime(x0)+x0
-  y1 <- f(x0)-lambda1
-  y2 <- f(x0)-lambda2
-  #----
-
-  #---- Approximation V2 ----
-  #   dx <- fprime(x0)
-  #   dx1 <- 1+dx^2   
-  #   ddx <- f2prime(x0)
-  #   
-  #   corr1 <- 1/a*(((dx1)-sqrt(dx1^2-2*ddx*(dx1-1)^2*ell.x0))/(ddx*dx))
-  #   corr2 <- 1/a*(((dx1)-sqrt(dx1^2+2*ddx*(dx1-1)^2*ell.x0))/(ddx*dx))
-  #   if(abs(dx)<.01) {
-  #     dx1 <- 2
-  #     corr1 <- corr2 <- 0
-  #   }
-  #   if(abs(dx)<.01){
-  #     ddx <- 1
-  #   }
-  #   
-  #   x1 = corr1 + x0
-  #   x2 = corr2 + x0
-  #   y1 = f(x1)+ell.x0
-  #   y2 = f(x2)-ell.x0
-  #----
-
-  df2 <- data.frame(x=x0, y=f(x0), deriv=fprime(x0),
-                    sec.xstart=x1, sec.xend = x2, 
-                    sec.ystart=y1, sec.yend = y2,
-                    ell.orig = 2*ell.x0)
-
-  df2$sec.ell1 <- a*.5*ell.x0/with(df2, sqrt((sec.yend-y)^2+(sec.xend-x)^2))
-  df2$sec.ell2 <- a*.5*ell.x0/with(df2, sqrt((y-sec.ystart)^2+(x-sec.xstart)^2))
-  #   df2$sec.ell1 <- with(df2, sqrt((sec.yend-y)^2+(sec.xend-x)^2))
-  #   df2$sec.ell2 <- with(df2, sqrt((y-sec.ystart)^2+(x-sec.xstart)^2))
-  df2$type <- "Perceived Width"
-  df2$a <- a
-  return(df2)
-}
-
-correctx <- function(z, fprime, a=0, b=2*pi, w=1) {
-  # w = 1/(shrink+1)
-  const <- integrate(function(x) abs(fprime(x)), a, b)$value
-  trans <- sapply(z, function(i) integrate(function(x) abs(fprime(x)), a, i)$value*(b-a)/const + a)
-  # alternatively to the rowMeans, you could report back  
-  # trans*(1-w) + z*w
-  trans*w + z*(1-w)
-}
-
-f <- function(x) 2*sin(x)
-fprime <- function(x) 2*cos(x)
-f2prime <- function(x) -2*sin(x)
-
-x <- seq(0, 2*pi, length=42)[2:41]
-data <- do.call("rbind", lapply(seq(-.5, .5, 1), function(i) data.frame(x=x, y=3*sin(x), z=i)))
-
-data.persp <- reshape2::acast(data, x~z, value.var="y")
-x <- sort(unique(data$x))
-y <- sort(unique(data$y))
-z <- sort(unique(data$z))
-dframe <- createSine(n = 50, len = 1, f=f, fprime=fprime, f2prime=f2prime)
-dframe %>%
-  ggplot(aes(x = xstart, xend = xend, y = ystart, yend = yend)) + geom_segment() + coord_fixed() + 
-  theme_void()
-
-par(mar = c(0, 0, 0, 0))
-persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1, 1), scale=FALSE, box=FALSE, expand=2.5/pi, d=50) # , ltheta=0, lphi=-15
-
-# persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1.75, 1.75), scale=FALSE, box=FALSE, expand=3/pi, d=3) # , ltheta=0, lphi=-15
-
-dframe <- createSine(n = 150, len = 1, f=f, fprime=fprime, f2prime=f2prime)
-dframe$ystartcts <- dframe$ystart
-dframe$yendcts <- dframe$yend
-dframe[1:150,c(2, 3, 5, 6)] <- NA
-dframe[(1:15)*10-5, c(2, 3)] <- dframe[(1:15)*10-5, 1] 
-dframe[(1:15)*10-5, 5] <- dframe[(1:15)*10-5, 4] - .5
-dframe[(1:15)*10-5, 6] <- dframe[(1:15)*10-5, 4] + .5
-dframe$type <- "Vertical Width"
-
-idx <- which(!is.na(dframe$xstart))
-dframe$ell <- dframe$ell/2
-dframe.1 <- getSecantSegment(dframe$xstart[idx], dframe, f, fprime, f2prime)
-dframe.1$x <- dframe$x[idx]
-dframe.1$y <- dframe$y[idx]
-dframe.1$ystartcts <- dframe$ystartcts[idx]
-dframe.1$yendcts <- dframe$yendcts[idx]
-names(dframe.1) <- c("x", "y", "deriv", "xstart", "xend", "ystart", "yend", "ell", "ell.quad1", "ell.quad2", "type", "a", "ystartcts", "yendcts")
-dframe.1$vangle <- with(dframe.1, atan(deriv))
-dframe <- bind_rows(dframe, dframe.1)
-dframe$type <- factor(dframe$type)
-
-ggplot(aes(x=x, y=y), data=dframe, colour=I("grey50")) + 
-  geom_line(color = "grey") + theme_void() + 
-  geom_line(aes(y=ystartcts), colour="grey50", alpha = .25) +
-  geom_line(aes(y=yendcts), colour="grey50", alpha = .25) +
-  geom_segment(data=subset(dframe, !is.na(type)), 
-               aes(x=xstart, xend = xend, y=ystart, yend=yend, colour=type, linetype=type), linewidth=0.8)  + 
-  xlab("") + ylab("")  +
-  coord_equal(ratio=1) + scale_colour_manual("", values=c("blue", "grey30")) + 
-  # geom_text(aes(label=paste("theta", "%~~%", round(abs(vangle)/pi*180), "^o", sep=""), 
-                # x=pmax(xstart, xend)/2+x/2+.35 , y=y-sign(vangle)*.6+.02), colour="blue",
-            # data=dframe.1, parse=TRUE, hjust=.9, vjust=.5, size=3) + 
-  scale_x_continuous(breaks=seq(0, 2*pi, by=pi/2), 
-                     labels=c("0", expression(paste(pi,"/2")), expression(pi), 
-                              expression(paste("3",pi, "/2")), expression(paste("2",pi)))) +
-  scale_linetype_manual("", values=c("11", "solid")) + 
-  theme(legend.key.width = unit(3, "line"), plot.margin = unit(c(0,0,0,0), "cm"), legend.position=c(.25, .2))
-
-```
+As 3D heuristics occasionally cause inaccurate communication when applied to 2D objects, it is reasonable to think that there might be some situations where charts making use of a realistic third dimension might be perceived more accurately than their 2D equivalents.}\label{fig:sine-illusion}
+\end{figure}
 
 The use of 3D graphics have been explored in multiple studies.
-@fisher_data_1997 explored user preference for 2D or 3D charts and found that subjects tended to prefer simpler 2D graphs when tasked with extracting information. 
-@barfield_effects_1989 compared 2D and 3D graphs presented on paper and on computers, showing that the accuracy of subject answers depended on their skill level: novice subjects were more accurate with 2D paper graphs, while experienced managers were more accurate with 3D computer graphs.
+\citet{fisher_data_1997} explored user preference for 2D or 3D charts and found that subjects tended to prefer simpler 2D graphs when tasked with extracting information.
+\citet{barfield_effects_1989} compared 2D and 3D graphs presented on paper and on computers, showing that the accuracy of subject answers depended on their skill level: novice subjects were more accurate with 2D paper graphs, while experienced managers were more accurate with 3D computer graphs.
 For both experience levels, participants were more confident in their answers when using 2D graphs.
 There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may better encode information - for instance, when X and Y are used to represent spatial dimensions, it may be preferable to use a 3D chart to convey numerical information instead of using color, which is perceived much less accurately.
-@brath_3d_2014 highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.
+\citet{brath_3d_2014} highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.
 
 There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data.
-However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.
+However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study \citeyearpar{heerCrowdsourcingGraphicalPerception2010b} in 2010.\\
 Digital graphics have developed quickly, along with the hardware necessary to support these software developments.
 As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
-As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].
+As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision \citep{fleetHowTactileGraphics2023}.
 
-In this paper, we discuss a study designed to examine Cleveland & McGill's experiments on grouped bar charts using modern graphics in two and three dimensions.
+In this paper, we discuss a study designed to examine Cleveland \& McGill's experiments on grouped bar charts using modern graphics in two and three dimensions.
 This study lays the groundwork for additional empirical studies on the use of 3D graphics, rendered and 3D printed, for visualizing complex data.
 In order to explore perception of fully 3D graphics, it is prudent to start with the simplest possible 3D graphic: one in which the third dimension is not necessary, so that we can easily compare to two-dimensional representations without loss of information.
-<!-- This is included because of a graphics paper section comment that really annoyed me -->
-In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from @clevelandGraphics1984 in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs.
+In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from \citet{clevelandGraphics1984} in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs.
 The next section presents the results, and we conclude the paper by discussing this experiment in the context of existing work on the perception of 2D and 3D graphical elements.
 
-# Methods
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-Our study is designed to replicate and expand upon the position-length experiment from Cleveland & McGill as closely as possible, with the hope of being able to integrate this study with previously reported results in a consistent manner.
+Our study is designed to replicate and expand upon the position-length experiment from Cleveland \& McGill as closely as possible, with the hope of being able to integrate this study with previously reported results in a consistent manner.
 In this section, we discuss the replication process and the design of our version of this experiment.
 
-## Replicating Cleveland and McGill
+\hypertarget{replicating-cleveland-and-mcgill}{%
+\subsection{Replicating Cleveland and McGill}\label{replicating-cleveland-and-mcgill}}
 
@@
 The exact numeric comparisons were not disclosed, but the comparison values used in our study were subjected to the constraints of having the same ratio values and that no value was used more than twice.
-\svp{Can we show the comparison values in a table?}
+{\textcolor{ForestGreen}{Can we show the comparison values in a table?}}
 
-\svp{In each graph, there are two sets of five bars each, and two of the bars are marked for identification with a circle and triangle}, \svp{as in \Cref{fig:bar-types}.}
-Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.\svp{What does excessive noise mean??}
-Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).
+{\textcolor{ForestGreen}{In each graph, there are two sets of five bars each, and two of the bars are marked for identification with a circle and triangle}}, {\textcolor{ForestGreen}{as in \Cref{fig:bar-types}.}}
+Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.{\textcolor{ForestGreen}{What does excessive noise mean??}}
+Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at \url{https://github.com/TWiedRW/2023-JDS-3dcharts}.
 
-## Stimuli Construction
+\hypertarget{stimuli-construction}{%
+\subsection{Stimuli Construction}\label{stimuli-construction}}
 
-The graphs share a common layout across all formats, where two groupings of five bars are identified by "A" and "B", respectively, and circles and triangles are used to identify the bars participants should compare.
+The graphs share a common layout across all formats, where two groupings of five bars are identified by ``A'' and ``B'', respectively, and circles and triangles are used to identify the bars participants should compare.
 Example graphs are shown in
@@
 There are some graphical elements that cannot be easily portrayed via 3D printing. For this reason, all graph types do not have axes, grid lines, or floating titles.
-<!-- The aspect ratio of the plots presented in the original paper is approximately 0.825:1. The aspect ratio of the plots presented in our study is closer to 0.864:1, but -->
-<!-- The aspect ratio of the 2d plots (including only the bars is 2:1), while the aspect ratio of the 3d printed charts is 11:9.5.
-This is something we probably need to fix....
-Until then, let's not go there -->
 
-```{r plotTypes, fig.cap = "Two dimensional, 3D digital rendering, and 3D-printed charts used in this study.",out.width = c("30%", "30%", "35%"), fig.show = "hold"}
-knitr::include_graphics("_images/Type1-Rep01.png", dpi = 300)
-knitr::include_graphics("_images/RenderedChart.png", dpi = 300)
-knitr::include_graphics("_images/Kit_of_charts.png", dpi = 300)
-```
+\begin{figure}
+\includegraphics[width=0.3\linewidth]{_images/Type1-Rep01} \includegraphics[width=0.3\linewidth]{_images/RenderedChart} \includegraphics[width=0.35\linewidth]{_images/Kit_of_charts} \caption{Two dimensional, 3D digital rendering, and 3D-printed charts used in this study.}\label{fig:plotTypes}
+\end{figure}
 
-The ggplot2 [@ggplot2] package was utilized to create the 2D bar charts. 
+The ggplot2 \citep{ggplot2} package was utilized to create the 2D bar charts.
 The scale axis was removed, leaving only the bars and a bar grouping identifier.
@@
 
-The 3D renderings and 3D printed charts were both created using OpenSCAD [@kintelOpenSCADDocumentation2023], which creates STL files from markup describing the object's geometric composition. 
+The 3D renderings and 3D printed charts were both created using OpenSCAD \citep{kintelOpenSCADDocumentation2023}, which creates STL files from markup describing the object's geometric composition.
 Charts were composed of a platform, with raised text labels centered in front of the A and B groups of bars.
@@
 
-Digital renderings of the generated STL files were created using RGL [@rgl], which integrates into Shiny [@shiny] using the WebGL [@mozillafoundationWebGL2D3D2023] extension.
+Digital renderings of the generated STL files were created using RGL \citep{rgl}, which integrates into Shiny \citep{shiny} using the WebGL \citep{mozillafoundationWebGL2D3D2023} extension.
 The RGL rendering of the STL file was initially angled to correspond to the perspective of default 3D bar charts present in Microsoft Excel, but in the WebGL interactive environment, participants could rotate the charts as they pleased to create a useful estimate of the bar ratio.
 Rendered charts were colored to correspond with the filament color used to print the physical chart for consistency.
-The default `rgl` lighting was replaced with three lights located in fixed positions around the rendered figure. 
-The lights were positioned so that one was behind the rendered figure<!--(x: 910, y: -1000, z:1000)-->, another in front of the figure<!--(x: 780, y: 1030, z: 1000)-->, and one light below the figure<!--(x: 65, y: 15, z: -100)-->.
+The default \texttt{rgl} lighting was replaced with three lights located in fixed positions around the rendered figure.
+The lights were positioned so that one was behind the rendered figure, another in front of the figure, and one light below the figure.
 This arrangement provided a consistent visual experience and minimized shadows that prevented visual estimation of the bars.
 
-## Experiment Design
+\hypertarget{experiment-design}{%
+\subsection{Experiment Design}\label{experiment-design}}
 
@@
 
-<!-- There are 21 kits in total so that each combination of ratios is accounted for. -->
-<!-- Each graph is randomly assigned either an adjacent or separated comparison. -->
+\begin{figure}
+\centering
+\includegraphics{_images/design.pdf}
+\caption{\label{fig:studyDesign}A graphical representation of the study design.
+Only five of the seven ratios were used in each kit; at least one of the smallest or largest ratios were included along with 4 other charts; each selected ratio was displayed in all 3 mediums.
+For each medium \(\times\) ratio combination, the comparison type (separated or adjacent) was randomly determined.
+A total of 21 kits of 3D printed charts were created to include all combinations of the five ratios.}
+\end{figure}
 
-```{r studyDesign, fig.cap = "A graphical representation of the study design.
-Only five of the seven ratios were used in each kit; at least one of the smallest or largest ratios were included along with 4 other charts; each selected ratio was displayed in all 3 mediums.
-For each medium $\\times$ ratio combination, the comparison type (separated or adjacent) was randomly determined.
-A total of 21 kits of 3D printed charts were created to include all combinations of the five ratios."}
-knitr::include_graphics("_images/design.pdf")
-```
+\hypertarget{participant-recruitment}{%
+\subsection{Participant Recruitment}\label{participant-recruitment}}
 
-## Participant Recruitment
-
-One interesting facet of @clevelandGraphical1984 is the participant recruitment methodology: "For each experiment the subjects fell into two categories: (1) a group of females, mostly housewives, without substantial technical experience; (2) a mixture of males and females with substantial technical training and working in technical jobs.
-Most of the subjects in the position-length experiment participated in the position-angle experiment; in all cases repeat subjects judged the position-angle graphs first." 
+One interesting facet of \citet{clevelandGraphical1984} is the participant recruitment methodology: ``For each experiment the subjects fell into two categories: (1) a group of females, mostly housewives, without substantial technical experience; (2) a mixture of males and females with substantial technical training and working in technical jobs.
+Most of the subjects in the position-length experiment participated in the position-angle experiment; in all cases repeat subjects judged the position-angle graphs first.''
 It seems likely that the authors recruited individuals within their respective departments as well as their wives.
@@
 
-```{r demographics, fig.cap="Demographic characteristics of participants in the study.", message=FALSE, warning=FALSE, out.width='100%', fig.width = 9, fig.height = 3.5}
-library(patchwork)
-valid.users$userAppStartTime <- as.POSIXct(valid.users$userAppStartTime,
-           origin = '1970-1-1')
+\begin{figure}
+\includegraphics[width=1\linewidth]{jds_files/figure-latex/demographics-1} \caption{Demographic characteristics of participants in the study.}\label{fig:demographics}
+\end{figure}
 
-age.levels = sort(unique(valid.users$age))
-p1 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
-  group_by(age) %>% 
-  summarise(count = n()) %>% 
-  ggplot(mapping = aes(x = factor(age, levels = rev(age.levels)), y = count)) + 
-  geom_bar(stat = 'identity') +
-  geom_text(aes(x = factor(age, levels = age.levels), y = count+.5, label = age), hjust = 0) + 
-  coord_flip() + 
-  labs(title = 'Age', y = 'Count') +
-  theme_bw() + 
-  scale_y_continuous(expand = expand_scale(mult = 0, add = c(.25, 4))) + 
-  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())
+While the results of any sample with recruitment methodology like this are not generalizable to the public, running our initial study on a comparable population to that of the initial study serves a purpose: other replications, such as \citet{heerCrowdsourcingGraphicalPerception2010b}, used online samples of paid participants, who likely have less experience using charts and graphics and making numerical estimations than those in a statistics department (and individuals who cohabitate with them).
+We have every intention of running this experiment in other populations, but the convenience of online sampling is not compatible with physical objects such as our 3D printed charts, so we will have to recruit in-person participants.
+Thus, an initial study in a population comparable to \citet{clevelandGraphical1984} is a reasonable first step into testing 3D charts.
 
-p2 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
-  group_by(gender) %>% 
-  summarise(count = n()) %>% 
-  ggplot(mapping = aes(x = gender, y = count)) + 
-  geom_bar(stat = 'identity') +
-  geom_text(aes(x = gender, y = count + .5, label = gender), hjust = 0) + 
-  coord_flip() + 
-  labs(title = 'Gender', y = 'Count') +
-  theme_bw() + 
-  scale_y_continuous(expand = expand_scale(mult = 0, add = c(.25, 4))) + 
-  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())
-
-educ.levels = sort(unique(valid.users$education))[c(2,4,3,1)]
-p3 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
-  group_by(education) %>% 
-  summarise(count = n()) %>% 
-  ggplot(mapping = aes(x = factor(education, levels = educ.levels), y = count)) + 
-  geom_bar(stat = 'identity') +
-  coord_flip() + 
-  geom_text(aes(x = education, y = count + .5, label = str_replace(education, "Graduate Degree", "Graduate\nDegree")), hjust = 0) + 
-  scale_y_continuous(expand = expand_scale(mult = 0, add = c(.25, 15))) + 
-  theme_bw() + 
-  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
-  labs(title = 'Education', y = 'Count')
-
-p = p1 + p2 + p3
-p
-```
-
-While the results of any sample with recruitment methodology like this are not generalizable to the public, running our initial study on a comparable population to that of the initial study serves a purpose: other replications, such as @heerCrowdsourcingGraphicalPerception2010b, used online samples of paid participants, who likely have less experience using charts and graphics and making numerical estimations than those in a statistics department (and individuals who cohabitate with them).
-We have every intention of running this experiment in other populations, but the convenience of online sampling is not compatible with physical objects such as our 3D printed charts, so we will have to recruit in-person participants.
-Thus, an initial study in a population comparable to @clevelandGraphical1984 is a reasonable first step into testing 3D charts.
-
-## Data Collection
+\hypertarget{data-collection}{%
+\subsection{Data Collection}\label{data-collection}}
 
@@
 Participants provided informed consent through the applet, and then were asked for demographic information (age, gender, education level).
-Then, participants were shown a "practice" page which allowed them to experiment with the data collection interface and practice estimating the ratio between the bars, as shown in \Cref{fig:practice}.
+Then, participants were shown a ``practice'' page which allowed them to experiment with the data collection interface and practice estimating the ratio between the bars, as shown in \Cref{fig:practice}.
 This practice interface did not provide any feedback as to participant correctness, but was merely intended to familiarize the participants with the questions which would be asked as well as the process of estimating the ratio between the two bars.
 
-```{r practice, fig.cap = "Screenshot of Shiny application practice screen.
+\begin{figure}
+\includegraphics[width=0.8\linewidth]{_images/03-Practice-2} \caption{Screenshot of Shiny application practice screen.
 Three 2D bar charts with different ratios were provided, along with sliders indicating the correct proportion.
-Participants could practice with the sliders and preview the questions that would be asked as part of the task.", out.width = "80%"}
-knitr::include_graphics("_images/03-Practice-2.png")
-```
+Participants could practice with the sliders and preview the questions that would be asked as part of the task.}\label{fig:practice}
+\end{figure}
 
@@
 
-```{r experiment3dRender, fig.cap = "Screenshot of the applet collecting data for a 3D rendered chart task.
-Participants were asked to select which bar (circle or triangle) was smaller, and then to estimate the ratio of the smaller bar to the larger bar.", out.width = "80%"}
-knitr::include_graphics("_images/05-Experiment-05-filled-in.png")
-```
+\begin{figure}
+\includegraphics[width=0.8\linewidth]{_images/05-Experiment-05-filled-in} \caption{Screenshot of the applet collecting data for a 3D rendered chart task.
+Participants were asked to select which bar (circle or triangle) was smaller, and then to estimate the ratio of the smaller bar to the larger bar.}\label{fig:experiment3dRender}
+\end{figure}
 
 Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in \Cref{fig:experiment3dRender}.
-It should be noted that we chose to use a slider for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries [@ruudUncertaintyCausesRounding2014a;@maineriSliderBarsMultiDevice2021].
+It should be noted that we chose to use a slider for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries \citep{ruudUncertaintyCausesRounding2014a, maineriSliderBarsMultiDevice2021}.
 Another benefit of this modification is that it should reduce participants' cognitive load, allowing them to focus on the ratio between bars rather than the numerical translation process.
-This differs from @clevelandGraphical1984, but maintains the spirit of the study while using methods which were not available for data entry at the time.
+This differs from \citet{clevelandGraphical1984}, but maintains the spirit of the study while using methods which were not available for data entry at the time.
 
-# Results
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-All responses ($n = `r nrow(res_all) - nrow(res)`$) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check.
+All responses (\(n = 32\)) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check.
 
-## Midmeans of Log Absolute Errors
+\hypertarget{midmeans-of-log-absolute-errors}{%
+\subsection{Midmeans of Log Absolute Errors}\label{midmeans-of-log-absolute-errors}}
+
 \citeauthor{clevelandGraphical1984} used
-$$\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$$
+\[\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)\]
 to measure accuracy of their participant's responses.
-In their study, log base 2 seemed appropriate due to "average relative errors changing by factors less than 10."
+In their study, log base 2 seemed appropriate due to ``average relative errors changing by factors less than 10.''
 They also added 1/8 to prevent distortions when the errors were close to zero.
-@heerCrowdsourcingGraphicalPerception2010b followed the same analysis method, replicating many (but not all) of the results presented in the original paper.
+\citet{heerCrowdsourcingGraphicalPerception2010b} followed the same analysis method, replicating many (but not all) of the results presented in the original paper.
 
@@
 
-```{r midmeans-log-errors, fig.cap='Midmeans and observed values of log absolute errors for the true ratio of bars.
-Summary lines are computed from raw data using a loess smooth.', message=FALSE, warning=FALSE, out.width='80%'}
-plot.types = c('2D', 'Rendered 3D', '3D Printed')
-names(plot.types) = c('2dDigital', '3dDigital', '3dPrint')
+\begin{figure}
+\includegraphics[width=0.8\linewidth]{jds_files/figure-latex/midmeans-log-errors-1} \caption{Midmeans and observed values of log absolute errors for the true ratio of bars.
+Summary lines are computed from raw data using a loess smooth.}\label{fig:midmeans-log-errors}
+\end{figure}
 
-tmp <- res %>% 
-  group_by(Height, graphtype, plot) %>% 
-  summarize(midmean = mean(response, trim=0.25, na.rm = T)) %>%
-  mutate(type = "Midmean") %>%
-  rename(response = midmean)
-
-res %>% 
-  mutate(type = "Raw Data") %>%
-  bind_rows(tmp) %>%
-  # group_by(Height, graphtype, plot) %>% 
-  ggplot(mapping = aes(x = Height*100, y = response,
-                       color = graphtype, fill = graphtype)) +
-  geom_smooth(se = T, alpha=.125,  expand = F, method = "loess") +
-  geom_point(aes(shape = type, alpha = type, size = type)) +
-  scale_shape_manual("Data", values = c(19, 1)) + 
-  scale_size_manual("Data", values = c(2, 1)) + 
-  scale_alpha_manual("Data", values = c(1, .25)) + 
-  scale_fill_discrete(labels = c('Adjacent', 'Separated')) +
-  scale_color_discrete(labels = c('Adjacent', 'Separated')) +
-  scale_x_continuous(limits = c(0, 100)) + 
-  facet_wrap(~plot, labeller = labeller(plot = plot.types)) +
-  labs(title = '',
-       x = 'True Proportional Difference (%)',
-       y = 'Log Error',fill = 'Comparison Type',
-       color = 'Comparison Type') +
-  theme_bw() +
-  theme(legend.position = 'bottom')
-```
-This is somewhat different than the results in @clevelandGraphical1984 and @heerCrowdsourcingGraphicalPerception2010b; in both cases the midmean log absolute errors increased until about 55% of the true proportional difference and then decreased.
+This is somewhat different than the results in \citet{clevelandGraphical1984} and \citet{heerCrowdsourcingGraphicalPerception2010b}; in both cases the midmean log absolute errors increased until about 55\% of the true proportional difference and then decreased.
 It is possible that this difference is due to the fact that we did not require an explicit numerical estimate of the ratio but instead asked participants to indicate the ratio on a slider (which is essentially a number line).
 It is possible that this change explains the lack of a difference between chart modalities we see, but there are other potential explanations as well.
-One interesting facet of this data is that there is a consistent reduction in log error when the true proportional difference is near 50%. 
-This may be because of implicit anchoring - we can typically bisect a span relatively accurately, which means that it might be expected that when the true proportion is near 50%, we would both notice that and be able to accurately transfer that information to the slider.
-This pattern is more noticable in separated (Type 3) comparisons, which also makes sense - as these comparisons are known to be less accurate, a locally  minimal error near 50% with increased error relative to adjacent (Type 1) comparisons could easily explain the exaggerated trends we see in \Cref{fig:midmeans-log-errors}.
+One interesting facet of this data is that there is a consistent reduction in log error when the true proportional difference is near 50\%.
+This may be because of implicit anchoring - we can typically bisect a span relatively accurately, which means that it might be expected that when the true proportion is near 50\%, we would both notice that and be able to accurately transfer that information to the slider.
+This pattern is more noticable in separated (Type 3) comparisons, which also makes sense - as these comparisons are known to be less accurate, a locally minimal error near 50\% with increased error relative to adjacent (Type 1) comparisons could easily explain the exaggerated trends we see in \Cref{fig:midmeans-log-errors}.
 
-## Linear Mixed Effects Model
+\hypertarget{linear-mixed-effects-model}{%
+\subsection{Linear Mixed Effects Model}\label{linear-mixed-effects-model}}
 
@@
 
-$$y_{ijklm}=\mu+S_i + R_j + G_k +T_l + \epsilon_{ijklm}$$
+\[y_{ijklm}=\mu+S_i + R_j + G_k +T_l + \epsilon_{ijklm}\]
 
@@
 
-- $y_{ijklm}=\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$
+\begin{itemize}
+\item
+  \(y_{ijklm}=\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)\)
+\item
+  \(S_i\sim N(0,\sigma^2_S)\) is the random effect of the \(i^{th}\) subject
+\item
+  \(R_j\) is the effect of the \(j^{th}\) ratio
+\item
+  \(G_k\) is the effect of the \(k^{th}\) display type
+\item
+  \(T_l\) is the effect of the \(l^{th}\) comparison type
+\item
+  \(\epsilon_{ijklm}\sim N(0,\sigma^2_\epsilon)\) is the random error
+\end{itemize}
 
-- $S_i\sim N(0,\sigma^2_S)$ is the random effect of the $i^{th}$ subject
+\begin{table}[tbp]
 
-- $R_j$ is the effect of the $j^{th}$ ratio
+\begin{center}
+\begin{threeparttable}
 
-- $G_k$ is the effect of the $k^{th}$ display type
+\caption{\label{tab:unnamed-chunk-2}Analysis of Fixed Effects}
 
-- $T_l$ is the effect of the $l^{th}$ comparison type
+\begin{tabular}{llll}
+\toprule
+Term & \multicolumn{1}{c}{$\hat{\beta}$} & \multicolumn{1}{c}{95\% CI} & \multicolumn{1}{c}{$t$}\\
+\midrule
+Intercept & 2.49 & {}[1.94, 3.04] & 8.84\\
+Ratio & 0.02 & {}[-0.04, 0.07] & 0.56\\
+Plot3dDigital & 0.15 & {}[-0.14, 0.43] & 1.02\\
+Plot3dPrint & -0.06 & {}[-0.35, 0.22] & -0.44\\
+Type & 0.11 & {}[-0.13, 0.35] & 0.94\\
+\bottomrule
+\end{tabular}
 
-- $\epsilon_{ijklm}\sim N(0,\sigma^2_\epsilon)$ is the random error
+\end{threeparttable}
+\end{center}
 
-```{r, include=T}
-library(papaja)
-library(lme4)
+\end{table}
 
-mod <- lmer(response ~ (1|subject) + ratio + plot + type,
-            data = res)
-modsum <- summary(mod)
-modaov <- anova(mod)
-modaovsum <- summary(modaov)
-
-f_df <- as.numeric(length(modsum$residuals)-modsum$ngrps - 1)
-
-apa_lm <- apa_print(mod)
-apa_table(apa_lm$table, caption = "Analysis of Fixed Effects")
-```
-
-No differences were detected for the true ratio of bars ($t = `r modsum$coefficients[2,3]`$), whether the bars were adjacent or separated ($t = `r modsum$coefficients[5,3]`$), or for the plot nested within the true ratio ($F(`r modaov$npar[2]`, `r f_df`) = `r modaov[2,4]`$).
+No differences were detected for the true ratio of bars (\(t = 0.5590574\)), whether the bars were adjacent or separated (\(t = 0.9381629\)), or for the plot nested within the true ratio (\(F(2, 441) = 1.1110313\)).
 While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes.
 
-## Interactivity with 3D Rendered Charts
+\hypertarget{interactivity-with-3d-rendered-charts}{%
+\subsection{Interactivity with 3D Rendered Charts}\label{interactivity-with-3d-rendered-charts}}
 
 When 3D rendered plots were shown, we recorded the number of user interactions (clicks, rotations) with the WebGL plot.
-\Cref{fig:clicks} shows the distribution of number of clicks over all 3D rendered trials in the experiment, as well as the number of clicks relative to the accuracy measure $\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$.
+\Cref{fig:clicks} shows the distribution of number of clicks over all 3D rendered trials in the experiment, as well as the number of clicks relative to the accuracy measure \(\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)\).
 It does not appear that participants who interacted with the WebGL rendering were more accurate than their peers who did not.
 
-```{r clicks, fig.cap = "Participant clicks on the WebGL interface (left) in order to interact with the plot.
+\begin{figure}
+\includegraphics[width=.45\textwidth]{jds_files/figure-latex/clicks-1} \includegraphics[width=.45\textwidth]{jds_files/figure-latex/clicks-2} \caption{Participant clicks on the WebGL interface (left) in order to interact with the plot.
 Most participants did not interact with the WebGL interface, perhaps because they did not realize that interacting was an option, but some participants utilized the interactivity heavily.
-(Right) Participant interactivity was not associated with increased accuracy.", fig.width = 4, fig.height = 4, out.width = ".45\\textwidth", fig.show = "hold"}
-res %>%
-  filter(plot == "3dDigital") %>%
-  ggplot(aes(x = plot3dClicks)) + geom_bar() + 
-  xlab("# Clicks on WebGL Plot") + ylab("Count") + ggtitle("Distribution of WebGL Interaction") + 
-  theme_bw()
+(Right) Participant interactivity was not associated with increased accuracy.}\label{fig:clicks}
+\end{figure}
 
-res %>%
-  filter(plot == "3dDigital") %>%
-  ggplot(aes(x = plot3dClicks, y = response)) + geom_point() + 
-  xlab("# Clicks on WebGL Plot") + ylab("Count") + ggtitle("Accuracy and Participant Interactivity") + 
-  geom_smooth(method = "lm", se = T) + 
-  theme_bw()
-```
-
-# Discussion and Future Work
+\hypertarget{discussion-and-future-work}{%
+\section{Discussion and Future Work}\label{discussion-and-future-work}}
 
@@
 The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger.
-However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @clevelandGraphical1984 do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.
+However, we should note that the analysis methods used in \citet{heerCrowdsourcingGraphicalPerception2010b} and \citet{clevelandGraphical1984} do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.
 
@@
 Future studies should include an additional fixed 3D perspective bar chart, which will at least enable us to examine whether modern 3D rendering environments allow for more accurate conclusions than fixed 3D perspectives.
-Future iterations of this study will include "traditional" 3D graphs created by Microsoft Excel (that is, graphs with a fixed 3D perspective rendered in 2D).
+Future iterations of this study will include ``traditional'' 3D graphs created by Microsoft Excel (that is, graphs with a fixed 3D perspective rendered in 2D).
 This option will allow us to examine fixed perspective 3D plots compared to 3D renderings and 2D plots; it will also enable online data collection in addition to the in-person data collection used in this experiment.
@@
 Another interesting aspect of our study is that the method used to record participant estimates is different from the method used in the original study as well as Heer \& Bostock's replication study.
-A method similar to our slider input, marking position on a line, was used in @spenceVisualPsychophysicsSimple1990, but Spence asked participants to estimate $A/(A+B)$, where we asked participants to estimate $A/B$; thus, our results are still not directly comparable to previous studies.
+A method similar to our slider input, marking position on a line, was used in \citet{spenceVisualPsychophysicsSimple1990}, but Spence asked participants to estimate \(A/(A+B)\), where we asked participants to estimate \(A/B\); thus, our results are still not directly comparable to previous studies.
 We expect that the specific ratio estimated would also have an effect on observed participant errors.
@@
 
-<!-- The goal of this study was to verify that previous research into graphics is upheld under a new set of circumstances.
-It started with a question, followed by research into prior work, and then by designing and conducting a study.
-This general framework of research is useful to all fields of study and is taught in most introductory statistics courses.
-The next step of this study is to collect more data while simultaneously providing students the opportunity to learn about the research process from both the perspectives of an experiment participant and the views of the researchers.
-At University of Nebraska-Lincoln, students taking introductory statistics will have the opportunity to participate in the experiment while also completing pre/post experiment surveys and reflections on a two-page abstract and a presentation of results.
--->
+\hypertarget{supplemental-material}{%
+\subsection{Supplemental Material}\label{supplemental-material}}
 
-## Supplemental Material
-Stimuli, code, and data for this experiment are provided at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).
+Stimuli, code, and data for this experiment are provided at \url{https://github.com/TWiedRW/2023-JDS-3dcharts}.
+
+\bibliography{JDS2023.bib}
+\bibliographystyle{jds}
+
+\end{document}

### 2024-01-22 jds.rmd
Update references for 3D sucking

@@
 For both experience levels, participants were more confident in their answers when using 2D graphs. 
-There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may better encode information - for instance, when X and Y are used to represent spatial dimensions, it may be preferable to use a 3D chart to convey numerical information instead of using color, which is perceived much less accurately.
+There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may provide a more natural way to encode information, at least in theory.
+For instance, when X and Y are used to represent spatial dimensions, it may be preferable to use a 3D chart to convey numerical information instead of using color, which is perceived much less accurately.
 @brath_3d_2014 highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.
 
-There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data. 
+There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data [@vanderplasTestingStatisticalCharts2020;@wilkeDonGo3D2020;@morganwallIntroducing3DGgplots20191006].
 However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.

### 2024-01-22 jds.tex
Update references for 3D sucking

@@
 For both experience levels, participants were more confident in their answers when using 2D graphs.
-There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may better encode information - for instance, when X and Y are used to represent spatial dimensions, it may be preferable to use a 3D chart to convey numerical information instead of using color, which is perceived much less accurately.
+There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may provide a more natural way to encode information, at least in theory.
+For instance, when X and Y are used to represent spatial dimensions, it may be preferable to use a 3D chart to convey numerical information instead of using color, which is perceived much less accurately.
 \citet{brath_3d_2014} highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.
 
-There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data.
+There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data \citep{vanderplasTestingStatisticalCharts2020, wilkeDonGo3D2020, morganwallIntroducing3DGgplots20191006}.
 However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study \citeyearpar{heerCrowdsourcingGraphicalPerception2010b} in 2010.\\

### 2024-01-22 jds.rmd
Residual plots

@@
 
-mod <- lmer(response ~ (1|subject) + ratio + plot + type,
+mod <- lmer(response ~ (1|subject) + factor(ratio)+plot+factor(type),
             data = res)
@@
 While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes.
+
+Model diagnostic plots show slight violations of normality with the error term, particularly with right-skewed residuals.
+Plots of the data for each ratio tend to suggest that some participants might have tried rounding to the ratio increments of fives and tens. 
+For example, participants tended to favor estimating the ratio of 0.562 to around 0.5.
+Since linear models are reasonably robust to violations of normality, we will assume that the model reasonably fits.
+Plots of the residual diagnostics can be found under the supplemental material in \Cref{fig:residualplots}.
 
@@
 Stimuli, code, and data for this experiment are provided at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).
+
+```{r residualplots}
+#| fig-width: 3
+#| fig-height: 3
+#| fig-cap: "Residual diagnostic plots for the linear mixed model."
+#| fig-subcap:
+#| - Fitted vs. predicted residuals
+#| - Q-Q plot
+#| out-width: ".49\\linewidth"
+#| fig-show: "hold"
+#| layout-ncol: 2
+
+r = residuals(mod)
+p = predict(mod)
+
+ggplot(mapping = aes(x = p, y = r)) +
+  geom_point() + 
+  geom_hline(yintercept = 0, linetype = 'dashed', color = 'red') + 
+  theme_bw() + 
+  labs(x = 'Fitted value',
+       y = 'Residual')
+
+ggplot(mapping = aes(sample = r)) + 
+  geom_qq() + 
+  geom_qq_line(color = 'red', linetype = 'dashed') + 
+  theme_bw() + 
+  labs(x = 'Theoretical Quantiles',
+       y = 'Sample Quantiles')
+
+```

### 2024-01-22 reviewer_response.qmd
point-by-point response

---
title: "Response to Reviewers"
format: pdf
---

## Editor Suggestions

- A more extensive literature review is needed; for instance, the authors claim on page 2 that "...the situation is different now than it was in 1984...it has even changed since 2014," but there are very few citations about this. I think even mentioning some application areas (with specific papers cited) where 3D charts and printed 3D charts are increasingly prevalent is important here to motivate why readers should be interested in this study.

> Fixed. It would be hard to cite printed 3D chart prevalence as there are very few applications of this at the moment (though there are a few US maps and things on Thingiverse). We expect they will be more common moving forward, though, which is why it's important to revisit the perception of these charts.

- Given that the authors claim this is a partial replication of Cleveland and McGill's study, I suggest creating a whole section devoted to more deeply explaining their work, strengths/limitations, and how the new study differs and improves on their work. This will help particularly for those not familiar with their study.

> We have added some limited additional clarity, but have also cited papers which go through the study in more depth, including the Annual Review paper "Testing Statistical Charts: What Makes A Good Graph" by Vanderplas et al.

- In its current state, the mixed model implemented in Section 3.2 needs to be vastly simplified; as both the authors and Reviewer 1 remark, this study simply has too few participants for this parameterized model to be sufficiently fit. At the very least, I would suggest fitting a simpler model and discussing this further (including assumptions), perhaps leaving a brief description of this mixed model as something of interest in a follow-up study or in an Appendix.

> The gam model we are currently using (after following up on Reviewer 2's questions) is a bit simpler but also much more interesting. We hope this model provides more useful results than the mixed-effects model in the previous version of this paper, which was pretty clearly not appropriate given the nonlinearity in the data.

## Reviewer 1 Suggestions

- This experiment seems very far from a replication of the source material. This is somewhat softened in the paper by calling it a “partial” replication and acknowledging many of the shortcomings of the study, but even with this: the statement seems inaccurate. My reasons are as follows:
  - The participants were allowed to practice, with the true feedback given. Was there such a practice round in the original study? This seems like something that would improve later performance.

> Participants were practicing with the slider mechanism, but no feedback was provided as to the correctness of the slider position. During beta testing of the survey, we noticed that some participants were defaulting to estimating A/A+B instead of A/B, and so we added a clarifying demo page.

  - A sliding error bar is a significantly different mechanism than simply stating a ratio. The experiment in this paper is, in a sense, asks participants to translate from a visual domain to a second visual domain. If the aim of the former visualization is to get to the latter visualization, why not present the latter visualization in the first place? What information would we like someone consuming a graph to get? This paper seems to want to comment on which medium best communicates the true numbers themselves, but what is tested is whether someone is able to translate one visual medium into another visual medium. It is the authors who then translate from this latter visual medium to the ratio.

> We agree that a sliding bar is different from estimating a ratio - the latter requires both the visual assessment of the ratio as well as translating this visual assessment into a number. We had hoped to reduce cognitive load by foregoing the numeric translation step; an added benefit to this is that it is typically very difficult to model numerical estimates from participants due to rounding effects that are censored but are not censored uniformly. We do have data from a different study suggesting that estimates using sliders are similar to numerical estimates (minus the rounding effects) - this paper will be submitted within the next 6 months or so but is not quite finished. At any rate, while we agree that the method for recording data is different (and that this might be an explanation for our failure to explicitly replicate the effects in the original paper), we are also working with very different technological constraints. The charts in Cleveland & McGill had axes which could be used to help participants with estimation; our 3D printed charts do not, which is one very real limitation of that medium. Because of this issue, as well as the cognitive load and technological considerations (e.g. we have the option of a slider, in 1984 that was much less feasible), we stand by our decision to deviate from the original protocol.

  - Why are the data where a participant incorrectly identified the smaller bar removed from the study? Did they do this in the original study? Is this information irrelevant? How many observations were removed?

> The original study is not particularly specific about the methods employed. Attention check questions are fairly standard in online studies, and we have added some language clarifying the number of observations removed. Fundamentally, if participants don't identify the smaller bar, it is not clear what the slider ratio would even represent, as it is fundamentally confined to $[0,1]$ and the participant isn't certain about what the denominator of the ratio is.

  - The original experiment also uses stacked bar charts; this experiment only compares grouped bar charts.
With these in mind, I recommend that this paper not argue that this experiment is in any way a replication, but a completely new experiment inspired by previous work. This will require that the authors acknowledge the poor sampling techniques used as their own.

> "Partial" is in this case a reference to the fact that Cleveland & McGill's 1984 study was an extensive assessment of many different simple graphical elements; we are specifically focusing in this study on the simple element of comparisons on an aligned scale (e.g. grouped, but not stacked, bar charts). When extending C&M to 3D, it is necessary to focus on one simple element at a time, if only because of technological constraints - the technology exists (but is extremely expensive) to print multiple color objects that would be needed for stacked bar charts, and it is entirely unclear how one would 3D print a scatterplot. While we agree that there are not very many 3D-printed charts at this point in time, it does seem to be a promising area for visualization and making data tactile, but we have to lay the groundwork first. We hope to follow up with a 3D heatmap style plot, where the 3rd dimension is not redundant, but comparing to existing studies seemed wise.

- There are details missing in the description of the experiment.
Are the ratios participants are asked to estimate the same across the three
plot types (2D, 3D, and 3D printed)? Are all plots compared either sepa-
rated or not? How many times is a participant asked to estimate a ratio?
There are a lot of missing details here. I would recommend an ordered list
of tasks a participant is asked to do that emphasizes what variables (true
ratio, separated/not separated, graph dimension) are controlled or varied
at each step

> We have revised Figure 2 and the experiment description to clarify the design.

- The linear mixed effects model does not represent a thorough
statistical analysis of the data. This is true for several reasons. The
model seems very heavily parameterized for such few data. With my
confusion described in the previous point, it is not immediately clear to me
how many fixed effects there are, but it seems like there are 26 parameters
(two s.d.s from the random effects, 7 ratios, 2 graph types nested in 7
ratios (7*2), and 3 graph types) to fit using a small dataset. This is too
ambitious of a model. Furthermore, the normality assumptions on the
error term and the random effect are not assessed in any way. I would
recommend a simplier model with a proper assessment of the model fit.
Examining p-values is meaningless unless the model accurately emulates
the data generating mechanism.If a linear model seems inappropriate for these data (which seems likely
without a random sample), the authors may want to forsake p-values and
consider a more Machine Learning approach.

> We have revised the model and are now using a generalized additive model with spline terms for ratio effects fit by comparison type and display mode and a random effect for participant. While this model has more flexibility, it does have fewer parameters and effective degrees of freedom than the previous model, and examination of random effect terms and model residuals suggest that the model fits reasonably well.

- Please clearly state the hypotheses before calculating p-values.

> We do not include model fit statistics for hypothesis testing purposes, but as a mechanism to graphically examine the components of the variance. While p-values are technically provided in the model output, we primarily include these tables in order to provide information with which to interpret the accompanying figures.

- The paper implies a consensus from existing literature, but the
literature review presents several conflicting viewpoints. In the
abstract, the paper claims that numerous studies advocate to avoid 3D
visualizations when the third dimension does not convey any additional
information to the user, yet it is never explicitly stated which studies claim
this. The papers mentioned in the second paragraph of 1.1 comment on
accuracy and encoding information (with conflicting conclusions) and no
mention is given of any “consensus on 2D v. 3D.” This paper would benefit
greatly from a more focused literature review.

> We have improved the literature review; while the "3D is bad" advice is almost a trope in the graphics world, it is reasonable that people who are not visualization experts may not be familiar with this advice -- it is so common that it is even addressed by authors of packages meant for 3D graphics, such as rayshader.

- Figure 6 seems to imply a significant confounding variable that
is neither addressed nor investigated. It would appear that there is
a drop in the midmean of the log error when the true proportion is around
50%. With this observation, I find the argument that certain proportions
are easier to distinguish than others quite compelling. This should be
investigated. Furthermore, if this is true, the graph that a participant
randomly selects would have a strong effect on their ability to determine
a correct ratio.

> When we switched to using the generalized additive model, we spent some time investigating this effect. This is a fairly common result in perceptual experiments resulting from anchoring at endpoints and the midpoint of the line; it does not seem to result from the default slider value of 0.5. This is an interesting finding that merits a follow-up study.

- Abstract: As mentioned earlier, the ’numerous studies’ are not explicitedly listed anywhere.

> This has been addressed

- The acronym EPTs is introduced, but never used. Perhaps most blaring: the paper fails to use it immediately after introducing it (at the beginning of the next sentence).

> This has been addressed

- What is Mechanical Turk? Please provide a short explanation.

> Added a short explanation

- Relative language such as “often” and “largely” detract from the point the paper is trying to make, especially when this point does not seem to be backed up in the literature review. For instance, each of these words appear in the first sentence of Section 1.2.

> Some of this relative language is because it is difficult to make absolute statements about charts - certainly some people create charts in 3D (as we have in this study), but the vast majority of visualizations are intended for display on flat, 2D screens or papers. Similarly, we cannot guarantee that perceptual heuristics will always be applied in certain cases - for instance, people who have extreme amblyopia or have been blind in one eye since birth are immune to the line-width illusion, while those with binocular vision cannot overcome it. There are very few absolutes in perceptual research. We have attempted to clarify the language, but it is very difficult to remove all relative language here.

- Also on the first sentence of 1.2: I have failed to understand this point. What is the visual system that is “assuming” something? Is it a computer? Our eyes? Is there evidence that chart perception is so affected by this assumption?

> We have cited papers indicating that chart perception is indeed affected by the human visual system's heuristics. We have also added language clarifying that we are talking about the human perceptual system. In addition, addressing the line-width illusion using an example allows us to directly provide evidence to the reader that these effects are real.

- Please give a short explanation of the line-width illusion.

> Added.

Do the papers at the end of the first paragraph discuss the specific idea that 3D perceptions affect perceptions of 2D objects? Or do they just talk about these specific illusions? If it’s the former: move them up in this paragraph. The beginning of this paragraph feels like something that the authors are saying without evidence, as written.

> I believe this has been addressed because we re-wrote the paragraph explaining the connection between the line-width illusion and 2D/3D perception. Vanderplas & Hofmann (2015) in JCGS makes the case for the connection between 3D perception and the line width illusion in a compelling way; there is also an unpublished case study involving someone without binocular depth perception presented in Chapter 3.3 of https://github.com/srvanderplas/Dissertation/blob/master/thesis.pdf, if the reviewer is interested.

- second paragraph, first sentence: Did the subjects select these graphs for use, or did they better understand them? Are subjects here the people comprehending the graphs, or creating them? This is not clearly written.

> These short summaries of papers are all in the context of user studies, so these are designed experiments suggesting that participants prefer 2D charts for extracting information (that is, participants are presented with charts; they are not creating them).

- second paragraph: Why are the experienced subjects referred to as ’managers?’

> This study was evaluating usage of charts in a corporate setting - the experienced participants were literally managers within the company.

- second paragraph, second-to-last sentence, ’there are times where 3D graphs may better...’: Are these the same ’times’ when 2D graphs perform better? The two parts of this sentence could be interpreted as disparate ideas.

> As with most graphics research, there are always papers with contradictory results. We have tried to clarify the overall tone of this paragraph to assist with any confusion this may cause, but human subjects research is inherently messy.

- 1.2, second paragraph, last sentence: ’its’ should be ’their.’

> I believe this has been resolved

- 1.2, third paragraph, third sentence: ’has’ should be ’have’, unless ’Digital Graphics’ is the proper name for a field. In which case, capitalize ’graphics.’

> Fixed, thanks

- 1.2: Is this paper really ’reconsidering’ 3D charts? It does not sound like the literature really put them down.

> We've done a better job at supporting the idea that 3D graphics have been denigrated in visualization literature for at least the last 70 years; they were blamed for misleading graphics in "How to Lie with Statistics" in 1954, and I would suspect there are older snide comments, but they get harder to find in the pre-digital era. While 3D charts may be acceptable in many fields, in graphics and visualization they are very frequently condemned, mocked, and denigrated, perhaps without sufficient evidence.

- 1.2, ’these charts provide the opportunity...’: Make it clear that this is in reference to the 3D printed charts specifically.

> Fixed

- 1.3: Briefly discuss Cleveland and McGill’s theory, or do not mention it.

> Cleveland & McGill's papers produced a hierarchy of simple graphical elements, and while they also produced various theories, in this case, we're not actually citing any of their theories. In any case, I believe the language that triggered this comment has been removed in the edits.

- Figure 2: What’s the difference between Type 1 or Type 3?

> This has been clarified in the introduction, and we have used adjacent and separated as terms throughout this paper because they are a bit more intuitive.

- Discuss, or at least mention, Figures 2 and 3 in the text.

> Done

- 2.2, last paragraph, first sentence: What is covered in these papers? Are these two methods for digital rendering techniques? How are they different (and why does one require ’integration’ into another)?

> These are software packages used in this work - Shiny creates the web applet used to collect data, WebGL provides the interactive 3D renderings to allow us to show 3D plots within the browser.

- 2.3: Is the ratio the same across graph type? How do we know that participants didn’t use information from another plot type to drive their answers?

> Yes, the ratio is the same. As this was a relatively informal sample, we did check with participants verbally to see if they had any comments or had observed any interesting design quirks; most had no idea that the ratios were repeated. If statisticians didn't notice, we are fairly confident that their roommates who are not statisticians were also not aware. Examining the data, in 3/164 cases of (participant x ratio) trials participants had 0 variance on the provided estimates; in every case, these trials were ratios adjacent to 0.5 where participants did not adjust the slider. While this doesn't completely confirm that participants *couldnt* use information from other trials, it suggests that it was not a significant factor.

- 3.2, first paragraph, ”(through given the midmean...”: As written, this sounds like you are not expecting to find any log absolute error plots.

> This language is no longer present

- 3.2: Give more details on this Heer & Bostock model. Is there president for using a linear mixed effects model in this case.

> We've moved away from the LME towards a gam model (for which there is no precedent); if we could get the data from Heer & Bostock, we could examine differences between their model/data and ours, but that data is no longer available and we have reached out by email but have gotten no response.

- 4, first sentence: I don’t think that your literature review really suggested this.

> Addressed earlier

- 4, first paragraph, last sentence: What discrepancy? The one between you in the literature? If so, you really need to argue a stronger consensus in the literature, if there is one.

> Addressed earlier

- 4: I wouldn’t highlight that this paper does significance testing without more careful modeling.

> Addressed with better modeling and a bit more caution

- 4: “measurement method” should be “the measurement method.”

> Fixed

## Reviewer 2 Suggestions

> Thank you so much for your careful comments and suggestions - they have greatly improved the paper and we appreciate the thought you put into this.

There are a few edits that I view as necessary to strengthen the paper. These edits will help clarify important
issues for the reader, and also help avoid pitfalls.

-  I recommend adding a brief explanation and refernce for the line-width illusion in section 1.2. This would help readers unfamiliar with graphics research.

> We've added a figure demonstrating the sine illusion (as the simplest case of the line width illusion) and discussed in more depth the connection between this illusion and 3D graphics.

- At the end of section 1.2, it would be useful to add references, if there are any, for the comments of accessibility.

> We've added a paper by Chancey Fleet, who is blind and runs the NY Public Library lab for making visualizations available to the visually impaired. Hopefully that is sufficient.

- Section 2.3 could be clarified. Undertstanding the exact design with only the words in section 2.3 was difficult. Figure 2 helps, but was not cited in the text. Please clarify this a bit. This will also help the reader understand the model later, since the design is linked with the G(R) inclusion rather than having main effects and then an interaction.

> We've added a diagram which we hope will make the design much clearer. In addition, we've simplified the model so that hopefully it is easier to understand how the model works relative to the design.

- In Figure 6 the default smoother in ggplot2 was used. This accentuates the dip around x = 0.5, and allows the smallest and largest x values to have good bit of pull on the smoother. Since these features don’t seem to be important to the narrative, I recommend an adjustment be made.

> We have adjusted the figure to use the underlying data instead of the midmeans; this ensures that the effect is not just due to the default smoother (though we understand where that perception came from). In the process, we decided to use a gam model instead of the linear mixed effects model to account for the clear nonlinearity of the ratio relationship to the error. This has made the paper much more interesting (in our opinion), so thank you for calling this out, because it led to a thorough dive into the data that was quite rewarding.

- Pages 7-8. I recommend avoiding the phrase “testing for significant differences”, since this could be read incorrectly as not testing hypotheses set a priori (by someone rushing).

> I can't find that phrase anymore, but I hope that we've been sufficiently clear in our description that we're using the model in an informative/exploratory context more than as a means to do hypothesis testing.

- Page 8, bullet 1, I recommend acnowledging the full subscript in words

> I'm not sure what this is referring to, as there are no bullets on Page 8 that I can see.

- Details are needed as to what tests you ran from the LME. Only reporting p-values is not informative enough.

> Fair, but no longer relevant with the gam model. We've included model tables for informational purposes, but we're not really using them for significance testing (which is just as well since very few effects were significant, and even those effects need follow-up studies to ensure they aren't just an artifact of the experimental design and model-fitting process).

- In section 4, I found myself wondering if you would tell whether a participant interacted with a plot in the
shiny app. For example, could you track whether they clicked or dragged?

> Yes, we can, and we've added a section about this. We were surprised by how few people did actually interact with the 3D rendering; this may be a case where we need to add a 3D render to the tutorial so that people can learn how to play with it before the study. Thanks for the suggestion! In more recent studies, we've actually taken to returning the rotation matrix on each interaction and we're looking forward to trying to analyze that data!

- Section 1.1, paragraph 2: “that 2D graphs” should read “than 2D graphs”

> Fixed

- Page 2, paragraph 2, “Digital graphics has” should be “Digital graphics have”

> Fixed

- In section 1.3, I would continue to use the Cleveland and McGill (1984) citation style. That’s a stylistic preference, however.

> Fixed

- Page 4, paragraph 1. The citations Murdoch and Adler (2023) and Change at al. (2023) should be changed to rgl and shiny with parenthetical citations.

> Fixed

- Page 4, end of section 2.4. Change “3” to “Figure 3”.

> Fixed

- Figures 4 and 5 were somewhat hard to read. Could these be clearer?

> If these are the app screenshots, I am not sure how to make them clearer while still meeting the image resolution requirements for the journal, but we have tried.

- Section 3, how many cases were discarded? A sentence or two would avoid readers asking this question
and getting distracted.

> Fixed, thanks. Good catch.

-  Page 9, first line “impact of measurement” could be “impact of the measurement”

> Fixed

-  In the supplemental materials, please change the relative file paths in _code so the user could start in that
main folder and not have to change the paths to rerun the analysis.

> Fixed, thanks.

### 2024-01-22 jds.rmd
point-by-point response

--- 2024-01-22 jds.rmd
@@
 As a result of the Cleveland and McGill [-@clevelandGraphical1984] study, it is possible to assemble an ordering of perceptual accuracy for the elements of length, position, and angle. 
-@heerCrowdsourcingGraphicalPerception2010b replicated some parts of @clevelandGraphical1984 in an online setting using Mechanical Turk, largely validating the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.
+@heerCrowdsourcingGraphicalPerception2010b replicated some parts of @clevelandGraphical1984 in an online setting using  Mechanical Turk, a platform for crowdsourcing human tasks that has been used for (among other things) assembling machine learning datasets.
+This replication study largely validated the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.
 
@@
 ## 3D Graphical Perception
-Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
-This occasionally causes problems: the line-width illusion[@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013], for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in @fig:sine-illusion).
-While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception.
-Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the stimuli.
+Chart perception is affected by the human visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial.
+In addition, these artificial stimuli are typically created and presented in two dimensions.
+The artificial scenes we create for data visualization have been documented to cause problems: for instance, the line-width illusion [@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013] has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, hammock plots, and Sankey diagrams.
+The line width illusion causes lines or error bands which are actually the same length to appear shorter at points with large amounts of curvature; a simple example of the classic sine illusion (one version of the line width illusion) is shown in \Cref{fig:sine-illusion}.
+While this illusion is problematic in two dimensions, when the depth cues are actually present in a three-dimensional situation, the same implicit visual heuristics contribute to accurate size perception (as in the middle panel of \Cref{fig:sine-illusion}).
+Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the visual display relative to the training of the visual system and the heuristics which are activated in order to make sense of the display.
 
@@
 
-The use of 3D graphics have been explored in multiple studies. 
+The use of 3D graphics have been explored in multiple studies, with mixed results. 
 @fisher_data_1997 explored user preference for 2D or 3D charts and found that subjects tended to prefer simpler 2D graphs when tasked with extracting information. 
@@
 @brath_3d_2014 highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.
+Nevertheless, it is extremely common in visualization literature to recommend that 3D plots be avoided at all costs [@vanderplasTestingStatisticalCharts2020;@wilkeDonGo3D2020;@morganwallIntroducing3DGgplots20191006].
 
-There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data [@vanderplasTestingStatisticalCharts2020;@wilkeDonGo3D2020;@morganwallIntroducing3DGgplots20191006].
+There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data [@huffHowLieStatistics1954]. 
 However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.
-Digital graphics have developed quickly, along with the hardware necessary to support these software developments. 
-As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
-As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].
+Computer graphics and rendering technology have developed quickly, along with the hardware necessary to support these software developments. 
+As a result, we have much more natural virtual rendering of 3D objects, and we can also print graphics in three dimensions using mass-market, relatively inexpensive 3D printers, moving artificial charts into a more natural, physical setting.
+As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].
 
@@
 <!-- This is included because of a graphics paper section comment that really annoyed me -->
-In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from @clevelandGraphics1984 in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs.
+In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from @clevelandGraphical1984 in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs.
 The next section presents the results, and we conclude the paper by discussing this experiment in the context of existing work on the perception of 2D and 3D graphical elements.
@@
 
+```{r, include = F}
+res %>%
+  group_by(subject, ratio) %>%
+  summarize(mean = mean(byHowMuch), sd = sd(byHowMuch))
+
+```
+
 # Results
@@
 ```{r midmeans-log-errors, fig.cap='Midmeans and observed values of log absolute errors for the true ratio of bars.
-Summary lines are computed from raw data using a loess smooth.', message=FALSE, warning=FALSE, out.width='80%'}
+Summary lines are computed from raw data using a loess smooth.', message=FALSE, warning=FALSE, out.width='\\textwidth'}
 plot.types = c('2D', 'Rendered 3D', '3D Printed')
@@
 
-## Linear Mixed Effects Model
+## Generalized Additive Model
 
-In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a linear mixed effects model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio.
-This allows us to test for significant differences (though given the midmean log absolute error plots, we do not expect to find any) as well as to quantify effect sizes for future studies.
+In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a generalized additive model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio.
 The formal statistical model is as follows:
 
-$$y_{ijklm}=\mu+S_i + R_j + G_k +T_l + \epsilon_{ijklm}$$
+$$y_{ijklm}= s(R_j * T_l) + s(R_j *G_k) + T_l + G_k + s(S_i) + \epsilon_{ijklm}$$
 
@@
 
-- $S_i\sim N(0,\sigma^2_S)$ is the random effect of the $i^{th}$ subject
+- $s(S_i)$ is the random effect of the $i^{th}$ subject
 
-- $R_j$ is the effect of the $j^{th}$ ratio
+- $s(R_j*T_l)$ is the effect of the $j^{th}$ ratio, by comparison type
+
+- $s(R_j*G_k)$ is the effect of the $j^{th}$ ratio, by display type
+
+- $T_l$ is the effect of the $l^{th}$ comparison type
 
@@
 
-- $T_l$ is the effect of the $l^{th}$ comparison type
-
 - $\epsilon_{ijklm}\sim N(0,\sigma^2_\epsilon)$ is the random error
 
-```{r, include=T}
+```{r, include=F}
 library(papaja)
@@
 
-mod <- lmer(response ~ (1|subject) + ratio + plot + type,
+mod <- lmer(response ~ (1|subject) + ratioLabel + plot + type,
             data = res)
@@
 
-No differences were detected for the true ratio of bars ($t = `r modsum$coefficients[2,3]`$), whether the bars were adjacent or separated ($t = `r modsum$coefficients[5,3]`$), or for the plot nested within the true ratio ($F(`r modaov$npar[2]`, `r f_df`) = `r modaov[2,4]`$).
-While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes.
+<!-- No differences were detected for the true ratio of bars ($t = `r modsum$coefficients[2,3]`$), whether the bars were adjacent or separated ($t = `r modsum$coefficients[5,3]`$), or for the plot nested within the true ratio ($F(`r modaov$npar[2]`, `r f_df`) = `r modaov[2,4]`$).
+-->
+<!-- The form of this linear mixed effects model assumes a linear relationship between ratio and response, but we know from \Cref{fig:midmeans-log-errors} that the actual relationship appears to be nonlinear.
+-->
+A generalized additive model with random effects for subject was fit using the `mgcv` package [@mgcv1;@mgcv2;@mgcv3].
+Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}.
+It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in @hollandsBiasProportionJudgments2000.
+
+While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only `r sum(res$byHowMuch == 50)` trials (`r sprintf("%.2f%%", mean(res$byHowMuch==50)*100)`) had slider values at exactly 0.5, and all but `r sum(res$byHowMuch == 50 & !res$ratioLabel %in% c(46.4, 56.2))` were for ratios on either side of 0.5 (.464, .562).
+This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5.
+In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types.
+It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum.
+A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points  could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.
+
+```{r gam}
+library(mgcv)
+library(gratia)
+res2 <- res %>%
+  mutate(participant = factor(as.numeric(factor(subject)))) %>%
+  mutate(ratio_prop = ratioLabel/100) %>%
+  mutate(type = factor(type, levels = 1:2, labels = c("Separated", "Adjacent"))) %>%
+  mutate(plot = factor(plot)) %>%
+  mutate(plottype = interaction(plot, type))
+# res2
+
+gam_mod <- gam(response ~ s(ratio_prop, by = plot, k = 6) + s(ratio_prop, by = type, k = 6) +  plot + type + s(participant, bs = "re"),
+               data = res2, method = 'REML')
+# gam_mod
+
+gam_sum <- summary(gam_mod)
+```
+
+```{r, results = 'asis'}
+gam_sum$p.table %>% knitr::kable(digits = 4, caption = "Parametric coefficients in gam model.", label = "gam-param", booktabs = T)
+
+gam_sum$s.table %>% as.data.frame() %>% 
+  mutate(Smooth = rownames(.),
+         Smooth = str_replace(Smooth, "ratio_prop", "Ratio")) %>%
+  select(Smooth, everything()) %>%
+  knitr::kable(digits = 4, row.names = F, caption = "Approximate significance of smooth terms in gam model.", label = "gam-smooth", booktabs = T)
+```
+
+```{r gam-pred-chart, fig.cap = "Predictions with standard errors for fixed effects in a generalized additive model with splines for ratio by comparison and ratio by display method.
+It is clear that the separated comparisons are easier to estimate when small, right around 50\\%, or large, and harder to estimate between these points.
+What is interesting is that no such trend is present for adjacent comparisons.
+", fig.width = 7, fig.height = 3, out.width = "\\textwidth"}
+# 
+# coef(gam_mod)
+# draw(gam_mod, ncol = 4)
+
+newdata <- expand.grid(ratio_prop = seq(0.15, .85, .001), type = c("Separated", "Adjacent"), plot = unique(res2$plot), participant = 1) %>%
+  mutate(plottype = interaction(plot, type))
+pred <- predict(gam_mod, newdata = newdata, type = "terms", se.fit = T) %>% as.data.frame()
+
+pred_se <- select(pred, matches("se.fit")) %>%
+  select(-matches("participant"))
+pred_fit <- select(pred, matches("^fit.")) %>%
+  select(-matches("participant"))
+
+newdata %>%
+  mutate(pred = rowSums(pred_fit), 
+         se = rowSums(pred_se)) %>%
+  ggplot(aes(x = ratio_prop, y = pred + gam_mod$coefficients[1], color = type)) + 
+  geom_ribbon(aes(ymin = pred + gam_mod$coefficients[1] - 1.96*se, ymax = pred+ gam_mod$coefficients[1] + 1.96*se, fill = type), alpha = .25) +  
+  geom_line() + facet_grid(.~plot) + 
+  scale_color_discrete("Comparison", na.translate = F) + scale_fill_discrete("Comparison", na.translate = F) +
+
+  geom_point(data = res, aes(x = ratioLabel/100, y = response, color = factor(type, levels = 1:2, labels = c("Separated", "Adjacent")))) +
+  theme_bw() + ylab("Prediction") + xlab("True Proportion Difference") + 
+  xlim(c(0, 1))
+
+```
+
+While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes; if we could get access to their data, this would allow for investigating whether the differences identified in our model were an artifact of the measurement method (e.g.
+slider vs.
+numerical entry) or if this effect was present in previous studies and can now be identified because of more advanced statistical modeling techniques.
 

### 2024-01-22 jds.tex
point-by-point response

@@
 Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
-This occasionally causes problems: the line-width illusion\citep{vanderplasSignsSineIllusion2015, daySineIllusion1991, hofmannCommonAnglePlots2013}, for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in \citet{fig:sine-illusion}).
+This occasionally causes problems: the line-width illusion\citep{vanderplasSignsSineIllusion2015, daySineIllusion1991, hofmannCommonAnglePlots2013}, for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in \Cref{fig:sine-illusion}.
 While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception.
@@
 In order to explore perception of fully 3D graphics, it is prudent to start with the simplest possible 3D graphic: one in which the third dimension is not necessary, so that we can easily compare to two-dimensional representations without loss of information.
-In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from \citet{clevelandGraphics1984} in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs.
+In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from \citet{clevelandGraphical1984} in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs.
 The next section presents the results, and we conclude the paper by discussing this experiment in the context of existing work on the perception of 2D and 3D graphical elements.
@@
 \begin{figure}
-\includegraphics[width=0.8\linewidth]{jds_files/figure-latex/midmeans-log-errors-1} \caption{Midmeans and observed values of log absolute errors for the true ratio of bars.
+\includegraphics[width=\textwidth]{jds_files/figure-latex/midmeans-log-errors-1} \caption{Midmeans and observed values of log absolute errors for the true ratio of bars.
 Summary lines are computed from raw data using a loess smooth.}\label{fig:midmeans-log-errors}
@@
 
-\hypertarget{linear-mixed-effects-model}{%
-\subsection{Linear Mixed Effects Model}\label{linear-mixed-effects-model}}
+\hypertarget{generalized-additive-model}{%
+\subsection{Generalized Additive Model}\label{generalized-additive-model}}
 
-In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a linear mixed effects model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio.
-This allows us to test for significant differences (though given the midmean log absolute error plots, we do not expect to find any) as well as to quantify effect sizes for future studies.
+In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a generalized additive model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio.
 The formal statistical model is as follows:
 
-\[y_{ijklm}=\mu+S_i + R_j + G_k +T_l + \epsilon_{ijklm}\]
+\[y_{ijklm}= s(R_j * T_l) + s(R_j *G_k) + T_l + G_k + s(S_i) + \epsilon_{ijklm}\]
 
@@
 \item
-  \(S_i\sim N(0,\sigma^2_S)\) is the random effect of the \(i^{th}\) subject
+  \(s(S_i)\) is the random effect of the \(i^{th}\) subject
 \item
-  \(R_j\) is the effect of the \(j^{th}\) ratio
+  \(s(R_j*T_l)\) is the effect of the \(j^{th}\) ratio, by comparison type
+\item
+  \(s(R_j*G_k)\) is the effect of the \(j^{th}\) ratio, by display type
+\item
+  \(T_l\) is the effect of the \(l^{th}\) comparison type
 \item
   \(G_k\) is the effect of the \(k^{th}\) display type
-\item
-  \(T_l\) is the effect of the \(l^{th}\) comparison type
 \item
@@
 
-\begin{table}[tbp]
+A generalized additive model with random effects for subject was fit using the \texttt{mgcv} package \citep{mgcv1, mgcv2, mgcv3}.
+Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}.
+It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in \citet{hollandsBiasProportionJudgments2000}.
 
-\begin{center}
-\begin{threeparttable}
+While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only 42 trials (8.79\%) had slider values at exactly 0.5, and all but 6 were for ratios on either side of 0.5 (.464, .562).
+This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5.
+In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types.
+It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum.
+A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.
 
-\caption{\label{tab:unnamed-chunk-2}Analysis of Fixed Effects}
+\begin{table}
 
-\begin{tabular}{llll}
+\caption{\label{tab:gam-param}Parametric coefficients in gam model.}
+\centering
+\begin{tabular}[t]{lrrrr}
 \toprule
-Term & \multicolumn{1}{c}{$\hat{\beta}$} & \multicolumn{1}{c}{95\% CI} & \multicolumn{1}{c}{$t$}\\
+  & Estimate & Std. Error & t value & Pr(>|t|)\\
 \midrule
-Intercept & 2.49 & {}[1.94, 3.04] & 8.84\\
-Ratio & 0.02 & {}[-0.04, 0.07] & 0.56\\
-Plot3dDigital & 0.15 & {}[-0.14, 0.43] & 1.02\\
-Plot3dPrint & -0.06 & {}[-0.35, 0.22] & -0.44\\
-Type & 0.11 & {}[-0.13, 0.35] & 0.94\\
+(Intercept) & 2.6862 & 0.1780 & 15.0931 & 0.0000\\
+plot3dDigital & 0.1357 & 0.1409 & 0.9626 & 0.3363\\
+plot3dPrint & -0.0563 & 0.1415 & -0.3980 & 0.6908\\
+typeAdjacent & 0.0968 & 0.1197 & 0.8088 & 0.4191\\
 \bottomrule
 \end{tabular}
-
-\end{threeparttable}
-\end{center}
-
 \end{table}
 
-No differences were detected for the true ratio of bars (\(t = 0.5590574\)), whether the bars were adjacent or separated (\(t = 0.9381629\)), or for the plot nested within the true ratio (\(F(2, 441) = 1.1110313\)).
-While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes.
+\begin{table}
+
+\caption{\label{tab:gam-smooth}Approximate significance of smooth terms in gam model.}
+\centering
+\begin{tabular}[t]{lrrrr}
+\toprule
+Smooth & edf & Ref.df & F & p-value\\
+\midrule
+s(Ratio):plot2dDigital & 0.0005 & 0.0009 & 0.0000 & 0.5000\\
+s(Ratio):plot3dDigital & 1.7934 & 2.2056 & 1.2807 & 0.2481\\
+s(Ratio):plot3dPrint & 1.0003 & 1.0005 & 2.4488 & 0.1183\\
+s(Ratio):typeSeparated & 4.2371 & 4.7338 & 5.8074 & 0.0001\\
+s(Ratio):typeAdjacent & 1.0007 & 1.0014 & 3.5452 & 0.0604\\
+\addlinespace
+s(participant) & 27.0449 & 34.0000 & 4.7264 & 0.0000\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{jds_files/figure-latex/gam-pred-chart-1} \caption{Predictions with standard errors for fixed effects in a generalized additive model with splines for ratio by comparison and ratio by display method.
+It is clear that the separated comparisons are easier to estimate when small, right around 50\%, or large, and harder to estimate between these points.
+What is interesting is that no such trend is present for adjacent comparisons.
+}\label{fig:gam-pred-chart}
+\end{figure}
+
+While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes; if we could get access to their data, this would allow for investigating whether the differences identified in our model were an artifact of the measurement method (e.g.~slider vs.~numerical entry) or if this effect was present in previous studies and can now be identified because of more advanced statistical modeling techniques.
 

### 2024-01-22 jds.rmd

@@
 
+Model diagnostic plots show slight violations of normality with the error term, particularly with right-skewed residuals.
+Plots of the data for each ratio tend to suggest that some participants might have tried rounding to the ratio increments of fives and tens. 
+For example, participants tended to favor estimating the ratio of 0.562 to around 0.5.
+Since linear models are reasonably robust to violations of normality, we will assume that the model reasonably fits.
+Plots of the residual diagnostics can be found under the supplemental material in \Cref{fig:residualplots}.
+
 ## Interactivity with 3D Rendered Charts
@@
 Stimuli, code, and data for this experiment are provided at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).
+
+```{r residualplots, include = F}
+#| fig-width: 3
+#| fig-height: 3
+#| fig-cap: "Residual diagnostic plots for the linear mixed model."
+#| fig-subcap:
+#| - Fitted vs. predicted residuals
+#| - Q-Q plot
+#| out-width: ".49\\linewidth"
+#| fig-show: "hold"
+#| layout-ncol: 2
+
+r = residuals(mod)
+p = predict(mod)
+
+ggplot(mapping = aes(x = p, y = r)) +
+  geom_point() + 
+  geom_hline(yintercept = 0, linetype = 'dashed', color = 'red') + 
+  theme_bw() + 
+  labs(x = 'Fitted value',
+       y = 'Residual')
+
+ggplot(mapping = aes(sample = r)) + 
+  geom_qq() + 
+  geom_qq_line(color = 'red', linetype = 'dashed') + 
+  theme_bw() + 
+  labs(x = 'Theoretical Quantiles',
+       y = 'Sample Quantiles')
+
+```

### 2024-01-22 jds.tex

@@
 As a result of the Cleveland and McGill \citeyearpar{clevelandGraphical1984} study, it is possible to assemble an ordering of perceptual accuracy for the elements of length, position, and angle.
-\citet{heerCrowdsourcingGraphicalPerception2010b} replicated some parts of \citet{clevelandGraphical1984} in an online setting using Mechanical Turk, largely validating the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.
+\citet{heerCrowdsourcingGraphicalPerception2010b} replicated some parts of \citet{clevelandGraphical1984} in an online setting using Mechanical Turk, a platform for crowdsourcing human tasks that has been used for (among other things) assembling machine learning datasets.
+This replication study largely validated the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.
 
@@
 
-Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
-This occasionally causes problems: the line-width illusion\citep{vanderplasSignsSineIllusion2015, daySineIllusion1991, hofmannCommonAnglePlots2013}, for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in \Cref{fig:sine-illusion}.
-While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception.
-Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the stimuli.
+Chart perception is affected by the human visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial.
+In addition, these artificial stimuli are typically created and presented in two dimensions.
+The artificial scenes we create for data visualization have been documented to cause problems: for instance, the line-width illusion \citep{vanderplasSignsSineIllusion2015, daySineIllusion1991, hofmannCommonAnglePlots2013} has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, hammock plots, and Sankey diagrams.
+The line width illusion causes lines or error bands which are actually the same length to appear shorter at points with large amounts of curvature; a simple example of the classic sine illusion (one version of the line width illusion) is shown in \Cref{fig:sine-illusion}.
+While this illusion is problematic in two dimensions, when the depth cues are actually present in a three-dimensional situation, the same implicit visual heuristics contribute to accurate size perception (as in the middle panel of \Cref{fig:sine-illusion}).
+Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the visual display relative to the training of the visual system and the heuristics which are activated in order to make sense of the display.
 
@@
 
-The use of 3D graphics have been explored in multiple studies.
+The use of 3D graphics have been explored in multiple studies, with mixed results.
 \citet{fisher_data_1997} explored user preference for 2D or 3D charts and found that subjects tended to prefer simpler 2D graphs when tasked with extracting information.
@@
 \citet{brath_3d_2014} highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.
+Nevertheless, it is extremely common in visualization literature to recommend that 3D plots be avoided at all costs \citep{vanderplasTestingStatisticalCharts2020, wilkeDonGo3D2020, morganwallIntroducing3DGgplots20191006}.
 
-There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data \citep{vanderplasTestingStatisticalCharts2020, wilkeDonGo3D2020, morganwallIntroducing3DGgplots20191006}.
+There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data \citep{huffHowLieStatistics1954}.
 However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study \citeyearpar{heerCrowdsourcingGraphicalPerception2010b} in 2010.\\
-Digital graphics have developed quickly, along with the hardware necessary to support these software developments.
-As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
-As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision \citep{fleetHowTactileGraphics2023}.
+Computer graphics and rendering technology have developed quickly, along with the hardware necessary to support these software developments.
+As a result, we have much more natural virtual rendering of 3D objects, and we can also print graphics in three dimensions using mass-market, relatively inexpensive 3D printers, moving artificial charts into a more natural, physical setting.
+As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts provide the opportunity to make data graphics accessible to those with limited or absent vision \citep{fleetHowTactileGraphics2023}.
 
@@
 
+Model diagnostic plots show slight violations of normality with the error term, particularly with right-skewed residuals.
+Plots of the data for each ratio tend to suggest that some participants might have tried rounding to the ratio increments of fives and tens.
+For example, participants tended to favor estimating the ratio of 0.562 to around 0.5.
+Since linear models are reasonably robust to violations of normality, we will assume that the model reasonably fits.
+Plots of the residual diagnostics can be found under the supplemental material in \Cref{fig:residualplots}.
+
 \hypertarget{interactivity-with-3d-rendered-charts}{%

### 2024-01-22 jds.rmd
diff for pdf

@@
 
-```{r demographics, fig.cap="Demographic characteristics of participants in the study.", message=FALSE, warning=FALSE, out.width='100%', fig.width = 9, fig.height = 3.5}
+```{r demographics, fig.cap="Demographic characteristics of participants in the study.", message=FALSE, warning=FALSE, out.width='100%', fig.width = 9, fig.height = 3.5, cache = F}
 library(patchwork)

### 2024-01-22 jds.rmd
Fixed merge issue

@@
 
-con = dbConnect(SQLite(), '../2022-Tyler-3D_Graphical_Perception/experiment_interface/department.db')
+con = dbConnect(SQLite(), '_data/department.db')
 dbListTables(con)
@@
 
-load('../2022-Tyler-3D_Graphical_Perception/experiment_interface/data/set85data.Rdata')
-load('../2022-Tyler-3D_Graphical_Perception/experiment_interface/data/kits.Rdata')
+load('_data/set85data.Rdata')
+load('_data/kits.Rdata')
 
@@
 ## 3D Graphical Perception
+
 Chart perception is affected by the human visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial.
@@
 The exact numeric comparisons were not disclosed, but the comparison values used in our study were subjected to the constraints of having the same ratio values and that no value was used more than twice.
-\svp{Can we show the comparison values in a table?}
 
-\svp{In each graph, there are two sets of five bars each, and two of the bars are marked for identification with a circle and triangle}, \svp{as in \Cref{fig:bar-types}.}
-Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.\svp{What does excessive noise mean??}
+```{r experiment-values}
+library(papaja)
+extract_values <- function(data){
+  data %>% 
+    dplyr::filter(!is.na(IDchr)) %>% 
+    dplyr::select(Height)
+}
+
+datasets %>% 
+  mutate(setID = str_extract(file, 'id-[0-9]{2}')) %>% 
+  filter(fileID < 15) %>% 
+  mutate(vals = map(data, extract_values)) %>% 
+  unnest(vals) %>% 
+  mutate(Height = round(Height,1)) %>% 
+  group_by(fileID) %>% 
+  mutate(Bar = ifelse(Height == min(Height), 'Smaller', 'Larger')) %>% 
+  ungroup() %>% 
+  select(fileID, Height, Bar, setID) %>% 
+  filter(fileID %% 2 == 1) %>% 
+  mutate(setID = paste0('ID: ', str_remove(setID, 'id-'))) %>% 
+  select(-fileID) %>% 
+  pivot_wider(names_from = Bar, values_from = Height) %>% 
+  mutate(`Ratio (%)` = round(Smaller/Larger,3)*100) %>% arrange(`Ratio (%)`) %>% 
+  pivot_longer(Smaller:`Ratio (%)`, values_to = 'Height', names_to = 'Bar') %>% 
+  pivot_wider(names_from = setID, values_from = Height) %>% 
+  apa_table(caption = 'Values used in the experiment to make ratio comparisons, sorted by ratio. The ID label corresponds to the file reference number.')
+
+```
+
+In each graph, there are two sets of five bars each, and two of the bars are marked for identification with a circle and triangle, as in \Cref{fig:bar-types}.
+Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution to provide random structure around our fixed values.
 Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).
@@
 \Cref{fig:midmeans-log-errors} shows the midmeans of the log absolute errors compared to the true ratio of the bars for each graph type and comparison type. 
-The results suggest that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types. In addition, there is a tendency
+The results suggest that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types.
 
@@
             data = res)
+
 modsum <- summary(mod)
@@
 #| - Q-Q plot
-#| out-width: ".49\\linewidth"
+#| out-width: ".40\\linewidth"
 #| fig-show: "hold"

### 2024-01-22 jds.rmd
Corrected graph axis label

@@
   ggplot(aes(x = plot3dClicks, y = response)) + geom_point() + 
-  xlab("# Clicks on WebGL Plot") + ylab("Count") + ggtitle("Accuracy and Participant Interactivity") + 
+  xlab("# Clicks on WebGL Plot") + ylab("Log Error") + ggtitle("Accuracy and Participant Interactivity") + 
   geom_smooth(method = "lm", se = T) + 

### 2024-01-22 jds.rmd
Removed residual plots

@@
 
-Model diagnostic plots show slight violations of normality with the error term, particularly with right-skewed residuals.
+<!---Model diagnostic plots show slight violations of normality with the error term, particularly with right-skewed residuals.
 Plots of the data for each ratio tend to suggest that some participants might have tried rounding to the ratio increments of fives and tens. 
@@
 Since linear models are reasonably robust to violations of normality, we will assume that the model reasonably fits.
-Plots of the residual diagnostics can be found under the supplemental material in \Cref{fig:residualplots}.
+Plots of the residual diagnostics can be found under the supplemental material in \Cref{fig:residualplots}.--->
 
@@
 ```
+
+```{r residualplots-gam, include = F}
+#| fig-width: 3
+#| fig-height: 3
+#| fig-cap: "Residual diagnostic plots for the linear mixed model."
+#| fig-subcap:
+#| - Fitted vs. predicted residuals
+#| - Q-Q plot
+#| out-width: ".40\\linewidth"
+#| fig-show: "hold"
+#| layout-ncol: 2
+
+r = residuals(gam_mod)
+p = predict(gam_mod)
+
+ggplot(mapping = aes(x = p, y = r)) +
+  geom_point() + 
+  geom_hline(yintercept = 0, linetype = 'dashed', color = 'red') + 
+  theme_bw() + 
+  labs(x = 'Fitted value',
+       y = 'Residual')
+
+ggplot(mapping = aes(sample = r)) + 
+  geom_qq() + 
+  geom_qq_line(color = 'red', linetype = 'dashed') + 
+  theme_bw() + 
+  labs(x = 'Theoretical Quantiles',
+       y = 'Sample Quantiles')
+
+```

### 2024-01-22 jds.tex
Removed residual plots

@@
 The exact numeric comparisons were not disclosed, but the comparison values used in our study were subjected to the constraints of having the same ratio values and that no value was used more than twice.
-{\textcolor{ForestGreen}{Can we show the comparison values in a table?}}
 
-{\textcolor{ForestGreen}{In each graph, there are two sets of five bars each, and two of the bars are marked for identification with a circle and triangle}}, {\textcolor{ForestGreen}{as in \Cref{fig:bar-types}.}}
-Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.{\textcolor{ForestGreen}{What does excessive noise mean??}}
+\begin{table}[tbp]
+
+\begin{center}
+\begin{threeparttable}
+
+\caption{\label{tab:experiment-values}Values used in the experiment to make ratio comparisons, sorted by ratio. The ID label corresponds to the file reference number.}
+
+\begin{tabular}{llllllll}
+\toprule
+Bar & \multicolumn{1}{c}{ID: 01} & \multicolumn{1}{c}{ID: 06} & \multicolumn{1}{c}{ID: 05} & \multicolumn{1}{c}{ID: 04} & \multicolumn{1}{c}{ID: 02} & \multicolumn{1}{c}{ID: 03} & \multicolumn{1}{c}{ID: 09}\\
+\midrule
+Smaller & 10.00 & 14.70 & 14.70 & 12.10 & 10.00 & 12.10 & 26.10\\
+Larger & 56.20 & 56.20 & 38.30 & 26.10 & 17.80 & 17.80 & 31.60\\
+Ratio (\%) & 17.80 & 26.20 & 38.40 & 46.40 & 56.20 & 68.00 & 82.60\\
+\bottomrule
+\end{tabular}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
+
+In each graph, there are two sets of five bars each, and two of the bars are marked for identification with a circle and triangle, as in \Cref{fig:bar-types}.
+Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution to provide random structure around our fixed values.
 Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at \url{https://github.com/TWiedRW/2023-JDS-3dcharts}.
@@
 \Cref{fig:midmeans-log-errors} shows the midmeans of the log absolute errors compared to the true ratio of the bars for each graph type and comparison type.
-The results suggest that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types. In addition, there is a tendency
+The results suggest that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types.
 
@@
 \addlinespace
-s(participant) & 27.0449 & 34.0000 & 4.7264 & 0.0000\\
+s(participant) & 27.0449 & 34.0000 & 4.7276 & 0.0000\\
 \bottomrule
@@
 While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes; if we could get access to their data, this would allow for investigating whether the differences identified in our model were an artifact of the measurement method (e.g.~slider vs.~numerical entry) or if this effect was present in previous studies and can now be identified because of more advanced statistical modeling techniques.
-
-Model diagnostic plots show slight violations of normality with the error term, particularly with right-skewed residuals.
-Plots of the data for each ratio tend to suggest that some participants might have tried rounding to the ratio increments of fives and tens.
-For example, participants tended to favor estimating the ratio of 0.562 to around 0.5.
-Since linear models are reasonably robust to violations of normality, we will assume that the model reasonably fits.
-Plots of the residual diagnostics can be found under the supplemental material in \Cref{fig:residualplots}.
 

### 2024-01-22 jds.rmd
Updates

@@
   filter(whichIsSmaller == 'Triangle (▲)',
-         as.Date.POSIXct(appStartTime) <= '2023-05-22') #Only verified results from before the start of SDSS
+         as.Date.POSIXct(appStartTime) <= '2023-05-22') %>% 
+  #Only verified results from before the start of SDSS 
+  filter(!is.na(ratio))
 
@@
 All responses ($n = `r nrow(res_all) - nrow(res)`$) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check.
+In addition, one trial was removed because the participant did not enter a valid ID for a 3D printed chart.
 
@@
 
-In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a generalized additive model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio.
-The formal statistical model is as follows:
-
-$$y_{ijklm}= s(R_j * T_l) + s(R_j *G_k) + T_l + G_k + s(S_i) + \epsilon_{ijklm}$$
+In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a generalized additive model [@hastie2017generalized] that accounts for participant variation as well as the effect of comparison type, graph type, and ratio.
+
+We define the following effects:
+
+- $R$, the numerical ratio of A/B where A is the length of the smaller bar and B is the length of the larger bar.
+There are $i=1, ..., 7$ levels of $R$ in this study, but we will treat $R$ as a numeric variable throughout this model.
+
+- $D_j$, the display mode, where $j = 1, 2, 3$ correspond to 2D, 3D rendered, and 3D printed charts respectively.
+
+- $T_k$, the type of comparison, where $k = 1, 2$ correspond to adjacent (Type 1) and separated (Type 3) comparisons, respectively
+
+- $P_l$, a random effect for participant which describes the overall skill of the participant at visual estimation of ratios.
+
+Generalized additive models depend on a smooth function $s(): \mathbb{R}\rightarrow\mathbb{R}$ which is applied to the independent variables; by default, the smooth function is created using thin-plate regression splines.
+In place of an interaction term when a categorical variable is involved, separate smooths are fit for each level of the categorical variable; we will indicate this by subscripting the smooth function with the categorical variable index, so that $s^T_1(R)$ indicates the smooth over numerical variable $R$ for the 1st level of the categorical variable T, that is, the smooth function over $R$ for adjacent comparisons.
+
+Then the formal statistical model is as follows:
+
+$$y= s^D_j(R) + s^T_k(R) + D_j + T_k + P_l + \epsilon$$
 
@@
 
-- $y_{ijklm}=\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$
-
-- $s(S_i)$ is the random effect of the $i^{th}$ subject
-
-- $s(R_j*T_l)$ is the effect of the $j^{th}$ ratio, by comparison type
-
-- $s(R_j*G_k)$ is the effect of the $j^{th}$ ratio, by display type
-
-- $T_l$ is the effect of the $l^{th}$ comparison type
-
-- $G_k$ is the effect of the $k^{th}$ display type
-
-- $\epsilon_{ijklm}\sim N(0,\sigma^2_\epsilon)$ is the random error
-
-```{r, include=F}
-library(papaja)
-library(lme4)
-
-mod <- lmer(response ~ (1|subject) + ratioLabel + plot + type,
-            data = res)
-
-modsum <- summary(mod)
-modaov <- anova(mod)
-modaovsum <- summary(modaov)
-
-f_df <- as.numeric(length(modsum$residuals)-modsum$ngrps - 1)
-
-apa_lm <- apa_print(mod)
-apa_table(apa_lm$table, caption = "Analysis of Fixed Effects")
-```
-
-<!-- No differences were detected for the true ratio of bars ($t = `r modsum$coefficients[2,3]`$), whether the bars were adjacent or separated ($t = `r modsum$coefficients[5,3]`$), or for the plot nested within the true ratio ($F(`r modaov$npar[2]`, `r f_df`) = `r modaov[2,4]`$).
--->
-<!-- The form of this linear mixed effects model assumes a linear relationship between ratio and response, but we know from \Cref{fig:midmeans-log-errors} that the actual relationship appears to be nonlinear.
--->
-A generalized additive model with random effects for subject was fit using the `mgcv` package [@mgcv1;@mgcv2;@mgcv3].
-Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}.
-It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in @hollandsBiasProportionJudgments2000.
-
-While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only `r sum(res$byHowMuch == 50)` trials (`r sprintf("%.2f%%", mean(res$byHowMuch==50)*100)`) had slider values at exactly 0.5, and all but `r sum(res$byHowMuch == 50 & !res$ratioLabel %in% c(46.4, 56.2))` were for ratios on either side of 0.5 (.464, .562).
-This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5.
-In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types.
-It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum.
-A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points  could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.
+- $y=\log_2(|\text{Judged Percent}_{ijkl} - \text{True Percent}_{ij}|+1/8)$
+
+- $s^D_j(R)$ is a thin-plate spline function for R accounting for the display type
+
+- $s^T_k(R)$ is a thin-plate spline function for R accounting for the comparison type
+
+- $D_j$ is the fixed effect of the display type (e.g. an intercept term)
+
+- $T_k$ is the fixed effect of the $k^{th}$ comparison type
+
+- $P_l$ is the random effect of the $l^{th}$ participant
+
+- $\epsilon$ is the random error
 
@@
 
-gam_mod <- gam(response ~ s(ratio_prop, by = plot, k = 6) + s(ratio_prop, by = type, k = 6) +  plot + type + s(participant, bs = "re"),
+gam_mod <- gam(response ~ 
+                 s(ratio_prop, by = plot, k = 6) +
+                 s(ratio_prop, by = type, k = 6) + 
+                 plot + type + 
+                 s(participant, bs = "re"),
                data = res2, method = 'REML')
@@
 ```
+
+A generalized additive model with random effects for subject was fit using the `mgcv` package [@mgcv1;@mgcv2;@mgcv3].
+Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}.
+It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in @hollandsBiasProportionJudgments2000.
+Residuals from the model indicate that there is some left skew in the tails, but this is not extreme; code for this analysis can be found in the supplemental materials on github.
 
@@
   xlim(c(0, 1))
-
-```
-
-While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes; if we could get access to their data, this would allow for investigating whether the differences identified in our model were an artifact of the measurement method (e.g.
-slider vs.
-numerical entry) or if this effect was present in previous studies and can now be identified because of more advanced statistical modeling techniques.
-
-<!---Model diagnostic plots show slight violations of normality with the error term, particularly with right-skewed residuals.
-Plots of the data for each ratio tend to suggest that some participants might have tried rounding to the ratio increments of fives and tens. 
-For example, participants tended to favor estimating the ratio of 0.562 to around 0.5.
-Since linear models are reasonably robust to violations of normality, we will assume that the model reasonably fits.
-Plots of the residual diagnostics can be found under the supplemental material in \Cref{fig:residualplots}.--->
-
-## Interactivity with 3D Rendered Charts
-
-When 3D rendered plots were shown, we recorded the number of user interactions (clicks, rotations) with the WebGL plot.
-\Cref{fig:clicks} shows the distribution of number of clicks over all 3D rendered trials in the experiment, as well as the number of clicks relative to the accuracy measure $\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$.
-It does not appear that participants who interacted with the WebGL rendering were more accurate than their peers who did not.
-
-```{r clicks, fig.cap = "Participant clicks on the WebGL interface (left) in order to interact with the plot.
-Most participants did not interact with the WebGL interface, perhaps because they did not realize that interacting was an option, but some participants utilized the interactivity heavily.
-(Right) Participant interactivity was not associated with increased accuracy.", fig.width = 4, fig.height = 4, out.width = ".45\\textwidth", fig.show = "hold"}
-res %>%
-  filter(plot == "3dDigital") %>%
-  ggplot(aes(x = plot3dClicks)) + geom_bar() + 
-  xlab("# Clicks on WebGL Plot") + ylab("Count") + ggtitle("Distribution of WebGL Interaction") + 
-  theme_bw()
-
-res %>%
-  filter(plot == "3dDigital") %>%
-  ggplot(aes(x = plot3dClicks, y = response)) + geom_point() + 
-  xlab("# Clicks on WebGL Plot") + ylab("Log Error") + ggtitle("Accuracy and Participant Interactivity") + 
-  geom_smooth(method = "lm", se = T) + 
-  theme_bw()
-```
-
-# Discussion and Future Work
-
-Previous work in 3D graphics would suggest that the errors for the 3D graphs would be larger than the errors for the 2D graphs. 
-While we did not find any significant results indicating that 3D graphs are read less accurately, there are two possibilities that might account for this discrepancy.
-
-The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger.
-However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @clevelandGraphical1984 do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.
-
-The second possibility is more interesting: we examined 3D charts using rendered 3D graphs and 3D printed charts; both of these options allow for participants to interact with the chart, rotating it, and generally perceiving it as one might perceive any other 3D, real, object.
-This is a far cry from the 3D perspective charts in the original study, which have a fixed angle and perspective and are thus not equivalent to our 3D charts. 
-Future studies should include an additional fixed 3D perspective bar chart, which will at least enable us to examine whether modern 3D rendering environments allow for more accurate conclusions than fixed 3D perspectives.
-Future iterations of this study will include "traditional" 3D graphs created by Microsoft Excel (that is, graphs with a fixed 3D perspective rendered in 2D).
-This option will allow us to examine fixed perspective 3D plots compared to 3D renderings and 2D plots; it will also enable online data collection in addition to the in-person data collection used in this experiment.
-
-Another interesting aspect of our study is that the method used to record participant estimates is different from the method used in the original study as well as Heer \& Bostock's replication study. 
-A method similar to our slider input, marking position on a line, was used in @spenceVisualPsychophysicsSimple1990, but Spence asked participants to estimate $A/(A+B)$, where we asked participants to estimate $A/B$; thus, our results are still not directly comparable to previous studies.
-We expect that the specific ratio estimated would also have an effect on observed participant errors.
-
-The slider method for input of ratio estimates should be easier for participants, as it does not require explicit transformation to the numerical domain. 
-What is clear is that it would be beneficial to assess the impact of the measurement method on participant errors directly, so that the results of these different studies might be explained and interpreted with regard both to the stimuli used and the measurement method employed in the experiment.
-Future iterations of this experiment will likely address this estimation difference; such modifications in experimental design are relatively straightforward in Shiny and will provide useful insight into the design of future experiments evaluating the perception of statistical graphics.
-
-<!-- The goal of this study was to verify that previous research into graphics is upheld under a new set of circumstances.
-It started with a question, followed by research into prior work, and then by designing and conducting a study.
-This general framework of research is useful to all fields of study and is taught in most introductory statistics courses.
-The next step of this study is to collect more data while simultaneously providing students the opportunity to learn about the research process from both the perspectives of an experiment participant and the views of the researchers.
-At University of Nebraska-Lincoln, students taking introductory statistics will have the opportunity to participate in the experiment while also completing pre/post experiment surveys and reflections on a two-page abstract and a presentation of results.
--->
-
-## Supplemental Material
-Stimuli, code, and data for this experiment are provided at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).
-
-```{r residualplots, include = F}
-#| fig-width: 3
-#| fig-height: 3
-#| fig-cap: "Residual diagnostic plots for the linear mixed model."
-#| fig-subcap:
-#| - Fitted vs. predicted residuals
-#| - Q-Q plot
-#| out-width: ".40\\linewidth"
-#| fig-show: "hold"
-#| layout-ncol: 2
-
-r = residuals(mod)
-p = predict(mod)
-
-ggplot(mapping = aes(x = p, y = r)) +
-  geom_point() + 
-  geom_hline(yintercept = 0, linetype = 'dashed', color = 'red') + 
-  theme_bw() + 
-  labs(x = 'Fitted value',
-       y = 'Residual')
-
-ggplot(mapping = aes(sample = r)) + 
-  geom_qq() + 
-  geom_qq_line(color = 'red', linetype = 'dashed') + 
-  theme_bw() + 
-  labs(x = 'Theoretical Quantiles',
-       y = 'Sample Quantiles')
 
@@
 
-r = residuals(gam_mod)
-p = predict(gam_mod)
+res2 <- res2 %>%
+  mutate(predict = predict(gam_mod, newdata = res2),
+         resid = response - predict)
+
+ggplot(res2, aes(x = predict, y = resid)) +
+  geom_point() + 
+  geom_hline(yintercept = 0, linetype = 'dashed', color = 'red') + 
+  theme_bw() + 
+  facet_wrap(~plottype,) + 
+  labs(x = 'Fitted value',
+       y = 'Residual')
+
+ggplot(res2, aes(sample = resid)) + 
+  geom_qq() + 
+  geom_qq_line(color = 'red', linetype = 'dashed') + 
+  facet_grid(type ~ plot) + 
+  theme_bw() + 
+  labs(x = 'Theoretical Quantiles',
+       y = 'Sample Quantiles')
+
+```
+
+While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only `r sum(res$byHowMuch == 50)` trials (`r sprintf("%.2f%%", mean(res$byHowMuch==50)*100)`) had slider values at exactly 0.5, and all but `r sum(res$byHowMuch == 50 & !res$ratioLabel %in% c(46.4, 56.2))` were for ratios on either side of 0.5 (.464, .562).
+This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5.
+In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types.
+It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum.
+A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points  could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.
+
+\citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, but this link has not been updated and is no longer active, so it is not possible to fit a similar model to their data for comparison purposes; if we could get access to their data, this would allow for investigating whether the differences identified in our model were an artifact of the measurement method (e.g.
+slider vs.
+numerical entry) or if this effect was present in previous studies and can now be identified because of more advanced statistical modeling techniques.
+
+<!---Model diagnostic plots show slight violations of normality with the error term, particularly with right-skewed residuals.
+Plots of the data for each ratio tend to suggest that some participants might have tried rounding to the ratio increments of fives and tens. 
+For example, participants tended to favor estimating the ratio of 0.562 to around 0.5.
+Since linear models are reasonably robust to violations of normality, we will assume that the model reasonably fits.
+Plots of the residual diagnostics can be found under the supplemental material in \Cref{fig:residualplots}.--->
+
+## Interactivity with 3D Rendered Charts
+
+When 3D rendered plots were shown, we recorded the number of user interactions (clicks, rotations) with the WebGL plot.
+\Cref{fig:clicks} shows the distribution of number of clicks over all 3D rendered trials in the experiment, as well as the number of clicks relative to the accuracy measure $\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$.
+It does not appear that participants who interacted with the WebGL rendering were more accurate than their peers who did not.
+
+```{r clicks, fig.cap = "Participant clicks on the WebGL interface (left) in order to interact with the plot.
+Most participants did not interact with the WebGL interface, perhaps because they did not realize that interacting was an option, but some participants utilized the interactivity heavily.
+(Right) Participant interactivity was not associated with increased accuracy.", fig.width = 4, fig.height = 4, out.width = ".45\\textwidth", fig.show = "hold"}
+res %>%
+  filter(plot == "3dDigital") %>%
+  ggplot(aes(x = plot3dClicks)) + geom_bar() + 
+  xlab("# Clicks on WebGL Plot") + ylab("Count") + ggtitle("Distribution of WebGL Interaction") + 
+  theme_bw()
+
+res %>%
+  filter(plot == "3dDigital") %>%
+  ggplot(aes(x = plot3dClicks, y = response)) + geom_point() + 
+  xlab("# Clicks on WebGL Plot") + ylab("Log Error") + ggtitle("Accuracy and Participant Interactivity") + 
+  geom_smooth(method = "lm", se = T) + 
+  theme_bw()
+```
+
+# Discussion and Future Work
+
+Previous work in 3D graphics would suggest that the errors for the 3D graphs would be larger than the errors for the 2D graphs. 
+While we did not find any significant results indicating that 3D graphs are read less accurately, there are two possibilities that might account for this discrepancy.
+
+The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger.
+However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @clevelandGraphical1984 do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.
+
+The second possibility is more interesting: we examined 3D charts using rendered 3D graphs and 3D printed charts; both of these options allow for participants to interact with the chart, rotating it, and generally perceiving it as one might perceive any other 3D, real, object.
+This is a far cry from the 3D perspective charts in the original study, which have a fixed angle and perspective and are thus not equivalent to our 3D charts. 
+Future studies should include an additional fixed 3D perspective bar chart, which will at least enable us to examine whether modern 3D rendering environments allow for more accurate conclusions than fixed 3D perspectives.
+Future iterations of this study will include "traditional" 3D graphs created by Microsoft Excel (that is, graphs with a fixed 3D perspective rendered in 2D).
+This option will allow us to examine fixed perspective 3D plots compared to 3D renderings and 2D plots; it will also enable online data collection in addition to the in-person data collection used in this experiment.
+
+Another interesting aspect of our study is that the method used to record participant estimates is different from the method used in the original study as well as Heer \& Bostock's replication study. 
+A method similar to our slider input, marking position on a line, was used in @spenceVisualPsychophysicsSimple1990, but Spence asked participants to estimate $A/(A+B)$, where we asked participants to estimate $A/B$; thus, our results are still not directly comparable to previous studies.
+We expect that the specific ratio estimated would also have an effect on observed participant errors.
+
+The slider method for input of ratio estimates should be easier for participants, as it does not require explicit transformation to the numerical domain. 
+What is clear is that it would be beneficial to assess the impact of the measurement method on participant errors directly, so that the results of these different studies might be explained and interpreted with regard both to the stimuli used and the measurement method employed in the experiment.
+Future iterations of this experiment will likely address this estimation difference; such modifications in experimental design are relatively straightforward in Shiny and will provide useful insight into the design of future experiments evaluating the perception of statistical graphics.
+
+<!-- The goal of this study was to verify that previous research into graphics is upheld under a new set of circumstances.
+It started with a question, followed by research into prior work, and then by designing and conducting a study.
+This general framework of research is useful to all fields of study and is taught in most introductory statistics courses.
+The next step of this study is to collect more data while simultaneously providing students the opportunity to learn about the research process from both the perspectives of an experiment participant and the views of the researchers.
+At University of Nebraska-Lincoln, students taking introductory statistics will have the opportunity to participate in the experiment while also completing pre/post experiment surveys and reflections on a two-page abstract and a presentation of results.
+-->
+
+## Supplemental Material
+Stimuli, code, and data for this experiment are provided at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).
+
+```{r, include=F, eval = F}
+library(papaja)
+library(lme4)
+
+mod <- lmer(response ~ (1|subject) + ratioLabel + plot + type,
+            data = res)
+
+modsum <- summary(mod)
+modaov <- anova(mod)
+modaovsum <- summary(modaov)
+
+f_df <- as.numeric(length(modsum$residuals)-modsum$ngrps - 1)
+
+apa_lm <- apa_print(mod)
+apa_table(apa_lm$table, caption = "Analysis of Fixed Effects")
+```
+
+```{r residualplots, include = F, eval = F}
+#| fig-width: 3
+#| fig-height: 3
+#| fig-cap: "Residual diagnostic plots for the linear mixed model."
+#| fig-subcap:
+#| - Fitted vs. predicted residuals
+#| - Q-Q plot
+#| out-width: ".40\\linewidth"
+#| fig-show: "hold"
+#| layout-ncol: 2
+
+r = residuals(mod)
+p = predict(mod)
 
@@
 ```
+
+```{r percent-error-model, eval = F}
+# Model using % error instead of the log2 error...
+
+res3 <- res2 %>%
+  mutate(estimation_error = (byHowMuch - ratioLabel)/(ratioLabel))
+res3 %>% 
+  mutate(type = "Raw Data") %>%
+  ggplot(mapping = aes(x = Height*100, y = estimation_error,
+                       color = graphtype, fill = graphtype)) +
+  geom_smooth(se = T, alpha=.125,  expand = F, method = "loess") +
+  geom_point(aes(shape = type, alpha = type, size = type)) +
+  scale_shape_manual("Data", values = c(19, 1)) + 
+  scale_size_manual("Data", values = c(2, 1)) + 
+  scale_alpha_manual("Data", values = c(1, .25)) + 
+  scale_fill_discrete(labels = c('Adjacent', 'Separated')) +
+  scale_color_discrete(labels = c('Adjacent', 'Separated')) +
+  scale_x_continuous(limits = c(0, 100)) + 
+  facet_wrap(~plot, labeller = labeller(plot = plot.types)) +
+  labs(title = '',
+       x = 'True Proportional Difference (%)',
+       y = 'Log Error',fill = 'Comparison Type',
+       color = 'Comparison Type') +
+  theme_bw() +
+  theme(legend.position = 'bottom')
+gam_mod2 <- gam(estimation_error ~ 
+                 s(ratio_prop, by = plot, k = 6) +
+                 s(ratio_prop, by = type, k = 6) + 
+                 plot + type + 
+                 s(participant, bs = "re"),
+               data = res3, method = 'REML')
+# gam_mod
+
+gam_sum2 <- summary(gam_mod2)
+
+gam_sum2$p.table %>% knitr::kable(digits = 4, caption = "Parametric coefficients in gam model.", label = "gam-param", booktabs = T)
+
+gam_sum2$s.table %>% as.data.frame() %>% 
+  mutate(Smooth = rownames(.),
+         Smooth = str_replace(Smooth, "ratio_prop", "Ratio")) %>%
+  select(Smooth, everything()) %>%
+  knitr::kable(digits = 4, row.names = F, caption = "Approximate significance of smooth terms in gam model.", label = "gam-smooth", booktabs = T)
+
+newdata <- expand.grid(
+  ratio_prop = seq(0.15, .85, .001), 
+  type = c("Separated", "Adjacent"), 
+  plot = unique(res3$plot), 
+  participant = 1)
+pred2 <- predict(gam_mod2, newdata = newdata, type = "terms", se.fit = T) %>% as.data.frame()
+
+pred_se2 <- select(pred2, matches("se.fit")) %>%
+  select(-matches("participant"))
+pred_fit2 <- select(pred2, matches("^fit.")) %>%
+  select(-matches("participant"))
+
+newdata %>%
+  mutate(pred = rowSums(pred_fit2) + gam_mod2$coefficients[1], 
+         se = rowSums(pred_se2) + gam_sum2$p.table[1,2]) %>%
+  ggplot(aes(x = ratio_prop, y = pred , color = type)) + 
+  geom_ribbon(aes(ymin = pred - 1.96*se, ymax = pred + 1.96*se, fill = type), alpha = .25) +  
+  geom_line() + facet_grid(.~plot) + 
+  scale_color_discrete("Comparison", na.translate = F) + scale_fill_discrete("Comparison", na.translate = F) +
+
+  geom_point(data = res3, aes(x = ratioLabel/100, y = estimation_error, color = type)) +
+  theme_bw() + ylab("Estimation Error") + xlab("True Proportion Difference") + 
+  xlim(c(0, 1))
+
+r2 = residuals(gam_mod2)
+p2 = predict(gam_mod2)
+
+ggplot(mapping = aes(x = p2, y = r2)) +
+  geom_point() + 
+  geom_hline(yintercept = 0, linetype = 'dashed', color = 'red') + 
+  theme_bw() + 
+  labs(x = 'Fitted value',
+       y = 'Residual')
+
+ggplot(mapping = aes(sample = r2)) + 
+  geom_qq() + 
+  geom_qq_line(color = 'red', linetype = 'dashed') + 
+  theme_bw() + 
+  labs(x = 'Theoretical Quantiles',
+       y = 'Sample Quantiles')
+```

### 2024-01-22 jds.tex
Updates

@@
 
-All responses (\(n = 32\)) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check.
+All responses (\(n = 33\)) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check.
+In addition, one trial was removed because the participant did not enter a valid ID for a 3D printed chart.
 
@@
 
-In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a generalized additive model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio.
-The formal statistical model is as follows:
+In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a generalized additive model \citep{hastie2017generalized} that accounts for participant variation as well as the effect of comparison type, graph type, and ratio.
 
-\[y_{ijklm}= s(R_j * T_l) + s(R_j *G_k) + T_l + G_k + s(S_i) + \epsilon_{ijklm}\]
+We define the following effects:
+
+\begin{itemize}
+\item
+  \(R\), the numerical ratio of A/B where A is the length of the smaller bar and B is the length of the larger bar.
+There are \(i=1, ..., 7\) levels of \(R\) in this study, but we will treat \(R\) as a numeric variable throughout this model.
+\item
+  \(D_j\), the display mode, where \(j = 1, 2, 3\) correspond to 2D, 3D rendered, and 3D printed charts respectively.
+\item
+  \(T_k\), the type of comparison, where \(k = 1, 2\) correspond to adjacent (Type 1) and separated (Type 3) comparisons, respectively
+\item
+  \(P_l\), a random effect for participant which describes the overall skill of the participant at visual estimation of ratios.
+\end{itemize}
+
+Generalized additive models depend on a smooth function \(s(): \mathbb{R}\rightarrow\mathbb{R}\) which is applied to the independent variables; by default, the smooth function is created using thin-plate regression splines.
+In place of an interaction term when a categorical variable is involved, separate smooths are fit for each level of the categorical variable; we will indicate this by subscripting the smooth function with the categorical variable index, so that \(s^T_1(R)\) indicates the smooth over numerical variable \(R\) for the 1st level of the categorical variable T, that is, the smooth function over \(R\) for adjacent comparisons.
+
+Then the formal statistical model is as follows:
+
+\[y= s^D_j(R) + s^T_k(R) + D_j + T_k + P_l + \epsilon\]
 
@@
 \item
-  \(y_{ijklm}=\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)\)
+  \(y=\log_2(|\text{Judged Percent}_{ijkl} - \text{True Percent}_{ij}|+1/8)\)
 \item
-  \(s(S_i)\) is the random effect of the \(i^{th}\) subject
+  \(s^D_j(R)\) is a thin-plate spline function for R accounting for the display type
 \item
-  \(s(R_j*T_l)\) is the effect of the \(j^{th}\) ratio, by comparison type
+  \(s^T_k(R)\) is a thin-plate spline function for R accounting for the comparison type
 \item
-  \(s(R_j*G_k)\) is the effect of the \(j^{th}\) ratio, by display type
+  \(D_j\) is the fixed effect of the display type (e.g.~an intercept term)
 \item
-  \(T_l\) is the effect of the \(l^{th}\) comparison type
+  \(T_k\) is the fixed effect of the \(k^{th}\) comparison type
 \item
-  \(G_k\) is the effect of the \(k^{th}\) display type
+  \(P_l\) is the random effect of the \(l^{th}\) participant
 \item
-  \(\epsilon_{ijklm}\sim N(0,\sigma^2_\epsilon)\) is the random error
+  \(\epsilon\) is the random error
 \end{itemize}
-
-A generalized additive model with random effects for subject was fit using the \texttt{mgcv} package \citep{mgcv1, mgcv2, mgcv3}.
-Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}.
-It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in \citet{hollandsBiasProportionJudgments2000}.
-
-While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only 42 trials (8.79\%) had slider values at exactly 0.5, and all but 6 were for ratios on either side of 0.5 (.464, .562).
-This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5.
-In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types.
-It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum.
-A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.
 
@@
 \addlinespace
-s(participant) & 27.0449 & 34.0000 & 4.7276 & 0.0000\\
+s(participant) & 27.0449 & 34.0000 & 4.7199 & 0.0000\\
 \bottomrule
@@
 \end{table}
+
+A generalized additive model with random effects for subject was fit using the \texttt{mgcv} package \citep{mgcv1, mgcv2, mgcv3}.
+Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}.
+It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in \citet{hollandsBiasProportionJudgments2000}.
+Residuals from the model indicate that there is some left skew in the tails, but this is not extreme; code for this analysis can be found in the supplemental materials on github.
 
@@
 
-While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes; if we could get access to their data, this would allow for investigating whether the differences identified in our model were an artifact of the measurement method (e.g.~slider vs.~numerical entry) or if this effect was present in previous studies and can now be identified because of more advanced statistical modeling techniques.
+While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only 42 trials (8.81\%) had slider values at exactly 0.5, and all but 6 were for ratios on either side of 0.5 (.464, .562).
+This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5.
+In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types.
+It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum.
+A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.
+
+\citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, but this link has not been updated and is no longer active, so it is not possible to fit a similar model to their data for comparison purposes; if we could get access to their data, this would allow for investigating whether the differences identified in our model were an artifact of the measurement method (e.g.~slider vs.~numerical entry) or if this effect was present in previous studies and can now be identified because of more advanced statistical modeling techniques.
 

### 2024-03-21 jds.rmd
Updated paper based on reviewer feedback

@@
          ratioLabel = round(100*Height, 1)) %>% 
-  filter(subject %in% valid.users$subject) %>% 
+  filter(subject %in% valid.users$subject
+         , as.Date.POSIXct(appStartTime) <= '2023-05-22'
+         ) %>% 
   arrange(appStartTime)
 
+valid.users <- valid.users %>% 
+  filter(subject %in% res_all$subject)
+
 res <- res_all %>% 
-  filter(whichIsSmaller == 'Triangle (▲)',
-         as.Date.POSIXct(appStartTime) <= '2023-05-22') %>% 
+  filter(whichIsSmaller == 'Triangle (▲)'
+         , as.Date.POSIXct(appStartTime) <= '2023-05-22'
+         ) %>% 
   #Only verified results from before the start of SDSS 
@@
 
-The accuracy of graphical forms has been studied for almost a century [@vonhuhnFurtherStudiesGraphic1927;@eellsRelativeMeritsCircles1926;@croxtonGraphicComparisonsBars1932;@croxtonBarChartsCircle1927], as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions.
+The accuracy of graphical forms has been studied for almost a century [@vonhuhnFurtherStudiesGraphic1927;@eellsRelativeMeritsCircles1926;@croxtonGraphicComparisonsBars1932;@croxtonBarChartsCircle1927]; as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions.
 This is particularly true in areas like graphics which are affected by the immense technological innovation in hardware and software which has taken place since the early 1990s.
@@
 As a result, we have much more natural virtual rendering of 3D objects, and we can also print graphics in three dimensions using mass-market, relatively inexpensive 3D printers, moving artificial charts into a more natural, physical setting.
-As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].
+As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts could eventually provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].
 
@@
 A diagram of the experimental design is provided in \Cref{fig:studyDesign}.
-The study is set up as a modified randomized incomplete block design, comprised of 3 blocks (presentation mode), 7 bar ratios (5 are selected for each participant).
+The study is set up as a modified randomized incomplete block design, comprised of 3 blocks (presentation mode) and 7 bar ratios (5 are selected for each participant).
 At each combination of ratio and presentation type, the comparison type, adjacent (Type 1) or separated (Type 3), is randomly selected.
@@
 
-A total of 48 participants completed the study; demographics are shown in \Cref{fig:demographics}.
+A total of `r nrow(valid.users)` participants completed the study; demographics are shown in \Cref{fig:demographics}.
+The participants provided a total of `nrow(res_all)` responses.
 
@@
 Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in \Cref{fig:experiment3dRender}.
-It should be noted that we chose to use a slider for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries [@ruudUncertaintyCausesRounding2014a;@maineriSliderBarsMultiDevice2021].
+It should be noted that we chose to use a slider without numeric identifiers or tick marks for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries [@ruudUncertaintyCausesRounding2014a;@maineriSliderBarsMultiDevice2021].
 Another benefit of this modification is that it should reduce participants' cognitive load, allowing them to focus on the ratio between bars rather than the numerical translation process.
@@
 
-All responses ($n = `r nrow(res_all) - nrow(res)`$) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check.
-In addition, one trial was removed because the participant did not enter a valid ID for a 3D printed chart.
+All responses that incorrectly identified the smaller bar ($n = `r nrow(res_all) - nrow(res)`$) were removed from the study before analysis; this serves as a basic attention check and leaves `r nrow(res)` valid responses.
+In addition, one response was removed because the participant did not enter a valid ID for a 3D printed chart.
 
@@
 It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in @hollandsBiasProportionJudgments2000.
-Residuals from the model indicate that there is some left skew in the tails, but this is not extreme; code for this analysis can be found in the supplemental materials on github.
+Residuals from the model indicate that there is some left skew in the tails, but this is not extreme; code for this analysis can be found in the supplemental materials on GitHub.
 
@@
 
-While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only `r sum(res$byHowMuch == 50)` trials (`r sprintf("%.2f%%", mean(res$byHowMuch==50)*100)`) had slider values at exactly 0.5, and all but `r sum(res$byHowMuch == 50 & !res$ratioLabel %in% c(46.4, 56.2))` were for ratios on either side of 0.5 (.464, .562).
+While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only `r sum(res$byHowMuch == 50)` responses (`r sprintf("%.2f%%", mean(res$byHowMuch==50)*100)`) had slider values at exactly 0.5, and all but `r sum(res$byHowMuch == 50 & !res$ratioLabel %in% c(46.4, 56.2))` were for ratios on either side of 0.5 (.464, .562).
 This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5.
@@
 
-The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger.
+The first potential explanation is that this study is underpowered - the effect size is small, and our `r nrow(valid.users)` participants were insufficient; the original study included 51 participants, which is slightly larger.
 However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @clevelandGraphical1984 do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.

### 2024-03-21 jds.tex
Updated paper based on reviewer feedback

@@
 
-The accuracy of graphical forms has been studied for almost a century \citep{vonhuhnFurtherStudiesGraphic1927, eellsRelativeMeritsCircles1926, croxtonGraphicComparisonsBars1932, croxtonBarChartsCircle1927}, as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions.
+The accuracy of graphical forms has been studied for almost a century \citep{vonhuhnFurtherStudiesGraphic1927, eellsRelativeMeritsCircles1926, croxtonGraphicComparisonsBars1932, croxtonBarChartsCircle1927}; as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions.
 This is particularly true in areas like graphics which are affected by the immense technological innovation in hardware and software which has taken place since the early 1990s.
@@
 As a result, we have much more natural virtual rendering of 3D objects, and we can also print graphics in three dimensions using mass-market, relatively inexpensive 3D printers, moving artificial charts into a more natural, physical setting.
-As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts provide the opportunity to make data graphics accessible to those with limited or absent vision \citep{fleetHowTactileGraphics2023}.
+As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts could eventually provide the opportunity to make data graphics accessible to those with limited or absent vision \citep{fleetHowTactileGraphics2023}.
 
@@
 
-A total of 48 participants completed the study; demographics are shown in \Cref{fig:demographics}.
+A total of 38 participants completed the study; demographics are shown in \Cref{fig:demographics}.
+The participants provided a total of \texttt{nrow(res\_all)} responses.
 
@@
 Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in \Cref{fig:experiment3dRender}.
-It should be noted that we chose to use a slider for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries \citep{ruudUncertaintyCausesRounding2014a, maineriSliderBarsMultiDevice2021}.
+It should be noted that we chose to use a slider without numeric identifiers or tick marks for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries \citep{ruudUncertaintyCausesRounding2014a, maineriSliderBarsMultiDevice2021}.
 Another benefit of this modification is that it should reduce participants' cognitive load, allowing them to focus on the ratio between bars rather than the numerical translation process.
@@
 
-All responses (\(n = 33\)) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check.
-In addition, one trial was removed because the participant did not enter a valid ID for a 3D printed chart.
+All responses that incorrectly identified the smaller bar (\(n = 18\)) were removed from the study before analysis; this serves as a basic attention check and leaves 477 valid responses.
+In addition, one response was removed because the participant did not enter a valid ID for a 3D printed chart.
 
@@
 \addlinespace
-s(participant) & 27.0449 & 34.0000 & 4.7199 & 0.0000\\
+s(participant) & 27.0449 & 34.0000 & 4.7276 & 0.0000\\
 \bottomrule
@@
 It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in \citet{hollandsBiasProportionJudgments2000}.
-Residuals from the model indicate that there is some left skew in the tails, but this is not extreme; code for this analysis can be found in the supplemental materials on github.
+Residuals from the model indicate that there is some left skew in the tails, but this is not extreme; code for this analysis can be found in the supplemental materials on GitHub.
 
@@
 
-While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only 42 trials (8.81\%) had slider values at exactly 0.5, and all but 6 were for ratios on either side of 0.5 (.464, .562).
+While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only 42 responses (8.81\%) had slider values at exactly 0.5, and all but 6 were for ratios on either side of 0.5 (.464, .562).
 This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5.
@@
 
-The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger.
+The first potential explanation is that this study is underpowered - the effect size is small, and our 38 participants were insufficient; the original study included 51 participants, which is slightly larger.
 However, we should note that the analysis methods used in \citet{heerCrowdsourcingGraphicalPerception2010b} and \citet{clevelandGraphical1984} do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.

### 2024-03-24 jds.rmd
Material updates for revision 2

@@
 This replicates the spirit of the recruitment method without the implicit assumptions that graduate students and professors are male, heterosexual, and have unemployed spouses.
-
-A total of `r nrow(valid.users)` participants completed the study; demographics are shown in \Cref{fig:demographics}.
-The participants provided a total of `nrow(res_all)` responses.
+A total of `r length(unique(res_all$subject))` participants completed our study; demographics are shown in \Cref{fig:demographics}.
 
@@
 
+```{r, include = F}
+res_sub_count <- res_all %>% 
+  group_by(subject) %>% 
+  count()
+```
+
+The participants provided a total of `r nrow(res_all)` responses.
+While each participant was provided with 15 trials, `r sum(res_sub_count$n < 15)` participants did not complete the entire experiment.
 All responses that incorrectly identified the smaller bar ($n = `r nrow(res_all) - nrow(res)`$) were removed from the study before analysis; this serves as a basic attention check and leaves `r nrow(res)` valid responses.
@@
 ```{r midmeans-log-errors, fig.cap='Midmeans and observed values of log absolute errors for the true ratio of bars.
-Summary lines are computed from raw data using a loess smooth.', message=FALSE, warning=FALSE, out.width='\\textwidth'}
+Summary lines are computed from raw data using a loess smooth.', message=FALSE, warning=FALSE, out.width='\\textwidth', fig.height=3}
 plot.types = c('2D', 'Rendered 3D', '3D Printed')
@@
   theme_bw() + ylab("Prediction") + xlab("True Proportion Difference") + 
-  xlim(c(0, 1))
+  xlim(c(0, 1)) + 
+  theme(legend.position = 'bottom')
 

### 2024-04-16 jds.rmd
Final revisions

@@
 This replicates the spirit of the recruitment method without the implicit assumptions that graduate students and professors are male, heterosexual, and have unemployed spouses.
-A total of `r length(unique(res_all$subject))` participants completed our study; demographics are shown in \Cref{fig:demographics}.
+A total of `r nrow(valid.users)` participants completed our study; demographics are shown in \Cref{fig:demographics}.
 
@@
 Then, participants were shown a "practice" page which allowed them to experiment with the data collection interface and practice estimating the ratio between the bars, as shown in \Cref{fig:practice}.
-This practice interface did not provide any feedback as to participant correctness, but was merely intended to familiarize the participants with the questions which would be asked as well as the process of estimating the ratio between the two bars.
+This practice interface did not provide any numeric feedback as to participant correctness, but was merely intended to familiarize the participants with the questions which would be asked as well as the process of estimating the ratio between the two bars.
 
@@
 res_sub_count <- res_all %>% 
-  group_by(subject) %>% 
+  group_by(nickname, participantUnique, appStartTime, subject) %>% 
   count()
@@
 The participants provided a total of `r nrow(res_all)` responses.
-While each participant was provided with 15 trials, `r sum(res_sub_count$n < 15)` participants did not complete the entire experiment.
+While each participant was provided with 15 trials, `r sum(res_sub_count$n < 15)` participants partially completed the experiment; one participant did not advance beyond the demographic page. 
 All responses that incorrectly identified the smaller bar ($n = `r nrow(res_all) - nrow(res)`$) were removed from the study before analysis; this serves as a basic attention check and leaves `r nrow(res)` valid responses.

### 2024-04-16 jds.tex
Final revisions

@@
 This replicates the spirit of the recruitment method without the implicit assumptions that graduate students and professors are male, heterosexual, and have unemployed spouses.
-
-A total of 38 participants completed the study; demographics are shown in \Cref{fig:demographics}.
-The participants provided a total of \texttt{nrow(res\_all)} responses.
+A total of 38 participants completed our study; demographics are shown in \Cref{fig:demographics}.
 
@@
 Then, participants were shown a ``practice'' page which allowed them to experiment with the data collection interface and practice estimating the ratio between the bars, as shown in \Cref{fig:practice}.
-This practice interface did not provide any feedback as to participant correctness, but was merely intended to familiarize the participants with the questions which would be asked as well as the process of estimating the ratio between the two bars.
+This practice interface did not provide any numeric feedback as to participant correctness, but was merely intended to familiarize the participants with the questions which would be asked as well as the process of estimating the ratio between the two bars.
 
@@
 
+The participants provided a total of 495 responses.
+While each participant was provided with 15 trials, 6 participants partially completed the experiment; one participant did not advance beyond the demographic page.
 All responses that incorrectly identified the smaller bar (\(n = 18\)) were removed from the study before analysis; this serves as a basic attention check and leaves 477 valid responses.
